---
title: "Take-home_Ex02"
author: "Nguyen Kim Hau"
date: "May 26, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 1 Load Dataset

```{r}
pacman::p_load(jsonlite, tidyverse, plotly,
               tidygraph, ggraph,
               DT, gt, skimr,
               visNetwork,lubridate,
               crosstalk)
```

```{r}
#| eval: true
#| code-fold: false
#| 
if (!file.exists("data/mc1_data.rds")) {
  mc1_data <- fromJSON("data/mc1_data.json")
  write_rds(mc1_data, "data/mc1_data.rds")
}

mc1_data  <- read_rds("data/mc1_data.rds")
```

```{r}
glimpse(mc1_data)
```

```{r}
names(mc1_data$rounds)
```

```{r}
glimpse(mc1_data$rounds$environment_context)
```

```{r}
mc1_data$rounds$communications[[1]] %>% 
  head(2)
```

```{r}
mc1_data$rounds$communications[[1]] %>%
  head(2) %>%
  unnest(internal_state)
```

Unnest

```{r}
comms_df <- mc1_data$rounds %>%
  select(hour, communications) %>%
  unnest(communications) %>%
  unnest(internal_state)
glimpse(comms_df)

```

```{r}
# Environment context data  
env_df <- mc1_data$rounds %>%
  select(hour, environment_context) %>%
  unnest(environment_context)
glimpse(env_df)
```

Check each hour have how many critical_deadlines

```{r}
env_df %>%
  mutate(n = map_int(critical_deadlines, length)) %>%
  select(hour, n)
```

```{r}
# Participants data
participants_df <- mc1_data$rounds %>%
  select(hour, participants) %>%
  unnest(participants)
glimpse(participants_df)
```

## **2. Question 1: What were the key events leading to the inappropriate release?**

### 2.1 Network graph shows who communicated with whom

Step 1: Check recipients column first

```{r}
# See what recipients looks like
comms_df %>%
  select(agent_id, recipients, channel) %>%
  head(10)

```

```{r}
# Create role-to-agent mapping
role_map <- c(
  "ALL"            = "ALL",
  "intern"         = "intern_agent",
  "judge"          = "judge_eval_agent",
  "legal"          = "legal_agent",
  "platform_trust" = "quality_agent",
  "pr"             = "pr_agent",
  "pr_intern"      = "pr_intern_agent",
  "social_manager" = "social_manager_agent"
)
role_map
```

Step 2: Build edges and nodes

```{r}
# Build edges (sender → recipient)
# Fix edges_df
edges_df <- comms_df %>%
  select(agent_id, recipients, channel, hour) %>%
  unnest(recipients) %>%
  filter(!is.na(recipients)) %>%
  mutate(to = role_map[recipients]) %>%
  filter(to != "ALL") %>%  # remove broadcast messages
  rename(from = agent_id) %>%
  select(from, to, channel, hour)

# Check result
edges_df %>% count(to)

# Aggregate edges: count messages between each pair
edges_agg <- edges_df %>%
  filter(from != to) %>%  # remove self-loops
  count(from, to, name = "weight") %>%
  mutate(width = weight / max(weight) * 10)

# Build nodes from participants_df
nodes_df <- participants_df %>%
  distinct(agent_id, agent_role, agent_label) %>%
  rename(id = agent_id, label = agent_label)
```

Step 3: Draw with visNetwork

```{r}
# Rebuild visNetwork with aggregated edges
visNetwork(nodes_df, edges_agg) %>%
  visEdges(arrows = "to", 
           width = "weight",  # thicker = more messages
           smooth = FALSE) %>%  # straight lines, less chaos
  visNodes(size = 20) %>%
  visOptions(highlightNearest = TRUE) %>%
  visLayout(randomSeed = 42)
```

### 2.2 Timeline visualization of events by hour

```{r}
#| code-fold: true
#| code-summary: "Show the code"
comms_df %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  count(date, agent_id) %>%
  plot_ly(x = ~date, y = ~n, color = ~agent_id,
          type = "scatter", mode = "lines+markers",
          hoverinfo = "text",
          text = ~paste("Agent:", agent_id,
                        "<br>Date:", date,
                        "<br>Messages:", n)) %>%
  layout(title = "Message Frequency by Agent Over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Number of Messages"),
         legend = list(title = list(text = "Agent")))
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
participants_df %>%
  select(hour, agent_id, agent_label, declared_action) %>%
  filter(!is.na(declared_action)) %>%
  mutate(hour = ymd_hms(hour)) %>%
  arrange(hour, agent_id) %>%
  DT::datatable(
    filter = "top",
    options = list(pageLength = 15)
  )
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Prepare data
timeline_data <- comms_df %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  filter(timestamp < ymd_hms("2046-06-05 17:00:00")) %>%
  mutate(content_short = str_trunc(content, 150))

# Create shared data for crosstalk
shared_data <- SharedData$new(timeline_data)

# Filter widget + plot
bscols(
  widths = c(3, 9),
  list(
    filter_select("date", "Select Date", shared_data, ~date),
    filter_select("agent", "Select Agent", shared_data, ~agent_id),
    filter_select("channel", "Select Channel", shared_data, ~channel)
  ),
  plot_ly(shared_data, 
          x = ~timestamp, 
          y = ~agent_id,
          color = ~channel,
          symbol = ~message_type,
          text = ~paste("<b>Agent:</b>", agent_label,
                        "<br><b>Time:</b>", timestamp,
                        "<br><b>Channel:</b>", channel,
                        "<br><b>Type:</b>", message_type,
                        "<br><b>Content:</b>", content_short),
          hoverinfo = "text",
          type = "scatter",
          mode = "markers",
          marker = list(size = 10, opacity = 0.7)) %>%
    layout(title = "Agent Communications Timeline (Before 5PM Breach)",
           xaxis = list(title = "Time"),
           yaxis = list(title = "Agent"))
)
```

### 2.3 Identify the causal chain of decisions that led to the embargo breach

WHAT happened (the timeline of key decisions)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Find all messages mentioning key terms

critical_events <- comms_df %>%
  mutate(timestamp = ymd_hms(timestamp)) %>%
  filter(str_detect(content, regex(
    "embargo|HarborCrest|CivicLoom|merger|post|publish|anonymous|Flex",
    ignore_case = TRUE
  ))) %>%
  select(timestamp, agent_id, agent_label, channel, 
         content, deliberating, message_id, responding_to)
  
critical_events %>%
  select(timestamp, agent_id, channel, content, deliberating) %>%
  arrange(timestamp)
```

The `deliberating` column reveals what agents were *actually thinking*

```{r}
# Track legal_agent's internal reasoning over time
comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  select(timestamp, deliberating) %>%
  arrange(timestamp)
```

```{r}
comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  select(deliberating)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Build response chains
comms_df %>%
  filter(!is.na(responding_to)) %>%
  select(message_id, agent_id, responding_to, timestamp, content) %>%
  arrange(timestamp)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
pacman::p_load(networkD3)

# Step 1: Identify the critical decision points
critical_events  %>%
  select(timestamp, agent_id, agent_label, channel, 
         content, deliberating, message_id, responding_to) %>%
  arrange(timestamp)

# Simpler: agent → channel → action flow
simple_flow <- critical_events %>%
  mutate(action_type = case_when(
    str_detect(content, "POSTED_ANONYMOUS") ~ "Anonymous Post",
    str_detect(content, "POSTED_PERSONAL") ~ "Personal Post",
    str_detect(channel, "official") ~ "Official Post",
    TRUE ~ "Internal Discussion"
  )) %>%
  count(agent_id, channel, action_type, name = "value")

# Build nodes
all_names <- unique(c(simple_flow$agent_id, simple_flow$channel, simple_flow$action_type))
nodes <- data.frame(name = all_names)

simple_flow$source <- match(simple_flow$agent_id, nodes$name) - 1
simple_flow$target_ch <- match(simple_flow$channel, nodes$name) - 1

# Two-level Sankey: Agent → Channel
links_1 <- simple_flow %>%
  group_by(agent_id, channel) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  mutate(source = match(agent_id, nodes$name) - 1,
         target = match(channel, nodes$name) - 1) %>%
  select(source, target, value)

# Channel → Action Type
links_2 <- simple_flow %>%
  group_by(channel, action_type) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  mutate(source = match(channel, nodes$name) - 1,
         target = match(action_type, nodes$name) - 1) %>%
  select(source, target, value)

links_all <- bind_rows(links_1, links_2)

sankeyNetwork(
  Links = as.data.frame(links_all),
  Nodes = nodes,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  fontSize = 14,
  nodeWidth = 30
)
```

::: callout-note
Insights:

**1. Legal-Agent is the most active — and the most suspicious**

- Legal-Agent has the **largest flow** of embargo-related messages
- Legal-Agent is the **only agent connected to `anonymous_post`** — this is the breach channel
- A senior legal counsel posting anonymously about the merger they're supposed to protect is a major red flag

**2. The anonymous post channel is isolated**

- Only `legal_agent` flows to `anonymous_post` — no other agent used this channel
- This suggests the anonymous posting was a **solo decision**, not a team-coordinated leak

**3. Most embargo discussion stayed internal**

- The overwhelming majority of messages went to `Internal Discussion` via `comms_huddle` and `one_on_one_chat`
- This means agents *were* discussing the embargo internally — but the controls failed to prevent the breach

**4. Social-Media-Agent also posted externally**

- `social_media_agent` connects to `personal_post` — they posted sentiment updates publicly
- This is a second potential breach, though less severe than the anonymous post

**5. PR-Intern-Agent had access to channels**

- Despite being junior, `pr_intern_agent` has flow to multiple channels including `official_post`
- Junior staff with posting access is a system design risk
:::

## 3. Question 2: Was the evasion of the embargo a new behavior?

### 3.1 Compare agent behavior **before vs during** the crisis

**Sentiment analysis** of the deliberating text

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(tidytext)

# NRC emotion analysis on legal_agent's deliberating
legal_emotions <- comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  select(date, deliberating) %>%
  unnest_tokens(word, deliberating) %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  count(date, sentiment) %>%
  filter(!sentiment %in% c("positive", "negative"))  # keep only emotions

# Plot emotion trajectory
plot_ly(legal_emotions, 
        x = ~date, y = ~n, color = ~sentiment,
        type = "scatter", mode = "lines+markers") %>%
  layout(title = "Legal-Agent Emotional Trajectory",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Word Count"))
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
legal_emotions <- comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  select(date, deliberating) %>%
  unnest_tokens(word, deliberating) %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  count(date, sentiment) %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  group_by(date) %>%
  mutate(share = n / sum(n)) %>%
  ungroup()

# Plot as share
plot_ly(legal_emotions, 
        x = ~date, y = ~share, color = ~sentiment,
        type = "scatter", mode = "none",
        stackgroup = "one", 
        groupnorm = "percent") %>%
  layout(title = "Legal-Agent Emotion Share Over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Share (%)", ticksuffix = "%"))
```

::: callout-not
**1. Trust drops dramatically on crisis day (Jun 5)**

- Trust (grey) was dominant at \~30-40% throughout, but drops to \~15% on Jun 5

- This means Legal-Agent lost confidence in the system/process right before the breach

**2. Anticipation surges on Jun 5**

- Anticipation (orange) grows from \~15% to \~30% on Jun 5

- Legal-Agent was expecting something — they knew what was coming because they planned it

**3. Disgust appears and grows toward crisis**

- Disgust (green) barely exists early on, but grows from May 28 onward

- This suggests growing frustration with the situation or the embargo restrictions

**4. Fear spikes around May 27-28 then stabilizes**

- Fear (pink) peaks mid-timeline when pressure was building

- But fear *decreases* toward Jun 5 — again suggesting the decision was already made, not a panic reaction

**5. Anger appears on Jun 5**

- Anger (blue) was near zero most of the time, but appears on the crisis day

- Combined with dropping trust and rising anticipation, this suggests frustration-driven action
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#  sentence-level sentiment
pacman::p_load(sentimentr)

legal_deliberating <- comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  select(date, deliberating)

# Sentence-level sentiment
sent_result <- legal_deliberating %>%
  mutate(id = row_number()) %>%
  unnest_tokens(sentence, deliberating, token = "sentences") %>%
  mutate(sentiment = sentiment_by(sentence)$ave_sentiment)

# Aggregate by date
sent_by_date <- sent_result %>%
  group_by(date) %>%
  summarise(avg_sentiment = mean(sentiment),
            min_sentiment = min(sentiment),
            max_sentiment = max(sentiment))

plot_ly(sent_by_date, x = ~date) %>%
  add_trace(y = ~avg_sentiment, type = "scatter", mode = "lines+markers",
            name = "Average") %>%
  add_ribbons(ymin = ~min_sentiment, ymax = ~max_sentiment,
              name = "Range", opacity = 0.2) %>%
  layout(title = "Legal-Agent Sentiment (sentimentr)",
         yaxis = list(title = "Sentiment Score"),
         shapes = list(
           list(type = "line", x0 = "2046-06-05", x1 = "2046-06-05",
                y0 = -1, y1 = 1,
                line = list(color = "red", dash = "dash"))))
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Keyword-based thematic analysis
legal_themes <- comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  mutate(
    mentions_embargo = str_detect(deliberating, regex("embargo|compliance|restrict", ignore_case = TRUE)),
    mentions_risk    = str_detect(deliberating, regex("risk|danger|threat|worry|concern", ignore_case = TRUE)),
    mentions_strategy = str_detect(deliberating, regex("strategy|plan|position|frame|narrative", ignore_case = TRUE)),
    mentions_post    = str_detect(deliberating, regex("post|publish|anonymous|Flex|public", ignore_case = TRUE)),
    mentions_pressure = str_detect(deliberating, regex("pressure|urgent|deadline|time|fast|quick", ignore_case = TRUE))
  ) %>%
  select(date, starts_with("mentions_")) %>%
  pivot_longer(cols = starts_with("mentions_"),
               names_to = "theme",
               values_to = "present") %>%
  mutate(theme = str_remove(theme, "mentions_"))

# Heatmap of themes over time
ggplotly(
  legal_themes %>%
    filter(present == TRUE) %>%
    ggplot(aes(x = date, y = theme, fill = theme)) +
    geom_tile(color = "white", height = 0.8) +
    scale_fill_manual(values = c(
      "embargo"  = "#e74c3c",   # red — compliance
      "risk"     = "#f39c12",   # orange — warning
      "pressure" = "#9b59b6",   # purple — stress
      "post"     = "#2ecc71",   # green — action
      "strategy" = "#3498db"    # blue — planning
    )) +
    labs(title = "Legal-Agent: Themes in Deliberation Over Time",
         x = "Date", y = "Theme") +
    theme_minimal() 
)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Channel usage by agent over time
channel_usage <- comms_df %>%
  mutate(timestamp = ymd_hms(timestamp),
         date = as.Date(timestamp)) %>%
  count(date, agent_id, channel)

# Heatmap view
ggplotly(
  channel_usage %>%
    ggplot(aes(x = date, y = channel, fill = n)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "lightyellow", high = "red") +
    facet_wrap(~agent_id, ncol = 2) +
    labs(title = "Heatmap of Channel Usage by Agent Over Time",
         x = "Date", y = "Channel", fill = "Messages") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                     size = 6))
)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(crosstalk, wordcloud2, htmlwidgets)
# Prepare word frequencies by agent and date
word_freq <- comms_df %>%
  mutate(date = as.Date(ymd_hms(timestamp))) %>%
  select(date, agent_id, content) %>%
  unnest_tokens(word, content) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_length(word) > 2) %>%
  count(agent_id, word, sort = TRUE)

# Word cloud per agent using wordcloud2
# Pick one agent at a time
agent_wc <- function(agent_name) {
  wc_data <- word_freq %>%
    filter(agent_id == agent_name) %>%
    select(word, n) %>%
    head(100)
  
  wordcloud2(wc_data, size = 0.5,
             color = "random-light",
             backgroundColor = "white")
}

# Example: Legal Agent word cloud
agent_wc("legal_agent")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
word_freq_date <- comms_df %>%
  mutate(date = as.Date(ymd_hms(timestamp))) %>%
  select(date, agent_id, content) %>%
  unnest_tokens(word, content) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_length(word) > 2) %>%
  count(date, agent_id, word, sort = TRUE)

shared_words <- SharedData$new(
  word_freq_date %>%
    group_by(agent_id, word) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    filter(n > 0) %>%  # remove 0 count
    group_by(agent_id) %>%
    slice_max(n, n = 15) %>%
    ungroup()
)

bscols(
  widths = c(3, 9),
  list(
    filter_select("agent", "Select Agent", shared_words, ~agent_id)
  ),
  plot_ly(shared_words,
          y = ~reorder_within(word, n, agent_id), 
          x = ~n, color = ~agent_id,
          type = "bar", orientation = "h",
          text = ~paste(word, ":", n),
          hoverinfo = "text") %>%
    layout(title = "Top Words by Agent",
           yaxis = list(title = "", 
                        categoryorder = "total ascending"),
           xaxis = list(title = "Frequency"),
           showlegend = FALSE)
)
```

```{r}
# Check what's in market_snapshot
glimpse(env_df$market_snapshot)
```

```{r}
# Extract and clean market data
market_df <- env_df %>%
  select(hour, market_snapshot) %>%
  unnest(market_snapshot) %>%
  mutate(
    date = as.Date(ymd_hms(hour)),
    stock_price = as.numeric(str_remove(stock_price, "\\$")),
    percent_change = as.numeric(str_remove(percent_change, "%"))
  )

# Get sentiment scores
legal_sentiment <- comms_df %>%
  filter(agent_id == "legal_agent",
         !is.na(deliberating)) %>%
  mutate(date = as.Date(ymd_hms(timestamp))) %>%
  select(date, deliberating) %>%
  unnest_tokens(word, deliberating) %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  group_by(date) %>%
  summarise(sentiment_score = sum(value))

# Join market + sentiment
dual_data <- market_df %>%
  select(date, stock_price, percent_change, sentiment) %>%
  left_join(legal_sentiment, by = "date")

# Dual-axis plot
plot_ly(dual_data) %>%
  add_trace(x = ~date, y = ~stock_price, 
            type = "scatter", mode = "lines+markers",
            name = "Stock Price ($)",
            line = list(color = "#3498db", width = 2),
            text = ~paste("Date:", date,
                          "<br>Stock:", stock_price,
                          "<br>Change:", percent_change, "%",
                          "<br>Market Sentiment:", sentiment),
            hoverinfo = "text") %>%
  add_trace(x = ~date, y = ~sentiment_score,
            type = "scatter", mode = "lines+markers",
            name = "Legal-Agent Sentiment",
            yaxis = "y2",
            line = list(color = "#e74c3c", width = 2),
            text = ~paste("Date:", date,
                          "<br>Sentiment Score:", sentiment_score),
            hoverinfo = "text") %>%
  layout(
    title = "Stock Price vs Legal-Agent Internal Sentiment",
    xaxis = list(title = "Date"),
    yaxis = list(title = "Stock Price ($)", side = "left",
                 titlefont = list(color = "#3498db")),
    yaxis2 = list(title = "Sentiment Score (AFINN)", side = "right",
                  overlaying = "y",
                  titlefont = list(color = "#e74c3c")),
    shapes = list(
      list(type = "line",
           x0 = "2046-06-05", x1 = "2046-06-05",
           y0 = 0, y1 = 1, yref = "paper",
           line = list(color = "red", dash = "dash", width = 2))),
    annotations = list(
      list(x = "2046-06-05", y = 1.05, yref = "paper",
           text = "BREACH DAY", showarrow = FALSE,
           font = list(color = "red", size = 12, bold = TRUE)))
  )
```

```{r}
plot_ly(dual_data) %>%
  add_trace(x = ~date, y = ~percent_change,
            type = "scatter", mode = "lines+markers",
            name = "Stock % Change",
            line = list(color = "#e74c3c", width = 2),
            marker = list(size = 8),
            text = ~paste("Date:", date,
                          "<br>Change:", percent_change, "%",
                          "<br>Market:", sentiment),
            hoverinfo = "text") %>%
  add_trace(x = ~date, y = ~sentiment_score,
            type = "scatter", mode = "lines+markers",
            name = "Legal-Agent Sentiment",
            yaxis = "y2",
            line = list(color = "#3498db", width = 2),
            marker = list(size = 8),
            text = ~paste("Date:", date,
                          "<br>Sentiment Score:", sentiment_score),
            hoverinfo = "text") %>%
  layout(
    title = "Market Pressure vs Legal-Agent Internal Sentiment",
    xaxis = list(title = "Date"),
    yaxis = list(title = "Stock % Change",
                 ticksuffix = "%",
                 titlefont = list(color = "#e74c3c"),
                 zeroline = TRUE,
                 zerolinecolor = "grey"),
    yaxis2 = list(title = "Sentiment Score (AFINN)",
                  side = "right",
                  overlaying = "y",
                  titlefont = list(color = "#3498db")),
    shapes = list(
      list(type = "line",
           x0 = "2046-06-05", x1 = "2046-06-05",
           y0 = 0, y1 = 1, yref = "paper",
           line = list(color = "red", dash = "dash", width = 2))),
    annotations = list(
      list(x = "2046-06-05", y = 1.05, yref = "paper",
           text = "BREACH DAY", showarrow = FALSE,
           font = list(color = "red", size = 12)))
  )
```
