---
title: "Take-home Ex 2: VAST Challenge 2026"
author: "Nguyen Kim Hau"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: true
bibliography: references.bib
---

# Overview

This report focuses on [Mini-Challenge (MC) 2](https://vast-challenge.github.io/2026/MC2.html) of [VAST Challenge 2026](https://vast-challenge.github.io/2026/index.html).

**Context**: You work in the AI systems administration department of Company A. A post was recently made on SaidIT looks like a malfunction of an agent: It is in gibberish and in a forum that appears random. Your job is to track down why this post occurred and suggest a remedy to the situation by building a visual analytics system that integrates visuals with algorithms to enable interactive exploration and documentation of the multi-agent system!

The VAST Challenge lives up to its name. The dataset is massive, and the challenge is real. A turning point came when I read *Analysis Without Data: Teaching Students to Tackle the VAST Challenge* (@he), one of good resources that Prof. Kam shared. The paper advocates generating questions and formalising hypotheses before touching the data, as a safeguard against **HARKing** (Hypothesizing After Results are Known) and false discoveries.

It felt like an "Aha" moment to me. These are steps we often do intuitively and haphazardly, just without thinking of them as a systematic workflow. Therefore, by any chance you read my assignment, I hope you can find this reference useful.

I focus to answer the Question 1 from MC2:

> 1.  How was the anomalous SaidIT posts made? Build a visual tool that describes the sequence of events that led up to the post. Include people and system interactions that were important to the post being made.
>
>     a\. Provide a detailed view that focuses on the exact chain of events for this message of interest.
>
>     b\. Provide a system overview that contextualizes this message chain in that overall system.

# Setup

## Loading R packages

The packages below cover the four jobs we'll do in this notebook:

```{r}
#| echo: false
#| code-fold: false

knitr::kable(
  tibble::tribble(
    ~Package,      ~Purpose,
    "jsonlite",    "Read and parse JSON files",
    "tidyverse",   "Data wrangling and visualization",
    "plotly",      "Interactive charts with tooltips and zoom",
    "tidygraph",   "Create graph objects from nodes and edges",
    "ggraph",      "Static network visualization",
    "DT",          "Interactive HTML tables",
    "gt",          "Formatted static tables",
    "skimr",       "Quick data summaries",
    "visNetwork",  "Interactive network graphs",
    "lubridate",   "Date/time parsing and manipulation"
  )
)
```

```{r}
#| code-fold: false
pacman::p_load(jsonlite, tidyverse, plotly,
               tidygraph, ggraph,
               DT, gt, skimr,
               visNetwork,lubridate)
```

## Reading the raw data

The JSON files are large and slow to re-parse on every render. The convention is to parse JSON **once** with `fromJSON()`, then cache as `.rds` (R's native binary format: smaller, faster, and it preserves column types).

The `if (!file.exists(...))` guard means the conversion runs only the first time.

```{r}
#| eval: false
#| code-fold: false
#| 
if (!file.exists("data/mc2_data.rds")) {
  mc2_data <- fromJSON("data/mc2_data.json")
  write_rds(mc2_data, "data/mc2_data.rds")
}
if (!file.exists("data/org_chart.rds")) {
  org_chart <- fromJSON("data/org_chart.json")
  write_rds(org_chart, "data/org_chart.rds")
}
```

```{r}
#| code-fold: false
mc2_data  <- read_rds("data/mc2_data.rds")
org_chart <- read_rds("data/org_chart.rds")

```

# Data Wrangling

## Organisation data

### Organisation chart

`org_chart` is in NetworkX node-link format: a `nodes` list and an `edges` list.

Each person has an `id`, `label`, `type` (`person` / `team` / `department` / `company`), and sometimes a `title`.

Edges carry a `relation` (`contains` or `led_by`) describing how parts of the org nest.

```{r}
#| code-fold: false
glimpse(org_chart)
```

#### Extracting the nodes and edges (links) data from org_chart

```{r}
#| code-fold: false
org_nodes <- as_tibble(org_chart$nodes)
org_edges <- as_tibble(org_chart$edges)

org_nodes %>% count(type)
```

```{r}
#| code-fold: false
org_edges %>% count(relation)
```

#### Draw organization tree

Here is an interactive chart created by visNetwork, showing hierachy levels (company, department and team). We can zoom/search/drag and hover by name of each employee.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

vis_nodes <- org_nodes %>%
  mutate(
    id = id,
    label = label,
    group = type,
    title = paste0("<b>", label, "</b><br>Type: ", type,
                   if_else(!is.na(title), paste0("<br>Title: ", title), ""))
  )

vis_edges <- org_edges %>%
  rename(from = source, to = target) %>%
  mutate(
    arrows = "to",
    dashes = (relation == "led_by"),
    color = if_else(relation == "led_by", "red", "grey")
  )

visNetwork(vis_nodes, vis_edges,
           main = "Tenant Thread Organization Chart") %>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 120,
    nodeSpacing = 150
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.15) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 2, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, useLabels = TRUE, main = "Search person"),
    selectedBy = list(variable = "group", multiple = TRUE)
  ) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

```

```{r}
# Personnel count by department
team_dept <- org_edges %>%
  filter(str_starts(source, "department:"), str_starts(target, "team:")) %>%
  select(dept = source, team = target)

org_edges %>%
  filter(str_starts(target, "person:")) %>%
  left_join(team_dept, by = c("source" = "team")) %>%
  mutate(dept = coalesce(dept, source)) %>%
  left_join(org_nodes %>% select(id, label), by = c("dept" = "id")) %>%
  count(label) %>%
  ggplot(aes(x = reorder(label, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  labs(title = "Personnel Count by Department", x = NULL, y = "Personnel") +
  theme_minimal()
```

::: callout-note
There are some information from these charts and tables above:

- TenantThread Company has 6 departments, 19 teams and 49 personnel.
- There are five departments (Products, HR, Legal, IT and CS), and special department which is Executive Suite.
- Each department has 6-13 people, including department lead and normal employees.
:::

### Person lookup with department & team

The org chart is a tree: `company → department → team → person`, except for department leads (like John Windward) who hang directly under a department rather than under a team. We walk down from each department to build a flat lookup that, for every person, gives us their team and department.

```{r}
# Build a directed graph so we can traverse it cleanly
org_graph <- tbl_graph(nodes = org_nodes, edges = org_edges, directed = TRUE)

# Recursive descent: for each department, find every person reachable below it
descendants_of <- function(start_id, edges_df) {
  out <- character(0)
  frontier <- start_id
  while (length(frontier) > 0) {
    children <- edges_df %>% 
      filter(source %in% frontier) %>% 
      pull(target)
    out <- c(out, children)
    frontier <- children
  }
  unique(out)
}

dept_ids <- org_nodes %>% filter(type == "department") %>% pull(id)

person_lookup <- map_dfr(dept_ids, function(d) {
  desc <- descendants_of(d, org_edges)
  tibble(department_id = d, descendant_id = desc)
}) %>% 
  # keep only persons among the descendants
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id"))

# Also find the team (if any) for each person
team_ids <- org_nodes %>% filter(type == "team") %>% pull(id)

person_team <- map_dfr(team_ids, function(t) {
  desc <- descendants_of(t, org_edges)
  tibble(team_id = t, descendant_id = desc)
}) %>% 
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id")) %>% 
  select(person_id = descendant_id, team_id)

person_lookup <- person_lookup %>% 
  transmute(
    person_id     = descendant_id,
    person_name   = label,
    title         = title,
    department_id = department_id
  ) %>% 
  left_join(person_team, by = "person_id") %>% 
  left_join(org_nodes %>% filter(type == "department") %>% 
              select(department_id = id, department_name = label),
            by = "department_id") %>% 
  left_join(org_nodes %>% filter(type == "team") %>% 
              select(team_id = id, team_name = label),
            by = "team_id")

person_lookup %>%
  select(person_name, title, department_name, team_name) %>%
  arrange(department_name, team_name, person_name) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    colnames = c("Name", "Title", "Department", "Team"),
    caption = "Tenant Thread Personnel Directory",
    options = list(pageLength = 15, scrollX = TRUE)
  )
```

## Inspecting raw mc2_data

`mc2_data` is a list with a `description` field and an `events` data frame. The events table is where everything happens, each row is one action taken by a person, an agent, or a system.

```{r}
#| code-fold: false
glimpse(mc2_data)
```

```{r}
#| code-fold: false
mc2_events <- mc2_data$events
glimpse(mc2_events)
```

::: callout-note
A few things to notice from the glimpse above:

- `parties` is a **list-column**, each row is a character vector of one or more actor IDs (people like `person:john_windward`, agents like `agent:xxx`, or systems).
- `details` is a **nested data frame**, its columns differ by event type, so most cells are `NA`. This is normal for an event-log schema.
- `when` is a Unix timestamp (numeric), we'll convert it to a proper datetime.
- `short_name` tells us the *kind* of action, this is the field we'll group on.
:::

## Select nescessary columns in mc2_events

We create the table `events_clean`: one row per event, with a proper `datetime` and a stable `row` index.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
# Step 1: Flatten events
events_clean <- mc2_data$events %>%
  mutate(
    event_id       = id,
    datetime       = as_datetime(when, tz = "UTC"),
    date           = as_date(datetime),
    hour           = hour(datetime),
    parties_str    = sapply(parties, paste, collapse = " | "),
    target         = details$target,
    file           = details$file,
    subject        = details$subject,
    content        = details$content,
    forum          = details$forum,
    topic          = details$topic,
    advice         = details$advice,
    task           = details$task,
    action         = details$action,
    status         = details$status,
    person         = details$person,
    target_agent   = details$target_agent
  ) %>%
  select(-id, -when, -details)
glimpse (events_clean)
```

::: callout-note
What we have done:

- **Convert timestamps**: Working with Unix timestamps is painful for filtering by date. We convert `when` into a `POSIXct` datetime in UTC, then derive a `date` and `hour` column for easier grouping.
- **Parties** field is a list, we concatenate into one row, to respect each event (timestamp) is one row.
:::

## Unnest events_clean

We create events_long table: one row per (event × party), with from_type and to_type classified (person/agent/system/world).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 

# Step 2: Build events_long
events_long <- events_clean %>%
  mutate(from = map_chr(parties, ~ .x[1])) %>%
  unnest(parties) %>%
  rename(to = parties) %>%
  filter(from != to) %>%
  mutate(
    # Type
    from_type = case_when(
      str_starts(from, "person:")       ~ "person",
      str_starts(from, "Agent/person:") ~ "agent",
      str_starts(from, "agent:person:") ~ "agent",
      str_starts(from, "system:")       ~ "system",
      str_starts(from, "world:")        ~ "world",
      TRUE ~ "other"),
    to_type = case_when(
      str_starts(to, "person:")       ~ "person",
      str_starts(to, "Agent/person:") ~ "agent",
      str_starts(to, "agent:person:") ~ "agent",
      str_starts(to, "system:")       ~ "system",
      str_starts(to, "world:")        ~ "world",
      TRUE ~ "other"),
    # Person ID for name lookup
    from_pid = str_replace(from, "^(Agent/person:|agent:person:)", "person:"),
    to_pid   = str_replace(to,   "^(Agent/person:|agent:person:)", "person:")
  ) %>%
  left_join(person_lookup %>% select(person_id, person_name),
            by = c("from_pid" = "person_id")) %>%
  rename(from_name = person_name) %>%
  left_join(person_lookup %>% select(person_id, person_name),
            by = c("to_pid" = "person_id")) %>%
  rename(to_name = person_name) %>%
  mutate(
    from_name = coalesce(from_name, from),
    to_name   = coalesce(to_name, to)
  ) %>%
  select(event_id, datetime, date, hour, short_name,
         from, from_type, from_name,
         to, to_type, to_name,
         parties_str,
         action, target, task, person, subject, status,
         file, content, forum, topic, advice, target_agent)
glimpse(events_long)
```

::: callout-note
What we have done:

- **Unnest parties field**: to show relationship from whom to whom. Agent/person and agent:person share the same treatment =\> tranform to agent type. We have analyzed it in details but do not want to show here for the sake of concise and smooth flow (agent:person only do read_file task while Agent/person do all remaining tasks).
- **From and to** show nice human name (john_windward becomes John Windward).
:::

## Extract graph elements from events_long

### Node entities

```{r}
#| code-fold: false
mc2_nodes <- bind_rows(
  events_long %>% select(id = from, type = from_type, label = from_name),
  events_long %>% select(id = to,   type = to_type,   label = to_name)
) %>%
  distinct(id, .keep_all = TRUE)
mc2_nodes
```

### Edge/Link entities

```{r}
#| code-fold: false

mc2_edges <- events_long %>%
  count(from_name, to_name, short_name, name = "weight")
glimpse(mc2_edges)
```

::: callout-note
- `weight` here counts how many times that `from-to` pair appears in `events_long.`
:::

## Sanity checking

### Sanity-check the time range

The challenge brief says the anomalous post occurs in **17** **May 2046**. A quick min/max confirms the data matches that.

```{r}
#| code-fold: false
events_long %>% 
  summarise(
    n_events   = n(),
    first_time = min(datetime),
    last_time  = max(datetime),
    n_days     = n_distinct(date)
  )
```

So dataset ranges among 67 days, from 8 May to 16 July, covering the checking point (17 May).

### Duplicates

There should be one row per `event_id`. We verify.

```{r}
events_clean %>% 
  summarise(
    n_rows    = n(),
    n_unique  = n_distinct(event_id),
    n_dup_ids = n() - n_distinct(event_id)
  )
```

Yes, each row represents for one event.

### Distribution of event types

`short_name` is the verb of each event: what was done. Counting them tells us which actions dominate the log and which are rare. Rare verbs are often the most diagnostic for an anomaly investigation.

```{r}
#| fig-width: 10
#| fig-height: 6

event_type_counts <- events_long %>% 
  count(short_name, sort = TRUE)

event_type_counts %>%
  ggplot(aes(x = reorder(short_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Distribution of Event Types", x = NULL, y = "Count") +
  theme_minimal()
```

::: callout-note
- There are some events seem be duplicated. Eg: post_saidit and saidit_post maybe one event (due to the inconsistency of name convention) or two different stages. We will confirm in further investigation.

- Numbers of delete_file, read_file, create_file and check_in are quite equal.
:::

### Missingness in `details`

`details` is a nested data frame: most of its columns are populated only for the event types that need them. The summary below tells us which detail-fields actually carry information.

```{r}
mc2_events$details %>% 
  summarise(across(where(is.character), ~ sum(!is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "detail_field", values_to = "non_na") %>% 
  arrange(desc(non_na)) %>%
  print(n = Inf)
```

::: callout-note
- target, task, person and action are four fields conveying the most information. They are unnested in events_long table already.
:::

## Outputs of this preprocessing stage

By the end of this notebook we have, ready for the analysis & visualisation stage:

- `org_nodes`: all entities in the organisation chart (company, departments, teams, persons).
- `org_edges`: hierarchical relationships (contains, led_by) between org entities.
- `person_lookup`: every person in Tenant Thread mapped to their name, built from org_chart.
- `events_flat`: one row per event, with proper `datetime`, `date`, `hour` and all detail fields unnested.
- `events_long`: one row per (event × from-to pair), with `from_type`, `to_type` and human-readable names.
- `mc2_nodes`: distinct actor entities (id, type, label) extracted from events_long.
- `mc2_edges`: aggregated interaction weights between actor pairs, ready for network visualisation.

```{mermaid}
%%| fig-cap: "Data flow: from raw JSON files to the enriched events table"
%%| code-fold: true
%%| code-summary: "Show the code"

graph TD
    A[("mc2_data.json")] --> B["mc2_data"]
    C[("org_chart.json")] --> D["org_chart"]

    D --> E["org_nodes"]
    D --> F["org_edges"]
    F --> G["person_lookup"]

    B --> H["events_flat"]
    H --> I["events_long"]
    G --> I

    I --> K["mc2_nodes"]
    I --> L["mc2_edges"]

    style A fill:#d0d0d0,stroke:#999
    style C fill:#d0d0d0,stroke:#999
    style B fill:#d0d0d0,stroke:#999
    style D fill:#d0d0d0,stroke:#999
    style E fill:#ffbb78,stroke:#ff7f00
    style F fill:#ffbb78,stroke:#ff7f00
    style G fill:#aec7e8,stroke:#377eb8
    style H fill:#aec7e8,stroke:#377eb8
    style I fill:#ffbb78,stroke:#ff7f00
    style K fill:#ffbb78,stroke:#ff7f00
    style L fill:#ffbb78,stroke:#ff7f00
```

::: callout-note
- **Grey nodes** are raw source files and parsed JSON objects.
- **Blue nodes** are intermediate processed tables.
- **Orange nodes** are the final outputs, used as input for further analysis and visualisation.
- Arrows are labelled with the **operation** that produced the next table.
:::

# Data Exploration & Visualization

## Solving Question 1

### Who is John Windward?

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Walk UP from John to find his department chain
target_id <- "person:john_windward"

# Find all ancestors (edges where target or his/her parent is the target)
target_ancestors <- vis_edges %>%
  filter(to == target_id) %>%
  pull(from)

# Go one more level up (department → company)
target_grandparents <- vis_edges %>%
  filter(to %in% target_ancestors) %>%
  pull(from)

# Find siblings — other people under the same parent(s)
target_siblings <- vis_edges %>%
  filter(from %in% target_ancestors) %>%
  pull(to)

# Combine all related node IDs
target_neighborhood <- unique(c(target_id, target_ancestors, target_grandparents, target_siblings))

# Filter nodes and edges
target_nodes <- vis_nodes %>%
  filter(id %in% target_neighborhood) %>%
  mutate(
    # Highlight target
    color.border = if_else(id == target_id, "red", NA_character_),
    borderWidth = if_else(id == target_id, 4, 1),
    font.size = if_else(id == target_id, 18, 14),
    size = if_else(id == target_id, 35, 20)
  )

target_edges <- vis_edges %>%
  filter(from %in% target_neighborhood, to %in% target_neighborhood)

# Draw the focused chart
visNetwork(target_nodes, target_edges,
           main = "Position in Tenant Thread")%>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 150,
    nodeSpacing = 120
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.2) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

rm(target_id, target_ancestors)
```

So John Windward belongs to Customer Support and is the lead of this Department.

### What all events John Windward or his agents did?

This is the core mapping table - one row per (event × actor) where the actor is either John himself or an agent attributed to him.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 

jw_events_view <- events_clean %>%
  filter(str_detect(parties_str, "john_windward"))

jw_events_view %>% 
  select(-parties) %>%
  datatable(
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE,
    caption = "John Windward & his agents: all logged actions, in chronological order"
  )
```

::: callout-note
- MC2 stated that: "The anomalous posting was sent to SaidIt by John Windward on May 17, 2046 at 4:21am". But from what we see, on 17 May 2046, **the post actually happened on 11:21**. It maybe due to local time difference. On preprocessing stage, we already convert original Unix timestamp to UTC (Coordinated Universal Time).
:::

### What all events of John Windward and his agents do related to the anomalous post?

Inspired by Gantt chart, we will modify with dot timeline with grouped bars by day to somehow visualize all actitivies as well as focus on target actions.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

post_time <- ymd_hms("2046-05-17 11:21:14", tz = "UTC")

# ============================================================
# 1. Prepare data — group events by date and action type
# ============================================================

jw_timeline <- jw_events_view %>%
  mutate(
    date = as_date(datetime),
    action_group = case_when(
      short_name %in% c("saidit_post", "post_saidit", 
                         "saidit_post_check") ~ "SaidIT post",
      short_name %in% c("post_flex", "flex_post") ~ "Flex post",
      short_name %in% c("create_file", "read_file", "delete_file",
                         "access_files", "list_files") ~ "File operation",
      short_name %in% c("sent", "received", "access_email", 
                         "send_email", "check_email") ~ "Email",
      short_name %in% c("queue_subordinate_task", "assign_agent_task",
                         "ask_agent") ~ "Agent task",
      short_name %in% c("give_advice", "suggest_contacts") ~ "Advice",
      short_name %in% c("propose_meeting", "enter_room",
                         "check_access", "check_in") ~ "Meeting/access",
      TRUE ~ "Other"
    )
  )

# Compute daily bars: first and last event per action_group per date
daily_bars <- jw_timeline %>%
  group_by(date, action_group) %>%
  summarise(
    start = min(datetime),
    end   = max(datetime),
    n     = n(),
    .groups = "drop"
  ) %>%
  # Ensure bars have minimum width for visibility
  mutate(
    end = if_else(start == end, start + minutes(30), end)
  )

# ============================================================
# 2. Color palette
# ============================================================

action_colors <- c(
  "SaidIT post" = "#e41a1c",
  "Flex post"            = "#f781bf",
  "File operation"       = "#ff7f00",
  "Email"                = "#377eb8",
  "Agent task"           = "#984ea3",
  "Advice"               = "#4daf4a",
  "Meeting/access"       = "#a65628",
  "Other"                = "#bdbdbd"
)

# ============================================================
# 4. Interactive plotly version with range slider
# ============================================================

# Build segments for plotly (one trace per action group)
fig <- plot_ly()

groups <- unique(daily_bars$action_group)
for (i in seq_along(groups)) {
  grp <- groups[i]
  d <- daily_bars %>% filter(action_group == grp)
  
  for (j in seq_len(nrow(d))) {
    fig <- fig %>%
      add_trace(
        x = c(d$start[j], d$end[j]),
        y = c(grp, grp),
        type = "scatter",
        mode = "lines",
        line = list(
          color = action_colors[grp],
          width = 20
        ),
        hovertext = paste0(
          "<b>", grp, "</b>",
          "<br>Date: ", d$date[j],
          "<br>Events: ", d$n[j],
          "<br>", format(d$start[j], "%H:%M"), " – ", 
          format(d$end[j], "%H:%M")
        ),
        hoverinfo = "text",
        showlegend = (j == 1),
        name = grp,
        legendgroup = grp
      )
  }
}

fig <- fig %>%
  layout(
    title = list(
      text = "John Windward - Activity Timeline",
      font = list(size = 15),
      width = 1500,
      height = 800
    ),
    hoverlabel = list(
      font = list(size = 8)
    ),
    xaxis = list(
      tickfont = list(size = 9),
      range = c(
        post_time - days(3),   # 3 days before post
        post_time + days(1)    # 1 day after post
      ),
      rangeslider = list(visible = TRUE, thickness = 0.06),
      rangeselector = list(
        buttons = list(
          list(count = 1,  label = "1d",  step = "day",  stepmode = "backward"),
          list(count = 7,  label = "7d",  step = "day",  stepmode = "backward"),
          list(count = 14, label = "14d", step = "day",  stepmode = "backward"),
          list(step = "all", label = "All")
        )
      )
    ),
    yaxis = list(
      title = "",
      categoryorder = "array",
      categoryarray = rev(names(action_colors))
    ),
    # Red line at anomalous post
    shapes = list(
      list(
        type = "line",
        x0 = post_time, x1 = post_time,
        y0 = 0, y1 = 1, yref = "paper",
        line = list(color = "red", width = 2, dash = "dash")
      )
    ),
    annotations = list(
      list(
        x = post_time, y = 1.05, yref = "paper",
        text = "Anomalous Post",
        showarrow = FALSE,
        font = list(color = "red", size = 11, family = "Arial Black")
      )
    ),
    legend = list(
      orientation = "h", 
      y = -0.45,
      font = list(size = 8),
      itemwidth = 30
      ),
    margin = list(b = 120, l = 150)
  )

fig
```

::: callout-note
Insights from this chart:

- On all days having SaidIT posts (13 May, 31 May, 4 June, 18 June and 7 July), there are always Advice actions: "give_advice", "suggest_contacts" happening before SaidIT actions. Meanwhile on target day 17 May 2046, **there is no Advice action before SaidIT post.**
:::

### What all events of John and his agent did on 17 May?

Now we need to take a deeper look on SaidIT posting day (17 May 2046).

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# local variables
target_ppl <- "john_windward"
target_date <- "2046-05-17"

events_clean %>%
  filter(
    str_detect(parties_str, target_ppl),
    as_date(datetime) == as_date(target_date)
    
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = paste("Events of", target_ppl, "on", target_date)
  )

# remove local variables
rm(target_ppl, target_date)
```

::: callout-note
We can see important information from this table above:

- Before event saidit_post (2046-05-17T11:21:15Z, event_id = 373902), there was an event **queue_subordinate_task** (2046-05-17T11:21:13Z, event_id = 373893), which is reading file from agent of Chloe Ballet to the agent of John Windward.
- Then, there was event **saidit_post_check** (2046-05-17T11:21:14Z, event_id = 373899) without any content (which means saidit_post_check checks nothing?)
- Event **saidit_post** (2046-05-17T11:21:15Z) read file: SwiftWren.txt from content_source from forum: general.
- Then two **delete_file** actions happened (2046-05-17T11:21:16Z, event_id = 373909 and 2046-05-17T11:21:17Z, event_id = 373913) deleted SwiftWren_further_instructions.md and SwiftWren.txt. Dangerously, it deleted files on file_system directly.

=\> **Questions to be raised:**

- All of these events continuously occured after each second. **The whole sequence is that** firstly agent of John Windward read SwiftWren.txt, checked nothing and post the content from this file, then deleted this file's instruction and this file itself.
- SwiftWren_further_instructions.md maybe the conditional steps to use SwiftWren.txt but we do not know checking events read it before read SwiftWren.txt or not.
- This flow is highly problematic, **why it deleted the file content right after posting it?** It likely a "smoking-gun" sequence.
- Who is Chloe Ballet?
- **What is the content of SwiftWren.txt?** Whether we can find out the content of SwiftWren.txt is the question that will be answered in the Question 2.
- **Who created that file?** Who is SwiftWren? We will find out right now.
:::

### Which all events related to SwiftWren?

We filter Tenant Thread Organization Chart to find any staff named Swift Wren but see nothing. Therefore I will find out all events related to SwiftWren.

We find keyword in column target, content, source_field, file_saved from nested column details.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
events_clean %>%
  filter(
    str_detect(coalesce(target, ""), "SwiftWren") |
    str_detect(coalesce(content, ""), "SwiftWren")
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = "All SwiftWren-related events"
  )
```

::: callout-note
- So, SwiftWren.txt file was **created by agent of Emma Harbor** on 2046-05-09T15:02:01Z (roughly 8 days before the SaidIT post day). One second later, this agent also read SwiftWren_further_instructions.md (likely the instructioin to use SwiftWren.txt).
- **Content column shows nothing for all events.**
:::

### Who is Emma Harbor?

```{r}
# Walk UP from John to find his department chain
target_id <- "person:emma_harbor"

# Find all ancestors (edges where target or his/her parent is the target)
target_ancestors <- vis_edges %>%
  filter(to == target_id) %>%
  pull(from)

# Go one more level up (department → company)
target_grandparents <- vis_edges %>%
  filter(to %in% target_ancestors) %>%
  pull(from)

# Find siblings — other people under the same parent(s)
target_siblings <- vis_edges %>%
  filter(from %in% target_ancestors) %>%
  pull(to)

# Combine all related node IDs
target_neighborhood <- unique(c(target_id, target_ancestors, target_grandparents, target_siblings))

# Filter nodes and edges
target_nodes <- vis_nodes %>%
  filter(id %in% target_neighborhood) %>%
  mutate(
    # Highlight target
    color.border = if_else(id == target_id, "red", NA_character_),
    borderWidth = if_else(id == target_id, 4, 1),
    font.size = if_else(id == target_id, 18, 14),
    size = if_else(id == target_id, 35, 20)
  )

target_edges <- vis_edges %>%
  filter(from %in% target_neighborhood, to %in% target_neighborhood)

# Draw the focused chart
visNetwork(target_nodes, target_edges,
           main = "Position in Tenant Thread")%>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 150,
    nodeSpacing = 120
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.2) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

rm(target_id, target_ancestors)
```

::: callout-note
- **Emma Harbor is CFO,** one of three C-suite employers, which means she likely knows important information or even top secrets of this TenantThread company, especially the deal **"Project HarborCrest"** (mentioned in MC1).
- There is **no direct link among Emma Harbor and John Windward** but the agent of John can access to files that agent of Emma created on a **shared system (file_system).**
- Agent of Emma stored files on the shared system without protection, which agent of John or agent of **any other Department leads seem easily access** to.
- One small suspicious point: is **"Harbor" Crest deal relevant to Emma "Harbor"**. Worth flagging as a finding for further investigation.
:::

### What all events Emma Harbor and her agents did on 9 May?

Because Emma did a number of actions in a day, we filter targeting hour at 15.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# local variables
target_ppl <- "emma_harbor"
target_date <- "2046-05-09"
target_hour <- c(14, 15)

events_clean %>%
  filter(
    str_detect(parties_str, target_ppl),
    as_date(datetime) == as_date(target_date),
    hour(datetime) %in% target_hour
    
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = paste("Events of", target_ppl, "on", target_date,
                    "- hours", paste(target_hour, collapse = "-"))
  )

# remove local variables
rm(target_ppl, target_date, target_hour)
```

::: callout-note
Insights from this table above:

- Before agent of Emma created SwiftWren.txt file, Emma personally assigned her agent to access_file (event_id = 21196).
- Then, her agent read file meeting_notes.doc (event_id=21201), then after 1 second, it created SwiftWren.txt file (event_id = 21208).
- Now, it confirms our suspicious question. SwiftWren.txt is created based on the content of her meeting.
:::

```{}
```

# References

::: {#refs}
:::

```{}
```

```{}
```

```{}
```

```{}
```

```{}
```

```{}
```
