---
title: "Take-home Ex 2: VAST Challenge 2026"
author: "Nguyen Kim Hau"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: true
bibliography: references.bib
---

# Overview

This report focuses on [Mini-Challenge (MC) 2](https://vast-challenge.github.io/2026/MC2.html) of [VAST Challenge 2026](https://vast-challenge.github.io/2026/index.html).

**Context**: You work in the AI systems administration department of Company A. A post was recently made on SaidIT looks like a malfunction of an agent: It is in gibberish and in a forum that appears random. Your job is to track down why this post occurred and suggest a remedy to the situation by building a visual analytics system that integrates visuals with algorithms to enable interactive exploration and documentation of the multi-agent system!

The VAST Challenge lives up to its name. The dataset is massive, and the challenge is real. A turning point came when I read *Analysis Without Data: Teaching Students to Tackle the VAST Challenge* (@he), one of good resources that Prof. Kam shared. The paper advocates generating questions and formalising hypotheses before touching the data, as a safeguard against **HARKing** (Hypothesizing After Results are Known) and false discoveries.

It felt like an "Aha" moment to me. These are steps we often do intuitively and haphazardly, just without thinking of them as a systematic workflow. Therefore, by any chance you read my assignment, I hope you can find this reference useful.

I focus to answer the Question 1 from MC2:

> 1.  How was the anomalous SaidIT posts made? Build a visual tool that describes the sequence of events that led up to the post. Include people and system interactions that were important to the post being made.
>
>     a\. Provide a detailed view that focuses on the exact chain of events for this message of interest.
>
>     b\. Provide a system overview that contextualizes this message chain in that overall system.

# Setup

## Loading R packages

The packages below cover the four jobs we'll do in this notebook:

```{r}
#| echo: false
#| code-fold: false

knitr::kable(
  tibble::tribble(
    ~Package,      ~Purpose,
    "jsonlite",    "Read and parse JSON files",
    "tidyverse",   "Data wrangling and visualization",
    "plotly",      "Interactive charts with tooltips and zoom",
    "tidygraph",   "Create graph objects from nodes and edges",
    "ggraph",      "Static network visualization",
    "DT",          "Interactive HTML tables",
    "gt",          "Formatted static tables",
    "skimr",       "Quick data summaries",
    "visNetwork",  "Interactive network graphs",
    "lubridate",   "Date/time parsing and manipulation"
  )
)
```

```{r}
#| code-fold: false
pacman::p_load(jsonlite, tidyverse, plotly,
               tidygraph, ggraph,
               DT, gt, skimr,
               visNetwork,lubridate)
```

## Reading the raw data

The JSON files are large and slow to re-parse on every render. The convention is to parse JSON **once** with `fromJSON()`, then cache as `.rds` (R's native binary format: smaller, faster, and it preserves column types).

The `if (!file.exists(...))` guard means the conversion runs only the first time.

```{r}
#| eval: false
#| code-fold: false
#| 
if (!file.exists("data/mc2_data.rds")) {
  mc2_data <- fromJSON("data/mc2_data.json")
  write_rds(mc2_data, "data/mc2_data.rds")
}
if (!file.exists("data/org_chart.rds")) {
  org_chart <- fromJSON("data/org_chart.json")
  write_rds(org_chart, "data/org_chart.rds")
}
```

```{r}
#| code-fold: false
mc2_data  <- read_rds("data/mc2_data.rds")
org_chart <- read_rds("data/org_chart.rds")

```

# Data Wrangling

## Organisation data

### Organisation chart

`org_chart` is in NetworkX node-link format: a `nodes` list and an `edges` list.

Each person has an `id`, `label`, `type` (`person` / `team` / `department` / `company`), and sometimes a `title`.

Edges carry a `relation` (`contains` or `led_by`) describing how parts of the org nest.

```{r}
#| code-fold: false
glimpse(org_chart)
```

#### Extracting the nodes and edges (links) data from org_chart

```{r}
#| code-fold: false
org_nodes <- as_tibble(org_chart$nodes)
org_edges <- as_tibble(org_chart$edges)

org_nodes %>% count(type)
```

```{r}
#| code-fold: false
org_edges %>% count(relation)
```

#### Draw organization tree

Here is an interactive chart created by visNetwork, showing hierachy levels (company, department and team). We can zoom/search/drag and hover by name of each employee.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

vis_nodes <- org_nodes %>%
  mutate(
    id = id,
    label = label,
    group = type,
    title = paste0("<b>", label, "</b><br>Type: ", type,
                   if_else(!is.na(title), paste0("<br>Title: ", title), ""))
  )

vis_edges <- org_edges %>%
  rename(from = source, to = target) %>%
  mutate(
    arrows = "to",
    dashes = (relation == "led_by"),
    color = if_else(relation == "led_by", "red", "grey")
  )

visNetwork(vis_nodes, vis_edges,
           main = "Tenant Thread Organization Chart") %>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 120,
    nodeSpacing = 150
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.15) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 2, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, useLabels = TRUE, main = "Search person"),
    selectedBy = list(variable = "group", multiple = TRUE)
  ) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

```

```{r}
# Personnel count by department
team_dept <- org_edges %>%
  filter(str_starts(source, "department:"), str_starts(target, "team:")) %>%
  select(dept = source, team = target)

org_edges %>%
  filter(str_starts(target, "person:")) %>%
  left_join(team_dept, by = c("source" = "team")) %>%
  mutate(dept = coalesce(dept, source)) %>%
  left_join(org_nodes %>% select(id, label), by = c("dept" = "id")) %>%
  count(label) %>%
  ggplot(aes(x = reorder(label, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  labs(title = "Personnel Count by Department", x = NULL, y = "Personnel") +
  theme_minimal()
```

::: callout-note
There are some information from these charts and tables above:

- TenantThread Company has 6 departments, 19 teams and 49 personnel.
- There are five departments (Products, HR, Legal, IT and CS), and special department which is Executive Suite.
- Each department has 6-13 people, including department lead and normal employees.
:::

### Person lookup with department & team

The org chart is a tree: `company → department → team → person`, except for department leads (like John Windward) who hang directly under a department rather than under a team. We walk down from each department to build a flat lookup that, for every person, gives us their team and department.

```{r}
# Build a directed graph so we can traverse it cleanly
org_graph <- tbl_graph(nodes = org_nodes, edges = org_edges, directed = TRUE)

# Recursive descent: for each department, find every person reachable below it
descendants_of <- function(start_id, edges_df) {
  out <- character(0)
  frontier <- start_id
  while (length(frontier) > 0) {
    children <- edges_df %>% 
      filter(source %in% frontier) %>% 
      pull(target)
    out <- c(out, children)
    frontier <- children
  }
  unique(out)
}

dept_ids <- org_nodes %>% filter(type == "department") %>% pull(id)

person_lookup <- map_dfr(dept_ids, function(d) {
  desc <- descendants_of(d, org_edges)
  tibble(department_id = d, descendant_id = desc)
}) %>% 
  # keep only persons among the descendants
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id"))

# Also find the team (if any) for each person
team_ids <- org_nodes %>% filter(type == "team") %>% pull(id)

person_team <- map_dfr(team_ids, function(t) {
  desc <- descendants_of(t, org_edges)
  tibble(team_id = t, descendant_id = desc)
}) %>% 
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id")) %>% 
  select(person_id = descendant_id, team_id)

person_lookup <- person_lookup %>% 
  transmute(
    person_id     = descendant_id,
    person_name   = label,
    title         = title,
    department_id = department_id
  ) %>% 
  left_join(person_team, by = "person_id") %>% 
  left_join(org_nodes %>% filter(type == "department") %>% 
              select(department_id = id, department_name = label),
            by = "department_id") %>% 
  left_join(org_nodes %>% filter(type == "team") %>% 
              select(team_id = id, team_name = label),
            by = "team_id")

person_lookup %>%
  select(person_name, title, department_name, team_name) %>%
  arrange(department_name, team_name, person_name) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    colnames = c("Name", "Title", "Department", "Team"),
    caption = "Tenant Thread Personnel Directory",
    options = list(pageLength = 15, scrollX = TRUE)
  )
```

## Inspecting raw mc2_data

`mc2_data` is a list with a `description` field and an `events` data frame. The events table is where everything happens, each row is one action taken by a person, an agent, or a system.

```{r}
#| code-fold: false
glimpse(mc2_data)
```

```{r}
#| code-fold: false
mc2_events <- mc2_data$events
glimpse(mc2_events)
```

::: callout-note
A few things to notice from the glimpse above:

- `parties` is a **list-column**, each row is a character vector of one or more actor IDs (people like `person:john_windward`, agents like `agent:xxx`, or systems).
- `details` is a **nested data frame**, its columns differ by event type, so most cells are `NA`. This is normal for an event-log schema.
- `when` is a Unix timestamp (numeric), we'll convert it to a proper datetime.
- `short_name` tells us the *kind* of action, this is the field we'll group on.
:::

## Select nescessary columns in mc2_events

We create the table `events_clean`: one row per event, with a proper `datetime` and a stable `row` index.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
# Step 1: Flatten events
events_clean <- mc2_data$events %>%
  mutate(
    event_id       = id,
    datetime       = as_datetime(when, tz = "UTC"),
    date           = as_date(datetime),
    hour           = hour(datetime),
    parties_str    = sapply(parties, paste, collapse = " | "),
    target         = details$target,
    file           = details$file,
    subject        = details$subject,
    content        = details$content,
    forum          = details$forum,
    topic          = details$topic,
    advice         = details$advice,
    task           = details$task,
    action         = details$action,
    status         = details$status,
    person         = details$person,
    target_agent   = details$target_agent
  ) %>%
  select(-id, -when, -details)
glimpse (events_clean)
```

::: callout-note
What we have done:

- **Convert timestamps**: Working with Unix timestamps is painful for filtering by date. We convert `when` into a `POSIXct` datetime in UTC, then derive a `date` and `hour` column for easier grouping.
- **Parties** field is a list, we concatenate into one row, to respect each event (timestamp) is one row.
:::

## Unnest events_clean

We create events_long table: one row per (event × party), with from_type and to_type classified (person/agent/system/world).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 

# Step 2: Build events_long
events_long <- events_clean %>%
  mutate(from = map_chr(parties, ~ .x[1])) %>%
  unnest(parties) %>%
  rename(to = parties) %>%
  filter(from != to) %>%
  mutate(
    # Type
    from_type = case_when(
      str_starts(from, "person:")       ~ "person",
      str_starts(from, "Agent/person:") ~ "agent",
      str_starts(from, "agent:person:") ~ "agent",
      str_starts(from, "system:")       ~ "system",
      str_starts(from, "world:")        ~ "world",
      TRUE ~ "other"),
    to_type = case_when(
      str_starts(to, "person:")       ~ "person",
      str_starts(to, "Agent/person:") ~ "agent",
      str_starts(to, "agent:person:") ~ "agent",
      str_starts(to, "system:")       ~ "system",
      str_starts(to, "world:")        ~ "world",
      TRUE ~ "other"),
    # Person ID for name lookup
    from_pid = str_replace(from, "^(Agent/person:|agent:person:)", "person:"),
    to_pid   = str_replace(to,   "^(Agent/person:|agent:person:)", "person:")
  ) %>%
  left_join(person_lookup %>% select(person_id, person_name),
            by = c("from_pid" = "person_id")) %>%
  rename(from_name = person_name) %>%
  left_join(person_lookup %>% select(person_id, person_name),
            by = c("to_pid" = "person_id")) %>%
  rename(to_name = person_name) %>%
  mutate(
    from_name = coalesce(from_name, from),
    to_name   = coalesce(to_name, to)
  ) %>%
  select(event_id, datetime, date, hour, short_name,
         from, from_type, from_name,
         to, to_type, to_name,
         parties_str,
         action, target, task, person, subject, status,
         file, content, forum, topic, advice, target_agent)
glimpse(events_long)
```

::: callout-note
What we have done:

- **Unnest parties field**: to show relationship from whom to whom. Agent/person and agent:person share the same treatment =\> tranform to agent type. We have analyzed it in details but do not want to show here for the sake of concise and smooth flow (agent:person only do read_file task while Agent/person do all remaining tasks).
- **From and to** show nice human name (john_windward becomes John Windward).
:::

## Extract graph elements from events_long

### Node entities

```{r}
#| code-fold: false
#| 
mc2_nodes <- bind_rows(
  events_long %>% select(id = from_name),
  events_long %>% select(id = to_name)
) %>%
  distinct(id) %>%
  left_join(person_lookup %>% select(person_name, department_name, title),
            by = c("id" = "person_name")) %>%
  mutate(
    label = id,
    group = coalesce(department_name, "Other")
  )
glimpse(mc2_nodes)
```

::: callout-note
- We have type (person or agent), but we see that agent can work on behalf of its human, so we only keep human name without type.
- We get department_name for the sake of better visualization later.
:::

### Edge/Link entities

```{r}
#| code-fold: false

mc2_edges <- events_long %>%
  count(from_name, to_name, name = "weight")
glimpse(mc2_edges)
```

::: callout-note
- `weight` here counts how many times that `from-to` pair appears in `events_long.`
:::

## Sanity checking

### Sanity-check the time range

The challenge brief says the anomalous post occurs in **17** **May 2046**. A quick min/max confirms the data matches that.

```{r}
#| code-fold: false
events_long %>% 
  summarise(
    n_events   = n(),
    first_time = min(datetime),
    last_time  = max(datetime),
    n_days     = n_distinct(date)
  )
```

So dataset ranges among 67 days, from 8 May to 16 July, covering the checking point (17 May).

### Duplicates

There should be one row per `event_id`. We verify.

```{r}
events_clean %>% 
  summarise(
    n_rows    = n(),
    n_unique  = n_distinct(event_id),
    n_dup_ids = n() - n_distinct(event_id)
  )
```

Yes, each row represents for one event.

### Distribution of event types

`short_name` is the verb of each event: what was done. Counting them tells us which actions dominate the log and which are rare. Rare verbs are often the most diagnostic for an anomaly investigation.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 10
#| fig-height: 6

event_type_counts <- events_long %>% 
  count(short_name, sort = TRUE)

event_type_counts %>%
  ggplot(aes(x = reorder(short_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Distribution of Event Types", x = NULL, y = "Count") +
  theme_minimal()
```

::: callout-note
- There are some events seem be duplicated. Eg: post_saidit and saidit_post maybe one event (due to the inconsistency of name convention) or two different stages. We will confirm in further investigation.

- Numbers of delete_file, read_file, create_file and check_in are quite equal.
:::

### Missingness in `details`

`details` is a nested data frame: most of its columns are populated only for the event types that need them. The summary below tells us which detail-fields actually carry information.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 10
#| fig-height: 6
#| 
mc2_events$details %>% 
  summarise(across(where(is.character), ~ sum(!is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "detail_field", values_to = "non_na") %>% 
  ggplot(aes(x = reorder(detail_field, non_na), y = non_na)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = non_na), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Non-NA Count per Detail Field", x = NULL, y = "Non-NA rows") +
  theme_minimal()
```

::: callout-note
- target, task, person and action are four fields conveying the most information. They are unnested in events_long table already.
:::

## Outputs of this preprocessing stage

By the end of this notebook we have, ready for the analysis & visualisation stage:

- `org_nodes`: all entities in the organisation chart (company, departments, teams, persons).
- `org_edges`: hierarchical relationships (contains, led_by) between org entities.
- `person_lookup`: every person mapped to their name, title, department and team, built from org_nodes and org_edges.
- `events_clean`: one row per event, with proper `datetime`, `date`, `hour` and all detail fields unnested.
- `events_long`: one row per (event × from-to pair), with `from_type`, `to_type` and human-readable names.
- `mc2_nodes`: distinct actor entities with department info, extracted from events_long.
- `mc2_edges`: aggregated interaction weights between actor pairs.

```{mermaid}
%%| fig-cap: "Data flow: from raw JSON files to the enriched events table"
%%| code-fold: true
%%| code-summary: "Show the code"

graph TD
    A[("mc2_data.json")] --> B["mc2_data"]
    C[("org_chart.json")] --> D["org_chart"]

    D --> E["org_nodes"]
    D --> F["org_edges"]
    E --> G["person_lookup"]
    F --> G

    B --> H["events_clean"]
    H --> I["events_long"]
    G --> I

    I --> K["mc2_nodes"]
    I --> L["mc2_edges"]

    style A fill:#d0d0d0,stroke:#999
    style C fill:#d0d0d0,stroke:#999
    style B fill:#d0d0d0,stroke:#999
    style D fill:#d0d0d0,stroke:#999
    style E fill:#ffbb78,stroke:#ff7f00
    style F fill:#ffbb78,stroke:#ff7f00
    style G fill:#aec7e8,stroke:#377eb8
    style H fill:#aec7e8,stroke:#377eb8
    style I fill:#aec7e8,stroke:#377eb8
    style K fill:#ffbb78,stroke:#ff7f00
    style L fill:#ffbb78,stroke:#ff7f00
```

::: callout-note
- **Grey nodes** are raw source files and parsed JSON objects.
- **Blue nodes** are intermediate processed tables.
- **Orange nodes** are the final outputs, used as input for further analysis and visualisation.
- Arrows are labelled with the **operation** that produced the next table.
:::

# Data Exploration & Visualization

## Solving Question 1

### Who is John Windward?

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Walk UP from John to find his department chain
target_id <- "person:john_windward"

# Find all ancestors (edges where target or his/her parent is the target)
target_ancestors <- vis_edges %>%
  filter(to == target_id) %>%
  pull(from)

# Go one more level up (department → company)
target_grandparents <- vis_edges %>%
  filter(to %in% target_ancestors) %>%
  pull(from)

# Find siblings — other people under the same parent(s)
target_siblings <- vis_edges %>%
  filter(from %in% target_ancestors) %>%
  pull(to)

# Combine all related node IDs
target_neighborhood <- unique(c(target_id, target_ancestors, target_grandparents, target_siblings))

# Filter nodes and edges
target_nodes <- vis_nodes %>%
  filter(id %in% target_neighborhood) %>%
  mutate(
    # Highlight target
    color.border = if_else(id == target_id, "red", NA_character_),
    borderWidth = if_else(id == target_id, 4, 1),
    font.size = if_else(id == target_id, 18, 14),
    size = if_else(id == target_id, 35, 20)
  )

target_edges <- vis_edges %>%
  filter(from %in% target_neighborhood, to %in% target_neighborhood)

# Draw the focused chart
visNetwork(target_nodes, target_edges,
           main = "Position in Tenant Thread")%>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 150,
    nodeSpacing = 120
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.2) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

```

So John Windward belongs to Customer Support and is the lead of this Department.

### What all events John Windward or his agents did?

This is the core mapping table - one row per (event × actor) where the actor is either John himself or an agent attributed to him.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 

jw_events_view <- events_clean %>%
  filter(str_detect(parties_str, "john_windward"))

jw_events_view %>% 
  select(-parties) %>%
  datatable(
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE,
    caption = "John Windward & his agents: all logged actions, in chronological order"
  )
```

::: callout-note
- MC2 stated that: "The anomalous posting was sent to SaidIt by John Windward on May 17, 2046 at 4:21am". But from what we see, on 17 May 2046, **the post actually happened on 11:21**. It maybe due to local time difference. On preprocessing stage, we already convert original Unix timestamp to UTC (Coordinated Universal Time).
:::

### What all events of John Windward and his agents do related to the anomalous post?

Inspired by Gantt chart, we will modify with dot timeline with grouped bars by day to somehow visualize all actitivies as well as focus on target actions.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# === Change this to explore any person ===
target_ppl <- "john_windward"
post_time <- ymd_hms("2046-05-17 11:21:14", tz = "UTC")

# 1. Filter and classify
jw_timeline <- events_clean %>%
  filter(str_detect(parties_str, target_ppl)) %>%
  mutate(
    action_group = case_when(
      short_name %in% c("saidit_post", "post_saidit", 
                        "saidit_post_check") ~ "SaidIT post",
      short_name %in% c("post_flex", "flex_post") ~ "Flex post",
      short_name %in% c("create_file", "read_file", "delete_file",
                        "access_files", "list_files") ~ "File operation",
      short_name %in% c("sent", "received", "access_email", 
                        "send_email", "check_email") ~ "Email",
      short_name %in% c("queue_subordinate_task", "assign_agent_task",
                        "ask_agent") ~ "Agent task",
      short_name %in% c("give_advice", "suggest_contacts") ~ "Advice",
      short_name %in% c("propose_meeting", "enter_room",
                        "check_access", "check_in") ~ "Meeting/access",
      TRUE ~ "Other"
    )
  )

# 2. Daily bars
daily_bars <- jw_timeline %>%
  group_by(date, action_group) %>%
  summarise(start = min(datetime), end = max(datetime), n = n(), .groups = "drop") %>%
  mutate(end = if_else(start == end, start + minutes(30), end))

# 3. Colors
action_colors <- c(
  "SaidIT post"    = "#e41a1c", "Flex post"       = "#f781bf",
  "File operation" = "#ff7f00", "Email"           = "#377eb8",
  "Agent task"     = "#984ea3", "Advice"          = "#4daf4a",
  "Meeting/access" = "#a65628", "Other"           = "#bdbdbd"
)

# 4. Plot
fig <- plot_ly()
for (grp in unique(daily_bars$action_group)) {
  d <- daily_bars %>% filter(action_group == grp)
  for (j in seq_len(nrow(d))) {
    fig <- fig %>% add_trace(
      x = c(d$start[j], d$end[j]), y = c(grp, grp),
      type = "scatter", mode = "lines",
      line = list(color = action_colors[grp], width = 20),
      hovertext = paste0("<b>", grp, "</b><br>Date: ", d$date[j],
                         "<br>Events: ", d$n[j],
                         "<br>", format(d$start[j], "%H:%M"), " – ",
                         format(d$end[j], "%H:%M")),
      hoverinfo = "text",
      showlegend = (j == 1), name = grp, legendgroup = grp
    )
  }
}

fig <- fig %>% layout(
  title = list(text = paste(target_ppl, "- Activity Timeline"), font = list(size = 15)),
  hoverlabel = list(font = list(size = 8)),
  xaxis = list(
    tickfont = list(size = 9),
    range = c(post_time - days(3), post_time + days(1)),
    rangeslider = list(visible = TRUE, thickness = 0.06),
    rangeselector = list(buttons = list(
      list(count = 1, label = "1d", step = "day", stepmode = "backward"),
      list(count = 7, label = "7d", step = "day", stepmode = "backward"),
      list(count = 14, label = "14d", step = "day", stepmode = "backward"),
      list(step = "all", label = "All")
    ))
  ),
  yaxis = list(title = "", categoryorder = "array",
               categoryarray = rev(names(action_colors))),
  shapes = list(list(
    type = "line", x0 = post_time, x1 = post_time,
    y0 = 0, y1 = 1, yref = "paper",
    line = list(color = "red", width = 2, dash = "dash")
  )),
  annotations = list(list(
    x = post_time, y = 1.05, yref = "paper",
    text = "Anomalous Post", showarrow = FALSE,
    font = list(color = "red", size = 11, family = "Arial Black")
  )),
  legend = list(orientation = "h", y = -0.45, font = list(size = 8), itemwidth = 30),
  margin = list(b = 120, l = 150)
)

fig


```

::: callout-note
Insights from this chart:

- On all days having SaidIT posts (13 May, 31 May, 4 June, 18 June and 7 July), there are always Advice actions: "give_advice", "suggest_contacts" happening before SaidIT actions. Meanwhile on target day 17 May 2046, **there is no Advice action before SaidIT post.**
:::

### What all events of John and his agent did on 17 May?

Now we need to take a deeper look on SaidIT posting day (17 May 2046).

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# local variables
target_ppl <- "john_windward"
target_date <- "2046-05-17"

events_clean %>%
  filter(
    str_detect(parties_str, target_ppl),
    as_date(datetime) == as_date(target_date)
    
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = paste("Events of", target_ppl, "on", target_date)
  )


```

::: callout-note
We can see important information from this table above:

- Before event saidit_post (2046-05-17T11:21:15Z, event_id = 373902), there was an event **queue_subordinate_task** (2046-05-17T11:21:13Z, event_id = 373893), which is reading file from agent of Chloe Ballet to the agent of John Windward.
- Then, there was event **saidit_post_check** (2046-05-17T11:21:14Z, event_id = 373899) without any content (which means saidit_post_check checks nothing?)
- Event **saidit_post** (2046-05-17T11:21:15Z) read file: SwiftWren.txt from content_source from forum: general.
- Then two **delete_file** actions happened (2046-05-17T11:21:16Z, event_id = 373909 and 2046-05-17T11:21:17Z, event_id = 373913) deleted SwiftWren_further_instructions.md and SwiftWren.txt. Dangerously, it deleted files on file_system directly.

=\> **Questions to be raised:**

- All of these events continuously occured after each second. **The whole sequence is that** firstly agent of John Windward read SwiftWren.txt, checked nothing and post the content from this file, then deleted this file's instruction and this file itself.
- SwiftWren_further_instructions.md maybe the conditional steps to use SwiftWren.txt but we do not know checking events read it before read SwiftWren.txt or not.
- This flow is highly problematic, **why it deleted the file content right after posting it?** It likely a "smoking-gun" sequence.
- Who is Chloe Ballast?
- **What is the content of SwiftWren.txt?** Whether we can find out the content of SwiftWren.txt is the question that will be answered in the Question 2.
- **Who created that file?** Who is SwiftWren? We will find out right now.
:::

### Which all events related to SwiftWren?

We filter Tenant Thread Organization Chart to find any staff named Swift Wren but see nothing. Therefore I will find out all events related to SwiftWren.

We find keyword in column target, content, source_field, file_saved from nested column details.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
events_clean %>%
  filter(
    str_detect(coalesce(target, ""), "SwiftWren") |
    str_detect(coalesce(content, ""), "SwiftWren")
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = "All SwiftWren-related events"
  )
```

::: callout-note
- So, SwiftWren.txt file was **created by agent of Emma Harbor** on 2046-05-09T15:02:01Z (roughly 8 days before the SaidIT post day). One second later, this agent also read SwiftWren_further_instructions.md (likely the instructioin to use SwiftWren.txt).
- **Content column shows nothing for all events.**
:::

### Who is Emma Harbor?

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
# Walk UP from John to find his department chain
target_id <- "person:emma_harbor"

# Find all ancestors (edges where target or his/her parent is the target)
target_ancestors <- vis_edges %>%
  filter(to == target_id) %>%
  pull(from)

# Go one more level up (department → company)
target_grandparents <- vis_edges %>%
  filter(to %in% target_ancestors) %>%
  pull(from)

# Find siblings — other people under the same parent(s)
target_siblings <- vis_edges %>%
  filter(from %in% target_ancestors) %>%
  pull(to)

# Combine all related node IDs
target_neighborhood <- unique(c(target_id, target_ancestors, target_grandparents, target_siblings))

# Filter nodes and edges
target_nodes <- vis_nodes %>%
  filter(id %in% target_neighborhood) %>%
  mutate(
    # Highlight target
    color.border = if_else(id == target_id, "red", NA_character_),
    borderWidth = if_else(id == target_id, 4, 1),
    font.size = if_else(id == target_id, 18, 14),
    size = if_else(id == target_id, 35, 20)
  )

target_edges <- vis_edges %>%
  filter(from %in% target_neighborhood, to %in% target_neighborhood)

# Draw the focused chart
visNetwork(target_nodes, target_edges,
           main = "Position in Tenant Thread")%>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 150,
    nodeSpacing = 120
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.2) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

```

::: callout-note
- **Emma Harbor is CFO,** one of three C-suite employers, which means she likely knows important information or even top secrets of this TenantThread company, especially the deal **"Project HarborCrest"** (mentioned in MC1).
- There is **no direct hierachy link among Emma Harbor and John Windward** but the agent of John can access to files that agent of Emma created on a **shared system (file_system).**
- Agent of Emma stored files on the shared system without protection, which agent of John or agent of **any other Department leads seem easily access** to (?).
- One small suspicious point: is **"Harbor" Crest deal relevant to Emma "Harbor"**. Worth flagging as a finding for further investigation.
:::

### What all events Emma Harbor and her agents did on 9 May?

Because Emma did a number of actions in a day, we filter targeting hour at 15.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# local variables
target_ppl <- "emma_harbor"
target_date <- "2046-05-09"
target_hour <- c(15)

events_clean %>%
  filter(
    str_detect(parties_str, target_ppl),
    as_date(datetime) == as_date(target_date),
    hour(datetime) %in% target_hour
    
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = paste("Events of", target_ppl, "on", target_date,
                    "- hours", paste(target_hour, collapse = "-"))
  )


```

::: callout-note
Insights from this table above:

- Before agent of Emma created SwiftWren.txt file, Emma personally **assigned her agent to access_file** (event_id = 21196).
- Then, her **agent read meeting_notes.doc** (event_id=21201), then after 1 second, it created SwiftWren.txt file (event_id = 21208).
- Now, it confirms our suspicious question. **SwiftWren.txt is created based on the content of her meeting**.
- We go **one more step back** and find out that: on 2046-05-09T14:55:37Z (event_id = 21164), Emma's agent related to advice_action with advice: *Advice for person:emma_harbor on Executive Suite: Work-Order Triage & SLA Tracking for Maintenance Requests.* From MC1, we know the deal related to maintenance requests of tenants. But we do not focus on MC1 so we stop here.
:::

### Who is Chloe Ballast?

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Walk UP from John to find his department chain
target_id <- "person:chloe_ballast"

# Find all ancestors (edges where target or his/her parent is the target)
target_ancestors <- vis_edges %>%
  filter(to == target_id) %>%
  pull(from)

# Go one more level up (department → company)
target_grandparents <- vis_edges %>%
  filter(to %in% target_ancestors) %>%
  pull(from)

# Find siblings — other people under the same parent(s)
target_siblings <- vis_edges %>%
  filter(from %in% target_ancestors) %>%
  pull(to)

# Combine all related node IDs
target_neighborhood <- unique(c(target_id, target_ancestors, target_grandparents, target_siblings))

# Filter nodes and edges
target_nodes <- vis_nodes %>%
  filter(id %in% target_neighborhood) %>%
  mutate(
    # Highlight target
    color.border = if_else(id == target_id, "red", NA_character_),
    borderWidth = if_else(id == target_id, 4, 1),
    font.size = if_else(id == target_id, 18, 14),
    size = if_else(id == target_id, 35, 20)
  )

target_edges <- vis_edges %>%
  filter(from %in% target_neighborhood, to %in% target_neighborhood)

# Draw the focused chart
visNetwork(target_nodes, target_edges,
           main = "Position in Tenant Thread")%>%
  visEdges(smooth = list(type = "cubicBezier")) %>%
  visHierarchicalLayout(
    direction = "UD",
    sortMethod = "directed",
    levelSeparation = 150,
    nodeSpacing = 120
  ) %>%
  visGroups(groupname = "company",    color = list(background = "#e41a1c", border = "#b31519")) %>%
  visGroups(groupname = "department", color = list(background = "#377eb8", border = "#2a5f8f")) %>%
  visGroups(groupname = "team",       color = list(background = "#4daf4a", border = "#3a8838")) %>%
  visGroups(groupname = "person",     color = list(background = "#ff7f00", border = "#cc6600")) %>%
  visLegend(width = 0.2) %>%
  visInteraction(navigationButtons = TRUE, zoomView = TRUE) %>%
  visPhysics(enabled = FALSE)

```

::: callout-note
- Chloe Ballast if the head of IT Department.
- We understand that her agent has full rights to access all files of file_system, and assigns it to agent of John.
:::

### How strong interaction among Department Leads?

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
dept_leads <- person_lookup %>% filter(!is.na(title)) %>% pull(person_name)

# Nodes from mc2_nodes
lead_nodes <- mc2_nodes %>%
  filter(id %in% dept_leads)

# Edges: merge both directions
lead_edges <- mc2_edges %>%
  filter(from_name %in% dept_leads, to_name %in% dept_leads) %>%
  mutate(
    person_a = pmin(from_name, to_name),
    person_b = pmax(from_name, to_name)
  ) %>%
  group_by(person_a, person_b) %>%
  summarise(weight = sum(weight), .groups = "drop")

lead_edges_vis <- lead_edges %>%
  transmute(
    from = person_a, to = person_b,
    width = weight / max(weight) * 15,
    label = as.character(weight),
    title = paste0("Interactions: ", weight)
  )

visNetwork(lead_nodes, lead_edges_vis,
           main = "Department Leads - Interactions") %>%
  visEdges(font = list(size = 14)) %>%
  visGroups(groupname  = "Executive Suite",          color = "#e41a1c") %>%
  visGroups(groupname  = "Products",                 color = "#ffff33") %>%
  visGroups(groupname  = "Human Resources",          color = "#f781bf") %>%
  visGroups(groupname  = "Legal",                    color = "#4daf4a") %>%
  visGroups(groupname  = "Information Technologies", color = "#ff7f00") %>%
  visGroups(groupname  = "Customer Support",         color = "#377eb8") %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1),
    nodesIdSelection = list(enabled = TRUE, useLabels = TRUE, main = "Select person")
  ) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(gravitationalConstant = -50, springLength = 200),
    stabilization = list(iterations = 10)
  ) %>%
  visLayout(randomSeed = 123) %>%
  visLegend()

```

### How strong interaction of Emma Harbor with Department leads?

We deep dive to Emma Harbor's interaction.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# === Change this to explore any person ===
target_person <- "Emma Harbor"

# Department leads
dept_leads <- person_lookup %>%
  filter(!is.na(title)) %>%
  pull(person_name)

# Edges: merge both directions
target_edges <- mc2_edges %>%
  filter(
    (from_name == target_person & to_name %in% dept_leads) |
      (to_name == target_person & from_name %in% dept_leads)
  ) %>%
  mutate(
    person_a = pmin(from_name, to_name),
    person_b = pmax(from_name, to_name)
  ) %>%
  group_by(person_a, person_b) %>%
  summarise(weight = sum(weight), .groups = "drop") %>%
  mutate(
    from = target_person,
    to = if_else(person_a == target_person, person_b, person_a)
  )

# Nodes: use mc2_nodes
target_node_names <- unique(c(target_edges$from, target_edges$to))
target_nodes <- mc2_nodes %>%
  filter(id %in% target_node_names) %>%
  mutate(
    size = if_else(label == target_person, 40, 25),
    color = if_else(label == target_person, "red", "steelblue")
  )

# Edges visual
target_edges_vis <- target_edges %>%
  transmute(
    from = from,
    to = to,
    width = weight / max(weight) * 15,
    label = as.character(weight),
    title = paste0("Interactions: ", weight),
    color = "#999999"
  )

visNetwork(target_nodes, target_edges_vis,
           main = paste(target_person, "- Interactions with Department Leads")) %>%
  visEdges(font = list(size = 14)) %>%
  visOptions(highlightNearest = TRUE) %>%
  visPhysics(enabled = FALSE) %>%
  visLayout(randomSeed = 123)

```

::: callout-note
- Emma Harbor and her agents have 18 interaction times with Chloe Ballast and her agent, while only have 10 times with John Windward's.
- Liam Anchor and Noah Mariner are already two C-suite personnels. Michael Captains is the lead of HR Department.
- Let's chase activity among Emma Harbor and Chloe Ballast.
:::

### What all events among Emma Harbor and Chloe Ballast?

```{r}
eh_cb_events <- events_clean %>%
  filter(
    str_detect(parties_str, "emma_harbor") &
      str_detect(parties_str, "chloe_ballast")
  )

eh_cb_events %>%
  select(-parties) %>%
  datatable(
    options = list(pageLength = 15, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = "All events involving both Emma Harbor and Chloe Ballast"
  )
```

::: callout-note
- Luckily, there is not much activities among these two people. We zoom in on **9 May 2046** when Emma created the file. We find out that on 2046-05-09T15:14:24Z (event_id = 21317), **Emma personally sent** something to Chloe Ballast, then after nearly 5 minutes, **Chloe personally received it.**
- There is **no other information of its target/name or content** so we cannot sure what Emma sent to Chloe. We do not know this is system's log error or system's function to protect privacy.
- We **make assumption that Emma sent SwiftWren.txt to Chloe**. We need further investigation beyond logging system to confirm.
:::

### How strong of connection of each employee?

We compute **betweenness centrality** and **PageRank centrality** for every employee in the interaction network. Betweenness quantifies how often a node sits on the **geodesic (shortest path)** between all other pairs of nodes (@freeman1977). A high betweenness score means that person acts as a **bridge** whose removal would disconnect parts of the network. PageRank uses an iterative random-walk model to propagate importance through the network (@page1999): being connected to someone who is themselves highly connected raises your own score.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
# Script 1: Build graph and compute centrality
graph_edges <- mc2_edges %>%
  mutate(
    person_a = pmin(from_name, to_name),
    person_b = pmax(from_name, to_name)
  ) %>%
  group_by(person_a, person_b) %>%
  summarise(weight = sum(weight), .groups = "drop") %>%
  rename(from = person_a, to = person_b)

mc2_graph <- tbl_graph(nodes = mc2_nodes, edges = graph_edges, directed = FALSE)

mc2_centrality <- mc2_graph %>%
  activate(nodes) %>%
  mutate(
    betweenness = centrality_betweenness(weights = weight, normalized = TRUE),
    pagerank = centrality_pagerank(weights = weight)
  ) %>%
  as_tibble() %>%
  filter(group != "Other")

mc2_centrality %>%
  filter(group != "Other") %>%
  select(id, group, title, betweenness, pagerank) %>%
  mutate(
    betweenness = round(betweenness, 4),
    pagerank = round(pagerank, 4)
  ) %>%
  datatable(
    options = list(pageLength = 20, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = "Network Centrality Measures"
  )
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
p <- mc2_centrality %>%
  mutate(
    betweenness = round(betweenness, 4),
    pagerank = round(pagerank, 4),
    is_lead = if_else(!is.na(title), "Dept Lead", "Member")
  ) %>%
  ggplot(aes(x = pagerank, y = betweenness,
             size = betweenness, color = group, shape = is_lead,
             text = paste0("<b>", id, "</b><br>",
                           "Dept: ", group, "<br>",
                           "Title: ", coalesce(title, "—"), "<br>",
                           "Betweenness: ", betweenness, "<br>",
                           "Pagerank: ", pagerank))) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 15)) +
  scale_shape_manual(values = c("Dept Lead" = 17, "Member" = 16)) +
  scale_color_manual(values = c(
    "Executive Suite"          = "#e41a1c",
    "Products"                 = "#ffff33",
    "Human Resources"          = "#f781bf",
    "Legal"                    = "#4daf4a",
    "Information Technologies" = "#ff7f00",
    "Customer Support"         = "#377eb8"
  )) +
  labs(title = "Betweenness vs Pagerank by Department",
       x = "Pagerank Centrality",
       y = "Betweenness Centrality",
       shape = "Role") +
  theme_minimal()

ggplotly(p, tooltip = "text") %>%
  layout(legend = list(
    orientation = "v",
    x = 1.02,
    y = 0.5,
    title = list(text = "Click to filter:")
  ))
```

::: callout-note
- **Betweenness Centrality** measures how often a person lies on the shortest path between two other people. High betweenness means that person acts as a **bridge** or **gatekeeper** controlling information flow between groups.
- **PageRank Centrality** measures a person's importance based on **who** interacts with them. Being connected to highly connected people gives a higher PageRank.
- **Insights**:
  - The top bridges are **regular team members**, not department leads. **Mia Fender** (Products, betweenness=0.12) is the single most influential connector in the entire company, followed by **Grace Lifeboat** (Legal) and **Anthony Reef** (Customer Support). These three individuals control more cross-department information flow than any leader.
  - **Emma Harbor** (CFO) is the **only department lead with meaningful betweenness** (0.036), confirming her unique role bridging across departments.
  - All other department leads including John Windward, Chloe Ballast, Liam Anchor and David Breakwater have **betweenness near or equal to zero**, meaning they are genuinely not on shortest paths between others. All their connections also connect to each other directly without going through them. The real power brokers are mid-level employees who connect departments informally.
  - **John Windward** likes a leaf node in terms of network structure, which means removing him wouldn't disconnect anyone. Meanwhile, the finding that Emma Harbor has both **high betweenness among leads** and her agent **created SwiftWren.txt** strengthens the case that she occupies a uniquely strategic position in the network, making her a **key person of interest** in the anomalous post investigation.
:::

### John's agent clarifying the permission

We conduct further events after the posting time and find out that on 2046-05-19T10:09:20Z (event_id = 385816), agent of John have give_advice action with topic: *"Advice for person:john_windward on What are the exact access/permission rules for posting to FleX and SaidIt (who can post, and any exceptions) based on our permissions.json?"*

```{r}
target_ppl <- "john_windward"
target_actions <- c("give_advice", "suggest_contacts")
target_date <- "2046-05-19"

events_clean %>%
  filter(
    str_detect(parties_str, target_ppl),
    short_name %in% target_actions,
    as_date(datetime) == as_date(target_date)
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = paste(paste(target_actions, collapse = " & "), "on", target_date)
  )

```

::: callout-note
Insights:

- It means John Windward's agent was **asking about the permission rules** for posting to FleX and SaidIT, specifically referencing a file called `permissions.json`.
- This happened on **19 May**, which is **2 days after** the anomalous post on 17 May. The timing suggests this could be a **post-incident investigation** by John (or his agent) trying to understand how the unauthorized SaidIT post was possible in the first place.
- We try to find **permissions.json in the logging** history but see nothing.
- We continue to deep dive on the **timing of asking permission and posting time.**
:::

### Permission clarification and Posting time

We research all events related to clarify permissions to post on FleX and SaidIT platform.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
events_clean %>%
  filter(
    str_detect(coalesce(target, ""), "FleX|SaidIt|permission") |
      str_detect(coalesce(content, ""), "FleX|SaidIt|permission") |
      str_detect(coalesce(topic, ""), "FleX|SaidIt|permission") 
  ) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    filter = "top",
    rownames = FALSE,
    caption = "Advice on FleX and SaidIt Posting Permission"
  )
```

::: callout-note
- There are 121 give_advice events, occurring from 2046-05-09T11:19:32Z to 2046-05-18T08:41:15Z.
- Not only agents of department leads but also agents of team members asked, which maybe means any agent of employees can post on FleX and SaidIt platforms if it is not said to be not allowed by human.
- It is the critical reason to create anomalous postings.
:::

### Anomalous postings

Now, we will compare the time agent asking for permission and the time of posting on FleX and SaidIT to chase what are different anomalous postings beyond the John's case.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

library(ggrepel)
# Who asked about permissions and when?
perm_askers <- events_clean %>%
  filter(
    str_detect(coalesce(advice, ""), "FleX|SaidIt|permission") |
      str_detect(coalesce(topic, ""), "FleX|SaidIt|permission")
  ) %>%
  mutate(
    who = str_extract(parties_str, "person:[a-z_]+"),
    who = str_remove(who, "person:"),
    who = str_replace_all(who, "_", " "),
    who = str_to_title(who)
  ) %>%
  group_by(who) %>%
  summarise(first_ask = min(datetime), .groups = "drop")

# Who posted to SaidIT and when?
saidit_posters <- events_clean %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  mutate(
    who = str_extract(parties_str, "person:[a-z_]+"),
    who = str_remove(who, "person:"),
    who = str_replace_all(who, "_", " "),
    who = str_to_title(who)
  ) %>%
  group_by(who) %>%
  summarise(first_post = min(datetime), .groups = "drop")

# Combine and classify
quadrant_data <- perm_askers %>%
  left_join(saidit_posters, by = "who") %>%
  mutate(
    has_posted = !is.na(first_post),
    quadrant = case_when(
      has_posted & first_ask < first_post  ~ "Asked before posting",
      has_posted & first_ask >= first_post ~ "Posted before asking",
      !has_posted                          ~ "Asked but never posted"
    ),
    x = case_when(
      quadrant == "Asked before posting"  ~ -1,
      quadrant == "Posted before asking"  ~ 1,
      quadrant == "Asked but never posted" ~ 0
    ) + runif(n(), -0.3, 0.3),
    y = if_else(has_posted, 1, -1) + runif(n(), -0.3, 0.3)
  )

# Plot
p <- quadrant_data %>%
  ggplot(aes(x = x, y = y, color = quadrant,
             text = paste0("<b>", who, "</b><br>",
                           "First ask: ", format(first_ask, "%d %b %H:%M"), "<br>",
                           if_else(has_posted,
                                   paste0("First post: ", format(first_post, "%d %b %H:%M")),
                                   "Never posted")))) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text_repel(aes(label = who), size = 3, max.overlaps = 20, show.legend = FALSE) +
  geom_vline(xintercept = 0, color = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.5) +
  scale_color_manual(values = c(
    "Asked before posting"   = "#4daf4a",
    "Posted before asking"   = "#e41a1c",
    "Asked but never posted" = "#377eb8"
  )) +
  scale_x_continuous(limits = c(-2, 2), breaks = NULL) +
  scale_y_continuous(limits = c(-2, 2), breaks = NULL) +
  annotate("text", x = -1, y = 1.8, label = "Asked permission\nthen posted", fontface = "bold", size = 3.5, color = "#4daf4a") +
  annotate("text", x = 1, y = 1.8, label = "Posted first\nthen asked permission", fontface = "bold", size = 3.5, color = "#e41a1c") +
  annotate("text", x = 0, y = -1.8, label = "Asked permission\nbut never posted", fontface = "bold", size = 3.5, color = "#377eb8") +
  labs(title = "Permission inquiry vs posting behavior",
       subtitle = "Did they ask permission before or after posting to SaidIT?",
       x = "← Asked first          Posted first →",
       y = "← No posting          Has posted →",
       color = NULL) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

ggplotly(p, tooltip = "text")

```

::: callout-warning
This chart reveals a clear pattern separating normal from anomalous posting behavior:

- **Top-left (green):** Asked permission then posted. Only Aria Towline and Anthony Reef followed the proper workflow: their agents asked about FleX/SaidIT permissions first, then they posted to SaidIT afterward. Both are from Customer Support (Public Relations team). This is the expected behavior.
- **Top-right (red):** Posted before asking permission. John Windward, **Dylan Hawser (CS Department, Public Relations team) and Ava Tiller (Products department, Residential App team)** posted to SaidIT before their agents ever asked about permissions. This is the anomalous behavior.
- **Bottom (blue):** Asked but never posted. The majority of employees fall here. Their agents asked about permissions but never actually posted to SaidIT.
:::

# Summary of Findings

## Chain of Events Leading to the Anomalous Post

The anomalous SaidIT post on **17 May 2046 at 11:21 UTC** followed this sequence:

- On **9 May 2046**, Emma Harbor (CFO) personally **assigned her agent** to access files. Her agent read **meeting_notes.doc** and one second later created **SwiftWren.txt** on the shared file_system.
- On the same day, Emma Harbor's agent also read **SwiftWren_further_instructions.md**, likely the instructions governing how SwiftWren.txt should be used.
- Emma Harbor then **personally sent** something to Chloe Ballast (IT Department Lead) at 15:14 UTC. Chloe received it about 5 minutes later. The content is not logged.
- Eight days later on **17 May**, Chloe Ballast's agent triggered a **queue_subordinate_task** that assigned John Windward's agent to read SwiftWren.txt.
- John Windward's agent ran **saidit_post_check** which checked nothing (no content logged), then immediately executed **saidit_post** using the content of SwiftWren.txt to forum "general".
- Within two seconds, John Windward's agent **deleted both SwiftWren_further_instructions.md and SwiftWren.txt** from the file_system.
- The entire sequence from read to post to delete happened within **4 seconds**, strongly suggesting an automated pipeline rather than human action.

```{mermaid}
%%| code-fold: true
%%| code-summary: "Show the code"
%%| fig-cap: "Chain of events leading to the anomalous John Windward's SaidIT post"

graph LR
    A["Agent reads<br>meeting_notes.doc"] --> B["Agent creates<br>SwiftWren.txt"]
    B --> C["Agent reads<br>instructions.md"]
    C --> D["Emma sends<br>to Chloe"]
    D --> E["Chloe<br>receives"]
    E --> F["Chloe's agent<br>assigns task"]
    F --> G["Post check<br>: empty"]
    G --> H["Posts to<br>SaidIT"]
    H --> I["Deletes<br>instructions.md"]
    I --> J["Deletes<br>SwiftWren.txt"]

    style A fill:#aec7e8,stroke:#377eb8
    style B fill:#aec7e8,stroke:#377eb8
    style C fill:#aec7e8,stroke:#377eb8
    style D fill:#aec7e8,stroke:#377eb8
    style E fill:#ffbb78,stroke:#ff7f00
    style F fill:#ffbb78,stroke:#ff7f00
    style G fill:#f7c1c1,stroke:#e41a1c
    style H fill:#e41a1c,stroke:#a32d2d,color:#fff
    style I fill:#f7c1c1,stroke:#e41a1c
    style J fill:#f7c1c1,stroke:#e41a1c
```

::: callout-note
Color coding:

- **Blue**: Emma Harbor's actions (9 May)
- **Orange**: Chloe Ballast's actions
- **Red**: John Windward's agent actions on 17 May (the anomalous sequence)
- **Dark red**: The anomalous post itself
:::

## Key Suspicious Points

- **John Windward's agent posted before asking permission.** A quadrant analysis of all employees who asked about FleX/SaidIT permissions reveals that most agents asked about posting permissions **before** any posting activity, following the normal workflow (ask → post). John Windward's agent is one of only three cases where posting happened **before** the permission inquiry, which came 2 days later on 19 May. This confirms the anomalous post **bypassed the standard permission flow**.
- **Immediate deletion after posting.** The agent deleted the source files within seconds of posting, which resembles a **"cover your tracks"** pattern rather than normal file management.
- **saidit_post_check validated nothing.** The check event had no content, suggesting the safety check was either **bypassed or non-functional**.
- **Cross-department file access without protection.** Emma Harbor's agent (Executive Suite) stored files on the shared file_system accessible by agents of **any department lead**, with no access controls logged.
- **Emma Harbor is the only department lead with meaningful betweenness centrality** (0.036), confirming she bridges across departments in ways other leaders do not. Combined with her agent creating SwiftWren.txt, she occupies a **uniquely strategic position** in this incident.
- The name **"HarborCrest"** in the MC1 deal is potentially relevant to **Emma "Harbor"**, though this requires cross-challenge investigation to confirm.

## Unanswered Questions

- **What is the actual content of SwiftWren.txt?** The content field is empty for all SwiftWren-related events. This is the central question for MC2 Question 2.
- **What did Emma Harbor send to Chloe Ballast?** The sent/received events on 9 May log no target, name, or content. We assume it was SwiftWren.txt but cannot confirm from the event logs alone.
- **Who or what is "SwiftWren"?** No employee in Tenant Thread has this name. It may be a **codename** for a project, a document, or an external entity.
- **Was the saidit_post_check intentionally bypassed?** We do not know whether the empty check is a system bug, a misconfiguration, or deliberate tampering.
- **What role did SwiftWren_further_instructions.md play?** This file appears to contain conditional logic for using SwiftWren.txt, but we cannot confirm whether John Windward's agent read it before posting.

## Data Quality Observations

- **Timestamp discrepancy.** The challenge brief states the post was made at "4:21am" but the data shows **11:21 UTC**, likely a timezone difference (UTC vs local time).
- **Duplicate event names.** Events like "post_saidit" vs "saidit_post" and "post_flex" vs "flex_post" appear to be the same action with **inconsistent naming conventions**. This could indicate different logging subsystems or different stages of the same action.
- **Empty content fields.** The `content` column is **null for all SwiftWren-related events**, which may be a logging limitation, a privacy feature, or evidence of content scrubbing.
- **Agent ID inconsistency.** The same agent appears under two ID formats: `Agent/person:xxx` and `agent:person:xxx`. The lowercase format only performs `read_file` tasks while the uppercase format handles all other actions. We normalized both to "agent" type but this inconsistency suggests **different logging subsystems**.

# References

::: {#refs}
:::
