---
title: "MC2 — Data Preprocessing"
author: "Nguyen Kim Hau"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

This report focuses on [Mini-Challenge 2](https://vast-challenge.github.io/2026/MC2.html) of [VAST Challenge 2026](https://vast-challenge.github.io/2026/index.html).

**Context**: You work in the AI systems administration department of Company A. A post was recently made on SaidIT looks like a malfunction of an agent: It is in gibberish and in a forum that appears random. Your job is to track down why this post occurred and suggest a remedy to the situation by building a visual analytics system that integrates visuals with algorithms to enable interactive exploration and documentation of the multi-agent system!

This part is data preprocessing stage, which contains following steps:

-   Loaded efficiently and inspected.
-   Flattened from its nested JSON structure into tidy tables.
-   Cleaned (timestamps converted, duplicates checked, missingness understood).
-   Joined with the organisation chart so that every actor in an event can be traced back to their **department**, **team**, and **human principal** (for autonomous agents).

The end-product of this stage is a single enriched events table (`events_mapped`) plus several lookup tables. From those, the very last chunk of this notebook produces the **John Windward mapping view**: what John and the agents tied to him actually did in the system.

# Setup

## Loading R packages

The packages below cover the four jobs we'll do in this notebook: read JSON (`jsonlite`), wrangle in the tidyverse, work with timestamps (`lubridate`), and walk the org chart as a graph (`tidygraph`). `DT` and `gt` are for rendering interactive / formatted tables in the report; `skimr` gives one-line data summaries.

```{r}
#| code-fold: false
pacman::p_load(jsonlite, tidyverse, lubridate,
               tidygraph, ggraph,
               DT, gt, skimr)
```

## Reading the raw data

The JSON files are large and slow to re-parse on every render. The convention is to parse JSON **once** with `fromJSON()`, then cache as `.rds` (R's native binary format: smaller, faster, and it preserves column types).

The `if (!file.exists(...))` guard means the conversion runs only the first time.

```{r}
#| eval: false
#| code-fold: false
#| 
if (!file.exists("data/mc2_data.rds")) {
  mc2_data <- fromJSON("data/mc2_data.json")
  write_rds(mc2_data, "data/mc2_data.rds")
}
if (!file.exists("data/org_chart.rds")) {
  org_chart <- fromJSON("data/org_chart.json")
  write_rds(org_chart, "data/org_chart.rds")
}
```

```{r}
#| code-fold: false
mc2_data  <- read_rds("data/mc2_data.rds")
org_chart <- read_rds("data/org_chart.rds")

```

# Data Wrangling

## Inspecting raw structure

### MC2 events

`mc2_data` is a list with a `description` field and an `events` data frame. The events table is where everything happens, each row is one action taken by a person, an agent, or a system.

```{r}
#| code-fold: false
glimpse(mc2_data)
```

```{r}
#| code-fold: false
mc2_events <- mc2_data$events
glimpse(mc2_events)
```

::: callout-note
A few things to notice from the glimpse above:

-   `parties` is a **list-column**, each row is a character vector of one or more actor IDs (people like `person:john_windward`, agents like `agent:xxx`, or systems).
-   `details` is a **nested data frame**, its columns differ by event type, so most cells are `NA`. This is normal for an event-log schema.
-   `when` is a Unix timestamp (numeric), we'll convert it to a proper datetime.
-   `short_name` tells us the *kind* of action, this is the field we'll group on.
:::

### Organisation chart

`org_chart` is in NetworkX node-link format: a `nodes` list and an `edges` list.

Each person has an `id`, `label`, `type` (`person` / `team` / `department` / `company`), and sometimes a `title`.

Edges carry a `relation` (`contains` or `led_by`) describing how parts of the org nest.

```{r}
#| code-fold: false
glimpse(org_chart)
```

#### Extracting the nodes and edges (links) data from org_chart

```{r}
#| code-fold: false
org_nodes <- as_tibble(org_chart$nodes)
org_edges <- as_tibble(org_chart$edges)

org_nodes %>% count(type)
```

```{r}
#| code-fold: false
org_edges %>% count(relation)
```

#### Draw organization tree

```{r}
#| code-fold: true
#| code-summary: "Show the code"
graph <- tbl_graph(nodes = org_nodes, 
                   edges = org_edges,
                   directed = TRUE)

ggraph(graph, layout = "tree") +
  geom_edge_diagonal(alpha = 0.4) +
  geom_node_point(aes(color = type), size = 3) +
  geom_node_text(aes(label = label), size = 2, repel = TRUE) +
  theme_void() +
  labs(title = "Tenant Thread Organization Chart")
```

## Cleaning the events table

### Convert timestamps and add row id

Working with Unix timestamps is painful for filtering by date. We convert `when` into a `POSIXct` datetime in UTC, then derive a `date` and `hour` column for easier grouping. We also keep an explicit `event_id` (the original `id` field) and a sequential `row` index for ordering.

```{r}
#| code-fold: false
events_clean <- mc2_events %>% 
  mutate(
    event_id  = id,
    datetime  = as_datetime(when, tz = "UTC"),
    date      = as_date(datetime),
    hour      = hour(datetime),
    .keep = "unused"  # drops `id` and `when`, keeps the rest
  ) %>% 
  mutate(row = row_number(), .before = 1)

events_clean %>% 
  select(row, event_id, datetime, short_name) %>% 
  head(5)
```

### Sanity-check the time range

The challenge brief says the anomalous post occurs in **17** **May 2046**. A quick min/max confirms the data matches that.

```{r}
#| code-fold: false
events_clean %>% 
  summarise(
    n_events   = n(),
    first_time = min(datetime),
    last_time  = max(datetime),
    n_days     = n_distinct(date)
  )
```

So dataset ranges among 70 days, from 8 May to 16 July, covering the checking point (17 May).

### Distribution of event types

`short_name` is the verb of each event: what was done. Counting them tells us which actions dominate the log and which are rare. Rare verbs are often the most diagnostic for an anomaly investigation.

```{r}
event_type_counts <- events_clean %>% 
  count(short_name, sort = TRUE)

event_type_counts
```

There are some events seem be duplicated. Eg: post_saidit and saidit_post maybe one event (due to the inconsistency of name convention) or two different stages. I will confirm in further investigation.

### Missingness in `details`

`details` is a nested data frame: most of its columns are populated only for the event types that need them. The summary below tells us which detail-fields actually carry information.

```{r}
events_clean$details %>% 
  summarise(across(where(is.character), ~ sum(!is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "detail_field", values_to = "non_na") %>% 
  arrange(desc(non_na)) %>%
  print(n = Inf)
```

### Duplicates

There should be one row per `event_id`. We verify.

```{r}
events_clean %>% 
  summarise(
    n_rows    = n(),
    n_unique  = n_distinct(event_id),
    n_dup_ids = n() - n_distinct(event_id)
  )
```

Yes, each row represents for one event.

## Transforming: long-format actor table

The hard part of working with this data is that **`parties` is a list-column**. To answer "what did John Windward do?", we need one row per (event × actor), not one row per event. We `unnest()` to long form.

```{r}
events_long <- events_clean %>% 
  select(row, event_id, datetime, date, hour, short_name, parties) %>% 
  unnest_longer(parties, values_to = "actor_id")

events_long %>% head(8)
```

Now every event appears as many times as it has actors. We can immediately ask "how many distinct actors appear in the log?"

```{r}
events_long %>% 
  summarise(
    n_event_actor_rows = n(),
    n_distinct_actors  = n_distinct(actor_id)
  )
```

### Classify actor IDs

Actor IDs follow a `prefix:name` convention. The prefix tells us whether the actor is a person, an autonomous agent, a system, or something else. We extract the prefix to make filtering and joins easier downstream.

```{r}
events_long <- events_long %>% 
  mutate(
    actor_kind = str_extract(actor_id, "^[^:]+"),   # part before the colon
    actor_name = str_remove(actor_id, "^[^:]+:")    # part after
  )

events_long %>% count(actor_kind, sort = TRUE)
```

::: callout-warning
The raw `actor_kind` extraction returns 54 categories, which is too many.

Inspection reveals that the data uses **two inconsistent ID formats for the same people**: most actors carry a prefix (`person:hannah_starboard`), but **49 bare person names** also appear (`Hannah Starboard`).

The number 49 exactly matches the number of persons in `org_chart`, so every employee shows up under both formats in different events.

It raises a question: **is there a pattern to which format gets used?** For example:

-   Does `person:` appear when the action is logged by the system (so the system knows the canonical ID)?
-   Does the bare name appear when the action originates from a free-text field (like an email body, a message, or a calendar invite where humans typed the name)?

For now, just normalise them and move on then we will investiage in later stage.

=\> We normalise the bare names to their `person:` equivalents using the org chart as the source of truth, and merge the two agent variants (`Agent/person` and `agent`) into a single `agent` kind. After normalisation there are four actor kinds: `person`, `agent`, `system`, `world`.
:::

```{r}
# 1. Build a lookup: bare name -> person ID, from the org chart
name_to_id <- org_nodes %>% 
  filter(type == "person") %>% 
  select(label, id) %>% 
  deframe()

# 2. Normalise actor_id, then derive actor_kind
events_long <- events_long %>% 
  mutate(
    # bare names get rewritten to person:xxx, prefixed IDs stay as-is
    actor_id = if_else(
      str_detect(actor_id, ":"),
      actor_id,
      name_to_id[actor_id]
    ),
    # extract the prefix, then merge the two agent variants
    actor_kind = str_extract(actor_id, "^[^:/]+"),
    actor_kind = if_else(actor_kind == "Agent", "agent", actor_kind),
    actor_name = str_remove(actor_id, "^[^:]+:")
  )

events_long %>% count(actor_kind, sort = TRUE)
```

## Building lookup tables from `org_chart`

### Person lookup with department & team

The org chart is a tree: `company → department → team → person`, except for department leads (like John Windward) who hang directly under a department rather than under a team. We walk down from each department to build a flat lookup that, for every person, gives us their team and department.

```{r}
# Build a directed graph so we can traverse it cleanly
org_graph <- tbl_graph(nodes = org_nodes, edges = org_edges, directed = TRUE)

# Recursive descent: for each department, find every person reachable below it
descendants_of <- function(start_id, edges_df) {
  out <- character(0)
  frontier <- start_id
  while (length(frontier) > 0) {
    children <- edges_df %>% 
      filter(source %in% frontier) %>% 
      pull(target)
    out <- c(out, children)
    frontier <- children
  }
  unique(out)
}

dept_ids <- org_nodes %>% filter(type == "department") %>% pull(id)

person_lookup <- map_dfr(dept_ids, function(d) {
  desc <- descendants_of(d, org_edges)
  tibble(department_id = d, descendant_id = desc)
}) %>% 
  # keep only persons among the descendants
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id"))

# Also find the team (if any) for each person
team_ids <- org_nodes %>% filter(type == "team") %>% pull(id)

person_team <- map_dfr(team_ids, function(t) {
  desc <- descendants_of(t, org_edges)
  tibble(team_id = t, descendant_id = desc)
}) %>% 
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id")) %>% 
  select(person_id = descendant_id, team_id)

person_lookup <- person_lookup %>% 
  transmute(
    person_id     = descendant_id,
    person_name   = label,
    title         = title,
    department_id = department_id
  ) %>% 
  left_join(person_team, by = "person_id") %>% 
  left_join(org_nodes %>% filter(type == "department") %>% 
              select(department_id = id, department_name = label),
            by = "department_id") %>% 
  left_join(org_nodes %>% filter(type == "team") %>% 
              select(team_id = id, team_name = label),
            by = "team_id")

person_lookup
```

A quick check: is John Windward in there with the right department?

```{r}
#| code-fold: false
person_lookup %>% filter(person_id == "person:john_windward") %>%
  t()
```

Customer Support, Department Lead matches the brief.

### Agent → principal lookup (derived from events)

Autonomous agents are not in the org chart, they only appear inside `mc2_events`. But the agent ID itself encodes the principal: `Agent/person:john_windward` literally means "the agent of John Windward". So the lookup is one line of string manipulation, no heuristic needed. Inspection also reveals two ID formats (`Agent/...` and `agent:...`) for what appear to be the same logical agents. We keep both as distinct rows in the lookup but map them to the same `principal_id`.

A peek at all distinct agent IDs in the log:

```{r}
events_long %>% 
  filter(actor_kind == "agent") %>% 
  distinct(actor_id) %>%
  print(n = Inf)
```

We strip either prefix (`Agent/` or `agent:`) from each agent ID to expose the `person:<name>` part underneath, which is the principal. The same person can appear under two agent IDs (one per format), both correctly map to the same `principal_id`.

```{r}
agent_lookup <- events_long %>% 
  filter(actor_kind == "agent") %>% 
  distinct(agent_id = actor_id) %>% 
  mutate(
    # Strip either "Agent/" or "agent:" prefix to expose "person:<name>"
    principal_id = str_replace(agent_id, "^(Agent/|agent:)", "")
  )

agent_lookup
```

```{r}
#| code-fold: false
agent_lookup %>% filter(principal_id == "person:john_windward")
```

```{r}
#| code-fold: false
events_long %>% 
  filter(str_starts(actor_id, "agent:")) %>% 
  count(short_name, sort = TRUE)
```

**`read_file` events use a different ID format by convention** (whoever logs `read_file` events writes agent IDs as `agent:person:X`, while every other event type writes them as `Agent/person:X`). Same agent, different logging style for this one event type.

## Master enriched table

Now we join everything: each (event × actor) row gets enriched with department, team, title for persons and for agents, their inferred principal plus that principal's department / team.

```{r}
# Person enrichment, join when actor is a person
events_mapped <- events_long %>% 
  left_join(person_lookup,
            by = c("actor_id" = "person_id")) %>% 
  rename(actor_person_name = person_name,
         actor_title       = title,
         actor_department  = department_name,
         actor_team        = team_name) %>% 
  select(-department_id, -team_id)

# Agent enrichment, look up the principal then re-join person info
events_mapped <- events_mapped %>% 
  left_join(agent_lookup %>% select(agent_id, principal_id),
            by = c("actor_id" = "agent_id")) %>% 
  left_join(person_lookup %>% 
              select(principal_id   = person_id,
                     principal_name = person_name,
                     principal_title = title,
                     principal_department = department_name,
                     principal_team = team_name),
            by = "principal_id")

# Tidy, unified "effective person" view, who is responsible for the action
events_mapped <- events_mapped %>% 
  mutate(
    effective_person_id   = coalesce(
      if_else(actor_kind == "person", actor_id, NA_character_),
      principal_id
    ),
    effective_person_name = coalesce(actor_person_name, principal_name),
    effective_department  = coalesce(actor_department, principal_department),
    effective_team        = coalesce(actor_team, principal_team),
    effective_title       = coalesce(actor_title, principal_title)
  )

events_mapped %>% 
  select(row, event_id, datetime, short_name, actor_kind, actor_id,
         effective_person_name, effective_department, effective_team) %>% 
  head(10)
```

After running, sanity check that John's events show up under both his own and agent rows:

```{r}
#| code-fold: false
events_mapped %>% 
  filter(effective_person_id == "person:john_windward") %>% 
  count(actor_kind)
```

Save the enriched table for the analysis stage:

```{r}
#| code-fold: false
#| eval: false
write_rds(events_mapped, "data/events_mapped.rds")
write_rds(person_lookup, "data/person_lookup.rds")
write_rds(agent_lookup,  "data/agent_lookup.rds")
```

## John Windward mapping view

This is the table the project requirement specifically asks for: **where does John Windward sit in the org**, and **what did he and his agents do**?

### John Windward in the org chart

```{r}
#| code-fold: false
person_lookup %>% 
  filter(person_id == "person:john_windward") %>% 
  gt() %>% 
  tab_header(title = "John Windward-organisational placement")
```

## Agents linked to John Windward

```{r}
#| code-fold: false
jw_agents <- agent_lookup %>% 
  filter(principal_id == "person:john_windward")

jw_agents %>% 
  gt()
```

## Every event involving John Windward or his agents

This is the core mapping table - one row per (event × actor) where the actor is either John himself or an agent attributed to him.

```{r}
jw_actor_ids <- c("person:john_windward", jw_agents$agent_id)

jw_events_view <- events_mapped %>% 
  filter(actor_id %in% jw_actor_ids) %>% 
  arrange(datetime) %>% 
  select(
    datetime, short_name,
    actor_kind, actor_id,
    effective_person_name, effective_department, effective_team,
    event_id
  )

jw_events_view %>% 
  datatable(
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE,
    caption = "John Windward & his agents — all logged actions, in chronological order"
  )
```

## What kinds of actions did John & his agents take?

A compact summary so we can see, at a glance, the action-mix that ended in the anomalous post.

```{r}
jw_events_view %>% 
  count(actor_kind, short_name, sort = TRUE) %>% 
  pivot_wider(names_from = actor_kind, values_from = n, values_fill = 0) %>% 
  gt() %>% 
  tab_header(title = "Action counts-John Windward and his agents")
```

## Lead-up to the anomalous SaidIt post

The post was sent on **2046-05-17 at 04:21 UTC**. We zoom into the 24 hours immediately preceding it.

```{r}
#| code-fold: false
post_time <- ymd_hms("2046-05-17 04:21:00", tz = "UTC")

jw_events_view %>% 
  filter(datetime >= post_time - hours(24),
         datetime <= post_time + minutes(5)) %>% 
  datatable(
    options = list(pageLength = 20, scrollX = TRUE),
    rownames = FALSE,
    caption = "Events involving John Windward or his agents in the 24h window around the SaidIt post"
  )
```

## Outputs of this preprocessing stage

By the end of this notebook we have, ready for the analysis & visualisation stage:

-   `events_clean`: one row per event, with a proper `datetime` and a stable `row` index.
-   `events_long`: one row per (event × actor), with `actor_kind` classified.
-   `person_lookup`: every person in Tenant Thread mapped to their team and department.
-   `agent_lookup`: every autonomous agent mapped to its human principal, parsed from the agent ID.
-   `events_mapped`: the master enriched table; this is what subsequent visualisations should build on.
-   `jw_events_view`: the focused John-Windward mapping that scaffolds the MC2 narrative.

```{mermaid}
%%| fig-cap: "Data flow: from raw JSON files to the enriched events table"

flowchart TD
    %% Raw sources
    A[(mc2_data.json)]:::raw
    B[(org_chart.json)]:::raw

    %% Top level objects
    A -->|fromJSON| C[mc2_data<br/><i>list</i>]
    B -->|fromJSON| D[org_chart<br/><i>list</i>]

    C -->|$events| E[mc2_events<br/><i>nested data frame</i>]
    D -->|$nodes, $edges| F[org_nodes + org_edges]

    %% events_clean
    E -->|mutate datetime, event_id, row| G[events_clean<br/>1 row per event]:::out

    %% events_long
    G -->|select + unnest_longer parties<br/>normalise bare names<br/>extract actor_kind| H[events_long<br/>1 row per event × actor]:::out

    %% person_lookup
    F -->|walk tree:<br/>department → team → person| I[person_lookup<br/>1 row per person]:::out

    %% agent_lookup
    H -->|filter actor_kind == agent<br/>parse principal from ID| J[agent_lookup<br/>1 row per agent_id]:::out

    %% events_mapped
    H -->|left_join| K[events_mapped<br/>enriched]:::final
    I -->|left_join on actor_id| K
    J -->|left_join on actor_id| K
    I -->|left_join on principal_id| K

    %% jw_events_view
    K -->|filter John + his agents| L[jw_events_view<br/>focused subset]:::final

    classDef raw fill:#f5f5f5,stroke:#999,stroke-width:1px,color:#333
    classDef out fill:#dceefb,stroke:#3b82f6,stroke-width:1px,color:#0c4a6e
    classDef final fill:#fef3c7,stroke:#f59e0b,stroke-width:2px,color:#78350f
```

::: callout-note
-   **Grey nodes** are raw source files.
-   **Blue nodes** are intermediate processed tables.
-   **Yellow nodes** are the final outputs, will be input for further analysis.
-   Arrows are labelled with the **operation** that produced the next tae.
:::
