---
title: "MC2 — Data Preprocessing"
author: "Nguyen Kim Hau"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

This document covers the **data preparation stage** for VAST Challenge 2026 Mini-Challenge 2 (MC2). The task is to reconstruct the chain of events that led to the anomalous SaidIt post sent by **John Windward** on **May 17, 2046 at 4:21am**.

Before any visualisation can be done, the raw data needs to be:

1.  Loaded efficiently and inspected,
2.  Flattened from its nested JSON structure into tidy tables,
3.  Cleaned (timestamps converted, duplicates checked, missingness understood),
4.  Joined with the organisation chart so that every actor in an event can be traced back to their **department**, **team**, and **human principal** (for autonomous agents).

The end-product of this stage is a single enriched events table (`events_mapped`) plus several lookup tables. From those, the very last chunk of this notebook produces the **John Windward mapping view**: what John and the agents tied to him actually did in the system.

# Setup

## Loading R packages

The packages below cover the four jobs we'll do in this notebook: read JSON (`jsonlite`), wrangle in the tidyverse, work with timestamps (`lubridate`), and walk the org chart as a graph (`tidygraph`). `DT` and `gt` are for rendering interactive / formatted tables in the report; `skimr` gives one-line data summaries.

```{r}
pacman::p_load(jsonlite, tidyverse, lubridate,
               tidygraph, ggraph,
               DT, gt, skimr)
```

## Reading the raw data

The JSON files are large and slow to re-parse on every render. The convention is to parse JSON **once** with `fromJSON()`, then cache as `.rds` (R's native binary format: smaller, faster, and it preserves column types). The `if (!file.exists(...))` guard means the conversion runs only the first time.

```{r}
#| eval: false
if (!file.exists("data/mc2_data.rds")) {
  mc2_data <- fromJSON("data/mc2_data.json")
  write_rds(mc2_data, "data/mc2_data.rds")
}
if (!file.exists("data/org_chart.rds")) {
  org_chart <- fromJSON("data/org_chart.json")
  write_rds(org_chart, "data/org_chart.rds")
}
```

```{r}
mc2_data  <- read_rds("data/mc2_data.rds")
org_chart <- read_rds("data/org_chart.rds")

```

# Inspecting raw structure

## MC2 events

`mc2_data` is a list with a `description` field and an `events` data frame. The events table is where everything happens, each row is one action taken by a person, an agent, or a system.

```{r}
glimpse(mc2_data)
```

```{r}
mc2_events <- mc2_data$events
glimpse(mc2_events)
```

A few things to notice from the glimpse above:

-   `parties` is a **list-column**, each row is a character vector of one or more actor IDs (people like `person:john_windward`, agents like `agent:xxx`, or systems).
-   `details` is a **nested data frame**, its columns differ by event type, so most cells are `NA`. This is normal for an event-log schema.
-   `when` is a Unix timestamp (numeric), we'll convert it to a proper datetime.
-   `short_name` tells us the *kind* of action, this is the field we'll group on.

## Organisation chart

`org_chart` is in NetworkX node-link format: a `nodes` list and an `edges` list. Each person has an `id`, `label`, `type` (`person` / `team` / `department` / `company`), and sometimes a `title`. Edges carry a `relation` (`contains` or `led_by`) describing how parts of the org nest.

```{r}
glimpse(org_chart)
```

```{r}
org_nodes <- as_tibble(org_chart$nodes)
org_edges <- as_tibble(org_chart$edges)

org_nodes %>% count(type)
```

```{r}
org_edges %>% count(relation)
```

Draw tree

```{r}
#| code-fold: false
#| code-summary: "Show the code"
graph <- tbl_graph(nodes = org_nodes, 
                   edges = org_edges,
                   directed = TRUE)

ggraph(graph, layout = "tree") +
  geom_edge_diagonal(alpha = 0.4) +
  geom_node_point(aes(color = type), size = 3) +
  geom_node_text(aes(label = label), size = 2, repel = TRUE) +
  theme_void() +
  labs(title = "Tenant Thread Organization Chart")
```

# Cleaning the events table

## Convert timestamps and add row id

Working with Unix timestamps is painful for filtering by date. We convert `when` into a `POSIXct` datetime in UTC, then derive a `date` and `hour` column for easier grouping. We also keep an explicit `event_id` (the original `id` field) and a sequential `row` index for ordering.

```{r}
events_clean <- mc2_events %>% 
  mutate(
    event_id  = id,
    datetime  = as_datetime(when, tz = "UTC"),
    date      = as_date(datetime),
    hour      = hour(datetime),
    .keep = "unused"  # drops `id` and `when`, keeps the rest
  ) %>% 
  mutate(row = row_number(), .before = 1)

events_clean %>% 
  select(row, event_id, datetime, short_name) %>% 
  head(5)
```

## Sanity-check the time range

The challenge brief says the events occur in **May 2046**. A quick min/max confirms the data matches that.

```{r}
events_clean %>% 
  summarise(
    n_events   = n(),
    first_time = min(datetime),
    last_time  = max(datetime),
    n_days     = n_distinct(date)
  )
```

## Distribution of event types

`short_name` is the verb of each event: what was done. Counting them tells us which actions dominate the log and which are rare. Rare verbs are often the most diagnostic for an anomaly investigation.

```{r}
event_type_counts <- events_clean %>% 
  count(short_name, sort = TRUE)

event_type_counts
```

## Missingness in `details`

`details` is a nested data frame: most of its columns are populated only for the event types that need them. The summary below tells us which detail-fields actually carry information.

```{r}
events_clean$details %>% 
  summarise(across(where(is.character), ~ sum(!is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "detail_field", values_to = "non_na") %>% 
  arrange(desc(non_na))
```

## Duplicates

There should be one row per `event_id`. We verify.

```{r}
events_clean %>% 
  summarise(
    n_rows    = n(),
    n_unique  = n_distinct(event_id),
    n_dup_ids = n() - n_distinct(event_id)
  )
```

# Transforming: long-format actor table

The hard part of working with this data is that **`parties` is a list-column**. To answer "what did John Windward do?", we need one row per (event × actor), not one row per event. We `unnest()` to long form.

```{r}
events_long <- events_clean %>% 
  select(row, event_id, datetime, date, hour, short_name, parties) %>% 
  unnest_longer(parties, values_to = "actor_id")

events_long %>% head(8)
```

Now every event appears as many times as it has actors. We can immediately ask "how many distinct actors appear in the log?"

```{r}
events_long %>% 
  summarise(
    n_event_actor_rows = n(),
    n_distinct_actors  = n_distinct(actor_id)
  )
```

## Classify actor IDs

Actor IDs follow a `prefix:name` convention. The prefix tells us whether the actor is a person, an autonomous agent, a system, or something else. We extract the prefix to make filtering and joins easier downstream.

```{r}
events_long <- events_long %>% 
  mutate(
    actor_kind = str_extract(actor_id, "^[^:]+"),   # part before the colon
    actor_name = str_remove(actor_id, "^[^:]+:")    # part after
  )

events_long %>% count(actor_kind, sort = TRUE)
```

# Building lookup tables from `org_chart`

## Person lookup with department & team

The org chart is a tree: `company → department → team → person`, except for department leads (like John Windward) who hang directly under a department rather than under a team. We walk down from each department to build a flat lookup that, for every person, gives us their team and department.

```{r}
# Build a directed graph so we can traverse it cleanly
org_graph <- tbl_graph(nodes = org_nodes, edges = org_edges, directed = TRUE)

# Recursive descent: for each department, find every person reachable below it
descendants_of <- function(start_id, edges_df) {
  out <- character(0)
  frontier <- start_id
  while (length(frontier) > 0) {
    children <- edges_df %>% 
      filter(source %in% frontier) %>% 
      pull(target)
    out <- c(out, children)
    frontier <- children
  }
  unique(out)
}

dept_ids <- org_nodes %>% filter(type == "department") %>% pull(id)

person_lookup <- map_dfr(dept_ids, function(d) {
  desc <- descendants_of(d, org_edges)
  tibble(department_id = d, descendant_id = desc)
}) %>% 
  # keep only persons among the descendants
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id"))

# Also find the team (if any) for each person
team_ids <- org_nodes %>% filter(type == "team") %>% pull(id)

person_team <- map_dfr(team_ids, function(t) {
  desc <- descendants_of(t, org_edges)
  tibble(team_id = t, descendant_id = desc)
}) %>% 
  inner_join(org_nodes %>% filter(type == "person"),
             by = c("descendant_id" = "id")) %>% 
  select(person_id = descendant_id, team_id)

person_lookup <- person_lookup %>% 
  transmute(
    person_id     = descendant_id,
    person_name   = label,
    title         = title,
    department_id = department_id
  ) %>% 
  left_join(person_team, by = "person_id") %>% 
  left_join(org_nodes %>% filter(type == "department") %>% 
              select(department_id = id, department_name = label),
            by = "department_id") %>% 
  left_join(org_nodes %>% filter(type == "team") %>% 
              select(team_id = id, team_name = label),
            by = "team_id")

person_lookup
```

A quick check: is John Windward in there with the right department?

```{r}
person_lookup %>% filter(person_id == "person:john_windward")
```

Customer Support, Department Lead matches the brief.

## Agent → principal lookup (derived from events)

Autonomous agents (`agent:*` IDs) are **not** in the org chart. They only show up inside `mc2_events`. To attach an agent to a human, we look at events whose `short_name` and `details` suggest assignment / ownership (anything containing words like *agent*, *assign*, *create*, *config*, *own*). For each such event we record the (agent, person) pair that appears together in `parties`.

```{r}
# All event types — pick out ones plausibly about agent ownership
events_clean %>% 
  count(short_name, sort = TRUE) %>% 
  filter(str_detect(short_name, regex("agent|assign|create|config|own|spawn|deploy",
                                      ignore_case = TRUE)))
```

```{r}
# Wide pairs of (agent, person) co-appearing in the same event's parties
agent_principal_pairs <- events_long %>% 
  filter(actor_kind %in% c("agent", "person")) %>% 
  select(row, actor_kind, actor_id) %>% 
  pivot_wider(names_from = actor_kind, values_from = actor_id,
              values_fn = list) %>% 
  unnest_longer(agent, values_to = "agent_id") %>% 
  unnest_longer(person, values_to = "person_id") %>% 
  filter(!is.na(agent_id), !is.na(person_id)) %>% 
  count(agent_id, person_id, name = "co_occurrences") %>% 
  arrange(desc(co_occurrences))

agent_principal_pairs %>% head(10)
```

For each agent, the **most frequent** human co-actor is taken as the likely principal. This is a heuristic, it should be revisited if specific assignment events exist in the data (look at the table above; if `short_name` values like `agent_created` or `agent_assigned` exist, prefer those as the source of truth).

```{r}
agent_lookup <- agent_principal_pairs %>% 
  group_by(agent_id) %>% 
  slice_max(co_occurrences, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  rename(principal_id = person_id, evidence_strength = co_occurrences)

agent_lookup %>% head(10)
```

# Master enriched table

Now we join everything: each (event × actor) row gets enriched with department, team, title for persons and for agents, their inferred principal plus that principal's department / team.

```{r}
# Person enrichment
events_mapped <- events_long %>% 
  # join person info directly when actor is a person
  left_join(person_lookup,
            by = c("actor_id" = "person_id")) %>% 
  rename(actor_person_name = person_name,
         actor_title       = title,
         actor_department  = department_name,
         actor_team        = team_name) %>% 
  select(-department_id, -team_id)

# Agent enrichment: look up the principal then re-join person info
events_mapped <- events_mapped %>% 
  left_join(agent_lookup %>% select(agent_id, principal_id),
            by = c("actor_id" = "agent_id")) %>% 
  left_join(person_lookup %>% 
              select(principal_id   = person_id,
                     principal_name = person_name,
                     principal_title = title,
                     principal_department = department_name,
                     principal_team = team_name),
            by = "principal_id")

# Tidy: a unified "effective person" view — who is morally responsible for the action
events_mapped <- events_mapped %>% 
  mutate(
    effective_person_id   = coalesce(actor_id %>% 
                                       if_else(actor_kind == "person", ., NA_character_),
                                     principal_id),
    effective_person_name = coalesce(actor_person_name, principal_name),
    effective_department  = coalesce(actor_department, principal_department),
    effective_team        = coalesce(actor_team, principal_team),
    effective_title       = coalesce(actor_title, principal_title)
  )

events_mapped %>% 
  select(row, event_id, datetime, short_name, actor_kind, actor_id,
         effective_person_name, effective_department, effective_team) %>% 
  head(10)
```

Save the enriched table for the analysis stage:

```{r}
#| eval: false
write_rds(events_mapped, "data/events_mapped.rds")
write_rds(person_lookup, "data/person_lookup.rds")
write_rds(agent_lookup,  "data/agent_lookup.rds")
```

# John Windward mapping view

This is the table the project requirement specifically asks for: **where does John Windward sit in the org**, and **what did he and his agents do**?

## John Windward in the org chart

```{r}
person_lookup %>% 
  filter(person_id == "person:john_windward") %>% 
  gt() %>% 
  tab_header(title = "John Windward-organisational placement")
```

## Agents linked to John Windward

```{r}
jw_agents <- agent_lookup %>% 
  filter(principal_id == "person:john_windward")

jw_agents %>% 
  gt() %>% 
  tab_header(title = "Agents whose most-frequent human co-actor is John Windward")
```

## Every event involving John Windward or his agents

This is the core mapping table - one row per (event × actor) where the actor is either John himself or an agent attributed to him.

```{r}
jw_actor_ids <- c("person:john_windward", jw_agents$agent_id)

jw_events_view <- events_mapped %>% 
  filter(actor_id %in% jw_actor_ids) %>% 
  arrange(datetime) %>% 
  select(
    datetime, short_name,
    actor_kind, actor_id,
    effective_person_name, effective_department, effective_team,
    event_id
  )

jw_events_view %>% 
  datatable(
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE,
    caption = "John Windward & his agents — all logged actions, in chronological order"
  )
```

## What kinds of actions did John & his agents take?

A compact summary so we can see, at a glance, the action-mix that ended in the anomalous post.

```{r}
jw_events_view %>% 
  count(actor_kind, short_name, sort = TRUE) %>% 
  pivot_wider(names_from = actor_kind, values_from = n, values_fill = 0) %>% 
  gt() %>% 
  tab_header(title = "Action counts-John Windward and his agents")
```

## Lead-up to the anomalous SaidIt post

The post was sent on **2046-05-17 at 04:21 UTC**. We zoom into the 24 hours immediately preceding it.

```{r}
post_time <- ymd_hms("2046-05-17 04:21:00", tz = "UTC")

jw_events_view %>% 
  filter(datetime >= post_time - hours(24),
         datetime <= post_time + minutes(5)) %>% 
  datatable(
    options = list(pageLength = 20, scrollX = TRUE),
    rownames = FALSE,
    caption = "Events involving John Windward or his agents in the 24h window around the SaidIt post"
  )
```

# Outputs of this preprocessing stage

By the end of this notebook we have, ready for the analysis & visualisation stage:

-   `events_clean`: one row per event, with a proper `datetime` and a stable `row` index.
-   `events_long`: one row per (event × actor), with `actor_kind` classified.
-   `person_lookup`: every person in Tenant Thread mapped to their team and department.
-   `agent_lookup`: every autonomous agent mapped to its most-likely human principal.
-   `events_mapped`: the master enriched table; this is what subsequent visualisations should build on.
-   `jw_events_view`: the focused John-Windward mapping that scaffolds the MC2 narrative.
