---
title: "Take-home_Ex02"
author: "Nguyen Kim Hau"
date: "May 26, 2026"
date-modified: "last-modified"
---

```{r}
pacman::p_load(jsonlite, tidyverse, 
               DT)
```

`mc1_data <- fromJSON("data/mc1_data.json")`

`write_rds(mc1_data, "data/mc1_data.rds")`

```{r}
mc1_data <- read_rds("data/mc1_data.rds") 
glimpse(mc1_data)
```

```{r}
names(mc1_data$rounds)
```

```{r}
glimpse(mc1_data$rounds$environment_context)
```

```{r}
mc1_data$rounds$communications[[1]] %>% 
  head(2)
```

```{r}
mc1_data$rounds$communications[[1]] %>%
  head(2) %>%
  unnest(internal_state)
```

```{r}
mc1_data$rounds$participants[[1]] %>% 
  head(2)
```

Unnest

unnest communications

internal_state is unnested df, recipients is list

```{r}
comms_df <- mc1_data$rounds %>%
  select(hour, communications) %>%
  unnest(communications) %>%
  unnest(internal_state)
glimpse(comms_df)

```

```{r}
# Environment context data  
env_df <- mc1_data$rounds %>%
  select(hour, environment_context) %>%
  unnest(environment_context)
glimpse(env_df)
```

Check each hour have how many critical_deadlines

```{r}
env_df %>%
  mutate(n = map_int(critical_deadlines, length)) %>%
  select(hour, n)
```

```{r}
# Participants data
participants_df <- mc1_data$rounds %>%
  select(hour, participants) %>%
  unnest(participants)
glimpse(participants_df)
```
