---
title: "Unmasking Motor Insurance Risk Drivers"
subtitle: "Executive Summary"
author: "Nguyen Kim Hau"
date: "May 24, 2026"
date-modified: "last-modified"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    show-slide-number: print
    transition: slide
    width: 1280
    height: 720
    code-fold: true
    code-summary: "Show the code"
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, readxl, kableExtra, patchwork, 
               ggrepel, scales, RColorBrewer)

motoinsu <- read_delim("data/Motor_insurance_portfolio.csv", delim = ";")

motoinsu <- motoinsu %>%
  mutate(
    insured_id = as.character(insured_id),
    across(c(year, policy_type, policy_status, business_type,
             payment_frequency, bonus_score, fuel_type, seats,
             vehicle_brand, municipality_type, circulation_area), 
           as.factor),
    across(c(liability_premium, property_damage_premium,
             theft_premium, fire_premium, glass_premium,
             legal_protection_premium, occupants_premium,
             total_exposure, liability_exposure), 
           as.double)
  )

motoinsu <- motoinsu %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value) %>%
  mutate(
    fire_incurred = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(occupants_incurred, 0)
  )

top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")
motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

theme_exec <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1B2A4A"),
    plot.subtitle = element_text(size = 12, color = "#5D6D7E"),
    axis.text = element_text(color = "#5D6D7E"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```
## Contents

::::: columns
::: {.column width="50%"}
**Portfolio Profile**

1.  Dataset and Portfolio Overview
2.  Categorical Variable Distributions
3.  Numerical Variable Distributions

**Risk and Profitability**

4.  Temporal Performance Trend
5.  Loss Ratio by Policy Type
6.  Loss Ratio by Bonus Score
:::

::: {.column width="50%"}
**Deep Dive Analysis**

7.  Optional Coverage Pricing Mismatch
8.  Vehicle Brand Risk Analysis
9.  Driver Age and Bonus Score Patterns
10. Correlation Structure

**Conclusion**

11. Key Findings and Recommendations
:::
:::::
## Insight 1: Dataset and Portfolio Overview

::::: columns
::: {.column width="50%"}
```{r}
#| label: overview-policy
#| fig-width: 5.5
#| fig-height: 3.8

comp_data <- tibble(
  Category = c("CC", "TPG", "COMP_E", "COMP_N", "TP"),
  Percentage = c(58.0, 20.7, 12.2, 7.5, 1.6)
)

ggplot(comp_data, aes(x = reorder(Category, Percentage), y = Percentage, fill = Category)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0(Percentage, "%")), hjust = -0.15, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("CC" = "#3D5A99", "TPG" = "#5B8DB8", 
                                "COMP_E" = "#E8913A", "COMP_N" = "#C0392B", "TP" = "#27AE60")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Policy Type Distribution", x = NULL, y = "Percentage (%)") +
  theme_exec + theme(legend.position = "none")
```
:::

::: {.column width="50%"}
```{r}
#| label: overview-business
#| fig-width: 5.5
#| fig-height: 3.8

biz_data <- tibble(
  Type = c("New Business", "Portfolio Renewal"),
  Pct = c(69.3, 30.7)
)

ggplot(biz_data, aes(x = reorder(Type, Pct), y = Pct, fill = Type)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(Pct, "%")), hjust = -0.15, size = 4.5, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("New Business" = "#C0392B", "Portfolio Renewal" = "#3D5A99")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Business Type Composition", x = NULL, y = "Percentage (%)") +
  theme_exec + theme(legend.position = "none")
```
:::