---
title: "Unmasking Motor Insurance Risk Drivers"
subtitle: "Executive Summary"
author: "Nguyen Kim Hau"
date: "May 24, 2026"
date-modified: "last-modified"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    show-slide-number: all
    transition: slide
    width: 1280
    height: 720
    code-fold: true
    code-summary: "Show the code"
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, readxl, tidyr, kableExtra, psych, dplyr, 
               patchwork, corrplot, RColorBrewer, gganimate, ggrepel)

motoinsu <- read_delim("data/Motor_insurance_portfolio.csv", delim = ";")

motoinsu <- motoinsu %>%
  mutate(
    insured_id = as.character(insured_id),
    across(c(year, policy_type, policy_status, business_type,
             payment_frequency, bonus_score, fuel_type, seats,
             vehicle_brand, municipality_type, circulation_area), 
           as.factor),
    across(c(liability_premium, property_damage_premium,
             theft_premium, fire_premium, glass_premium,
             legal_protection_premium, occupants_premium,
             total_exposure, liability_exposure), 
           as.double)
  )

motoinsu <- motoinsu %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value) %>%
  mutate(
    fire_incurred = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(occupants_incurred, 0)
  )

top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")
motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

theme_exec <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1B2A4A"),
    plot.subtitle = element_text(size = 12, color = "#5D6D7E"),
    axis.text = element_text(color = "#5D6D7E"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```

## Contents

::::: columns
::: {.column width="50%"}
**Portfolio Profile**

1.  Dataset and Portfolio Overview
2.  Categorical Variable Distributions
3.  Numerical Variable Distributions

**Risk and Profitability**

4.  Temporal Performance Trend
5.  Loss Ratio by Policy Type
6.  Loss Ratio by Bonus Score
:::

::: {.column width="50%"}
**Deep Dive Analysis**

7.  Optional Coverage Pricing Mismatch
8.  Vehicle Brand Risk Analysis
9.  Driver Age and Bonus Score Patterns
10. Correlation Structure

**Conclusion**

11. Key Findings and Recommendations
:::
:::::

## Insight 1: Dataset and Portfolio Overview

::::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 5.5
#| fig-height: 3.8

ggplot(motoinsu, aes(x = fct_rev(fct_infreq(policy_type)))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", hjust = -0.2, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Distribution of policy Type", x = "policy_type", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
  
```
:::

::: {.column width="50%"}
```{r}
#| label: overview-business
#| fig-width: 5.5
#| fig-height: 3.8

ggplot(motoinsu, aes(x = fct_rev(fct_infreq(business_type)))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue", width = 0.5) +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", hjust = -0.2, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Distribution of Business Type", x = "businesss_type", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```
:::
:::::

::: {style="font-size: 0.5em;"}
-   The portfolio contains **354,140 records** from a Spanish motor insurance company, across **47 variables** (11 categorical, 35 numerical) from **Jan 2022 to Dec 2024** .
-   **Policy type**: CC dominates (58%), meaning most customers choose combined third-party + additional coverage. Pure TP is rare (1.6%) - customers prefer more protection.
-   **Business type**: 69.3% new business (NB) vs 30.7% portfolio (P), which means the insurer is heavily acquiring new customers rather than renewing existing ones. This is a risk signal, new customers have unknown claims history.
:::

## Insight 2: Loss Ratio by Policy Type

```{r}
#| fig-width: 18
#| fig-height: 6

# Chart 1: Loss Ratio
group_var <- "policy_type"

loss_data <- motoinsu %>%
  group_by(!!sym(group_var)) %>%
  summarise(loss_ratio = sum(total_incurred)/sum(total_premium)*100) %>%
  mutate(status = ifelse(loss_ratio >= 70, "Above 70%", "Below 70%"))

p1 <- ggplot(loss_data, aes(x = reorder(!!sym(group_var), loss_ratio), 
                       y = loss_ratio, fill = status)) +
  geom_col() +
  geom_hline(yintercept = 70, linetype = "dashed", color = "grey30", linewidth = 0.75) +
  annotate("text", x = 0.5, y = 72, label = "70% Benchmark", 
           hjust = 0, size = 3, color = "grey30") +
  geom_text(aes(label = paste0(round(loss_ratio, 1), "%")), hjust = -0.2, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("Above 70%" = "#F1948A", "Below 70%" = "#AED6F1")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = paste("Loss Ratio by", group_var), x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Chart 2: Total Claims Boxplot
x_var <- "policy_type"
y_var <- "total_claims"

median_labels1 <- motoinsu %>%
  filter(!!sym(y_var) > 0) %>%         
  group_by(!!sym(x_var)) %>%
  summarise(med = median(!!sym(y_var)))

p2 <- motoinsu %>%
  filter(!!sym(y_var) > 0) %>%
  ggplot(aes(y = !!sym(y_var), x = !!sym(x_var), fill = !!sym(x_var))) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_label(data = median_labels1,
             aes(x = !!sym(x_var), y = med,
                 label = scales::comma(round(med, 0))),
             fill = "white", size = 3, inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  labs(title = paste("Boxplot of", y_var, "by", x_var),
       x = NULL, y = paste(y_var, "(log scale)")) +
  theme_minimal() +
  theme(legend.position = "none")

# Chart 3: Total Incurred Boxplot
y_var2 <- "total_incurred"

median_labels2 <- motoinsu %>%
  filter(!!sym(y_var2) > 0) %>%         
  group_by(!!sym(x_var)) %>%
  summarise(med = median(!!sym(y_var2)))

p3 <- motoinsu %>%
  filter(!!sym(y_var2) > 0) %>%
  ggplot(aes(y = !!sym(y_var2), x = !!sym(x_var), fill = !!sym(x_var))) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_label(data = median_labels2,
             aes(x = !!sym(x_var), y = med,
                 label = scales::comma(round(med, 0))),
             fill = "white", size = 3, inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  labs(title = paste("Boxplot of", y_var2, "by", x_var),
       x = NULL, y = paste(y_var2, "(log scale)")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.title.y = element_blank())

p1 + p2 + p3 + plot_layout(ncol = 3)
```

::: {style="font-size: 0.5em;"}
-   **COMP_N**: **Most dangerous product**. Highest median incurred (1,392) and widest claim frequency (IQR 1 to 5). No deductible encourages excessive claiming, driving up both frequency and severity. **Immediate repricing needed**.
-   **COMP_E**: Median incurred (954) is lower than TP despite broader coverage, proving the deductible effectively filters small claims. However, more outliers reach 10+ claims, suggesting it attracts **higher risk customers**.
-   **TP**: **Most profitable** by loss ratio, but surprisingly second highest median incurred (1,114). Low frequency (median 1) but **costly when claims occur** due to bodily injury liability.
-   **CC**: Lowest median incurred (882) with median 1 claim. High volume product with mixed, smaller claims. **Stable and well priced**.
-   **TPG**: Same low median incurred (882) and flat claim frequency as CC. Glass coverage adds **minimal risk** to the base product.
:::

## Insight 3: Optional Coverage Pricing Mismatch

:::::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 3

premium_vars <- c("liability_premium", "property_damage_premium",
                   "theft_premium", "fire_premium", "glass_premium",
                   "legal_protection_premium", "occupants_premium")

premium_long <- motoinsu %>%
  select(all_of(premium_vars)) %>%
  pivot_longer(cols = everything(), 
               names_to = "premium_type", 
               values_to = "premium_value") %>%
  filter(premium_value > 0)

median_labels <- premium_long %>%
  group_by(premium_type) %>%
  summarise(median_val = median(premium_value))

ggplot(premium_long, aes(x = reorder(premium_type, premium_value, FUN = median), 
                          y = premium_value, fill = premium_type)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.1) +
  geom_label(data = median_labels,
             aes(x = premium_type, y = median_val,
                 label = round(median_val, 1)),
             hjust = -0.2, size = 3, fontface = "bold",
             fill = "white", alpha = 0.8, label.size = 0,
             inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Pastel1") +
  coord_flip() +
  labs(title = "Distribution of Premium Types",
       subtitle = "log scale, zero values excluded") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::

::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 3

incurred_vars <- c("liability_incurred", "liability_property_incurred",
                    "liability_injury_incurred", "property_incurred", "theft_incurred",
                    "fire_incurred", "glass_incurred", "legal_protection_incurred",
                    "occupants_incurred")

incurred_long <- motoinsu %>%
  select(all_of(incurred_vars)) %>%
  pivot_longer(cols = everything(), 
               names_to = "incurred_type", 
               values_to = "incurred_value") %>%
  filter(incurred_value > 0)

median_labels <- incurred_long %>%
  group_by(incurred_type) %>%
  summarise(median_val = median(incurred_value))

ggplot(incurred_long, aes(x = reorder(incurred_type, incurred_value, FUN = median), 
                           y = incurred_value, fill = incurred_type)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.1) +
  geom_label(data = median_labels,
             aes(x = incurred_type, y = median_val,
                 label = round(median_val, 1)),
             hjust = -0.2, size = 3, fontface = "bold",
             fill = "white", alpha = 0.8, label.size = 0,
             inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Pastel2") +
  coord_flip() +
  labs(title = "Distribution of Incurred Loss Types",
       subtitle = "log scale, zero values excluded",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )
```
:::

::: {style="font-size: 0.4em;"}
**Safe (well priced)**:

-   **Property Damage**: Most balanced coverage. Both violins are wide and well spread, showing effective risk segmentation in pricing that matches actual claim variation.

-   **Liability**: Mature pricing with a broad, full bodied premium violin showing good differentiation. Incurred violin is also wide with a long right tail, reflecting inherently high bodily injury severity.

-   **Legal Protection**: Only optional coverage adequately priced. Both violins are compact and narrow, meaning low variation on both sides and proportional shapes.

**Risky (underpriced)**:

-   **Fire**: Most critical. Premium violin is flat and concentrated near zero, while incurred violin is the widest in the entire chart, showing highly unpredictable expensive claims against nearly no pricing.

-   **Occupants**: Premium violin is the narrowest (no risk differentiation at all), yet incurred violin is broad and boxy, suggesting wide variation in claim severity.

-   **Glass**: Incurred violin has a distinctive diamond/star shape tightly clustered around a fixed cost, suggesting standardized repairs. Premium violin is thin and low, still too small to cover them.

-   **Theft**: Premium violin is wide and spread showing some differentiation, but incurred violin is a compact rectangle in a predictable band. The shapes do not match, pricing variation is misaligned with actual claim patterns.

**Overall**: Mandatory coverages have wide, full bodied violins on both sides, confirming mature pricing. Optional coverages have flat/narrow premium violins against much larger and varied incurred violins, confirming systematic underpricing.
:::
::::::

## Insight 4: Vehicle Brand Risk

::::: columns
::: {.column width="50%" style="font-size: 0.5em;"}
```{r}
#| fig-width: 6
#| fig-height: 3
#| 
motoinsu %>%
  group_by(vehicle_brand_grouped) %>%
  summarise(
    loss_ratio = sum(total_incurred)/sum(total_premium)*100,
    total_premium = sum(total_premium)
  ) %>%
  filter(vehicle_brand_grouped != "Others") %>%
  ggplot(aes(x = total_premium, y = loss_ratio,    # x = total_premium now
             size = total_premium, color = vehicle_brand_grouped)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = vehicle_brand_grouped),
            vjust = -1.5, size = 3, color = "black") +
  geom_text(aes(label = paste0("€", scales::comma(round(total_premium/1e6, 1)), "M")),
            vjust = 2.5, size = 2.8, color = "grey30") +
  scale_x_continuous(labels = scales::comma) +    
  scale_size_continuous(range = c(3, 18), guide = "none") +
  labs(
    title = "Loss Ratio vs Vehicle Brand (sized by Total Premium)",
    x = "Total Premium (€)", y = "Loss Ratio (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 8)
  )
```

-   **OPEL** (77%) and **SEAT** (74%) have the **highest loss ratios** with the **smallest premium volumes**, making them the **most concerning** brands.

-   **VOLKSWAGEN** (71%) carries the **largest premium volume** among named brands, requiring **careful management** given its revenue weight.

-   **CITROEN** (69%) offers the **best balance** of volume and profitability among French brands.

-   **PEUGEOT** (66%) and **RENAULT** (67%) deliver the **lowest loss ratios** with **strong premium volumes**, making them the **most profitable** brands in the portfolio.
:::

::: {.column width="50%" style="font-size: 0.5em;"}
```{r}
#| fig-width: 6
#| fig-height: 3
# Bin
top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")

motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

# Draw chart
ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand_grouped))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_blank(),
        axis.title = element_blank()
        )
```

-   The near-equal spread among top brands means risk is **not concentrated** in any single brand’s characteristics.
-   Among top 7, French brands (RENAULT, CITROEN, PEUGEOT) collectively represent \~27%, reflecting **Spain’s strong preference for French** automobiles.
-   SEAT’s presence (\~7.9%) is notable as a **Spanish domestic brand**, expected in a Spanish portfolio.
:::
:::::

## Insight 5: Temporal Analysis by Year

::::: columns
::: {.column width="50%" style="font-size: 0.5em;"}
```{r}
#| fig-width: 6
#| fig-height: 4
#| 
motoinsu %>%
  group_by(year) %>%
  summarise(
    avg_incurred = mean(total_incurred, na.rm = TRUE),
    avg_premium  = mean(total_premium, na.rm = TRUE),
    loss_ratio   = sum(total_incurred)/sum(total_premium)*100
  ) %>%
  pivot_longer(-year, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = as.numeric(as.character(year)), y = value, 
             color = metric, group = metric)) +
  geom_line() + 
  geom_point() +
  facet_wrap(~metric, scales = "free_y", nrow = 3) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  theme_minimal() +
  labs(title = "Portfolio Performance Trend 2022-2024", x = NULL, y = NULL)
```

-   **Average incurred** rose steadily from \~187 to \~224, signaling **growing claims severity** year over year.

-   **Average premium** peaked in 2023 then declined, suggesting **premium cuts to attract new business** while claims costs kept rising.

-   **Loss ratio** jumped from \~66.5% to \~75%, with the **sharpest spike in 2024** (\~8 percentage points in one year).

=\> **Warning:** Rising claims, declining premiums, and worsening loss ratio indicate **accelerating financial stress**. Heavy new business acquisition without adequate premium adjustment is an **urgent risk** requiring immediate repricing action.
:::

::: {.column width="50%" style="font-size: 0.5em;"}
```{r}
#| fig-width: 6
#| fig-height: 4
#| 
anim_data <- motoinsu %>%
  filter(vehicle_brand_grouped != "Others") %>%
  group_by(year, vehicle_brand_grouped) %>%
  summarise(
    loss_ratio = sum(total_incurred) / sum(total_premium) * 100,
    .groups = "drop"
  ) %>%
  mutate(year_num = as.numeric(as.character(year)))

p <- ggplot(anim_data, aes(x = year_num, y = loss_ratio, 
                            color = vehicle_brand_grouped,
                            group = vehicle_brand_grouped)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = vehicle_brand_grouped),
            vjust = 1.5, size = 3, show.legend = FALSE, fontface = "bold") +
  geom_hline(yintercept = 70, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  labs(title = "Loss Ratio Trend by Vehicle Brand (2022–2024)",
       x = "Year", y = "Loss Ratio (%)", color = "Brand") +
  theme_minimal() +
  theme(axis.title.x = element_blank()) +
  transition_reveal(year_num) +
  ease_aes("linear")

animate(p, nframes = 60, fps = 10, width = 800, height = 500,
        renderer = gifski_renderer(),
        res = 100, units = "px")
```

-   Across three year window, **RENAULT and PEUGEOT** are the most stable, staying in the 63 to 68% range. They are the safest brands in the portfolio and should be prioritized for customer retention.
-   **OPEL** continues to show warning to be the highest volatile brand. This instability suggests a small portfolio where a few large claims can swing the ratio wildly. Unpredictable and hard to price.
-   **SEAT** is consistently the worst among mid range brands, hovering around 70 to 80% across all three years. It never dips below the benchmark, confirming it is structurally underpriced.
:::
:::::
