---
title: "Technical Report"
author: "Nguyen Kim Hau"
date: "May 15, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
bibliography: references.bib
---

# 1. Overview

# Data wrangling

## Importing dataset

```{r}
pacman::p_load(tidyverse, readxl, tidyr, kableExtra, psych, dplyr, 
               patchwork, corrplot, RColorBrewer, gganimate, ggrepel) 
```

The dataset contains anonymised microdata on automobile insurance policies and covers the main operations of the non-life portfolio from January 2022 to December 2024. The database shows 354,140 records and 47 variables, where each row corresponds to a policy transaction and each column to a specific attribute of that transaction.

This dataset uses semicolon delimiters, we use read_delim() to deal with semicolon delimiters.

```{r}
motoinsu <- read_delim(
  "data/Motor_insurance_portfolio.csv",
  delim = ";"
)
```

Out of a total of 47 attributes in the motoinsu tibble data frame, 29 of them are numerical data type (dbl - double) and 28 features are categorical data type (char - character).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
describe(motoinsu %>% 
           select(where(is.numeric)) # Only show numerical columns
         ) %>%
  select(vars, min, max, mean, skew) # Only get desired columns
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
var_desc <- read_excel("data/Descriptive_of_variables.xlsx") %>%
  fill(everything(), .direction = "down") %>%
  group_by(Variables, Group) %>%
  summarise(Description = paste(Description, collapse = " "), .groups = "drop")

var_desc %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
head(motoinsu, n=3)
```

## Conversing datatype

We realize that we need to change data type of some variables as follows.

```{r}
motoinsu <- motoinsu %>%
  mutate(
    # from dbl to chr (identifier)
    insured_id = as.character(insured_id),
    
    # from dbl to fac (categorical)
    across(c(year, policy_type, policy_status, business_type,
             payment_frequency, bonus_score, fuel_type, seats,
             vehicle_brand, municipality_type, circulation_area), 
           as.factor),
    
    # from char to dbl
    across(c(liability_premium, property_damage_premium,
             theft_premium, fire_premium, glass_premium,
             legal_protection_premium, occupants_premium,
             total_exposure, liability_exposure), 
           as.double)
  )
```

::: callout-tip
-   dbl stands for: double, numbers with decimals.
-   fct: factor, text with fixed, known categories. Factor uses less memory, which matters in such large database.
-   char: character, plain text without predefined categories.
:::

After changing data type, we have 1 identifier, 11 categorical variables and 35 numerical variables.

## Handling missing values

There are six features having missing values. Although all of these rates are small, under 1%, we treat them differently.

```{r}
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

```{r}
colMeans(is.na(motoinsu)) %>% # Caculate missing rate in each column
  round(5) %>%
  .[. > 0] # Filter ratio > 0
```

-   `vehicle_age`, `age_driving_licence` only lack two values, we drop these missing rows.
-   `fire_incurred`, `occupants_incurred` having missing values likely means no claim, in insurance context. Therefore we can change to zero.
-   `fuel_type`, `vehicle_value` have more missing values, which make I consider between using mode/medium or remove missing rows. Because they are seem important features, it is risky to impute data. This issue should be reported to the relevant stakeholders.

```{r}
motoinsu <- motoinsu %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value ) %>%
  mutate(
    fire_incurred    = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(occupants_incurred, 0)
  )
# Check after modification
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

## Checking outliers

To check whether we need to remove abnormal records, we use Z-score. It measures how many standard deviations a value is away from the mean.

`z = (value - mean) / standard deviation`

In general statistics, any data point with an absolute Z-score greater than 3 is widely flagged as an potential outlier, some researchs use threshold 3.5 (@1.3.5.17b). However in insurance field with heavy-tailed distributions, values far beyond 6 standard deviations occur with non-negligible probability (@embrechts1994).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Check all numeric variables at once
numeric_vars <- motoinsu %>% 
  select(where(is.numeric)) %>% 
  names()

outlier_summary <- map_dfr(numeric_vars, function(var) {
  x <- motoinsu[[var]]
  x_nonzero <- x[x > 0]  # for skewed insurance data, check non-zero only
  
  Q1  <- quantile(x, 0.25, na.rm = TRUE)
  Q3  <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  
  tibble(
    variable     = var,
    max_val      = max(x, na.rm = TRUE),
    p999         = quantile(x, 0.999, na.rm = TRUE),
    max_z        = if(length(x_nonzero) > 1) max(abs(scale(x_nonzero))) else NA,
    n_over_3sd   = if(length(x_nonzero) > 1) sum(abs(scale(x_nonzero)) > 3) else NA,
    pct_over_3sd = if(length(x_nonzero) > 1) round(mean(abs(scale(x_nonzero)) > 3) * 100, 3) else NA,
    iqr_fence    = Q3 + 3 * IQR_val,
    n_beyond_iqr = sum(x > Q3 + 3 * IQR_val, na.rm = TRUE)
  )
})

outlier_summary %>%
  kable(format.args = list(big.mark = ",", scientific = FALSE)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

So we are safe to move on further analysis.

# Visual Analysis

## Univariate Analysis

We demonstrate all distributions by each variable (categorical and numerical) to have a comprehensive analysis.

### Categorical variables

Among 11 categorical columns, vehicle_brand has too many distinct values. Therefore we will visualize its distribution and impute it later. Following is the distributions of 10 categorical features.

We mainly use histogram chart to demonstrate distrubutions in univariate analysis.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 15 # Adjust width of canvas
#| fig-height: 15 # Adjust height of canvas
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

cat_vars <- c("year", "policy_type", "policy_status", "business_type",
              "payment_frequency", "bonus_score", 
              "fuel_type", "seats",
              "municipality_type", "circulation_area")

plots <- lapply(cat_vars, function(col) {
  ggplot(motoinsu, aes(x = fct_infreq(!!sym(col)))) +
    geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
    geom_text(aes(y = after_stat(count/sum(count) * 100),
                  label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
              stat = "count", vjust = -0.5, size = 3) +
    labs(title = paste("Distribution of", col), x = col, y = "Percentage (%)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights: Categorical variables

#### Policy & Business Characteristics

-   **Year**: Data is unbalanced - 2024 (47.4%) has the most records, likely because policies accumulate over time. Dataset covers 3 years of a growing portfolio.
-   **Policy type**: CC dominates (58%), meaning most customers choose combined third-party + additional coverage. Pure TP is rare (1.6%) - customers prefer more protection.
-   **Policy status**: 89.2% active (A) vs 10.8% cancelled (C) - relatively low cancellation rate, portfolio is stable.
-   **Business type**: 69.3% new business (NB) vs 30.7% portfolio (P), which means the insurer is heavily acquiring new customers rather than renewing existing ones. This is a risk signal, new customers have unknown claims history.
-   **Payment frequency**: 81.3% pay annually (A) - most customers prefer single payment, simplifying cash flow for the insurer.
-   **Bonus score**: 96.5% have score G (good/favorable claims history) - the portfolio is dominated by low-risk customers historically.

#### Vehicle Characteristics

-   **Fuel type**: 64.2% diesel (D) vs 35.8% gasoline (G) - diesel vehicles dominate, which may have different risk profiles.
-   **Seats**: 84.9% have 5 seats - standard passenger cars dominate, very homogeneous portfolio.

#### Geographic

-   **Municipality type**: 64.6% type I (Inland) where typically have higher accident/theft risk than coastal and island areas.
-   **Circulation area**: Nearly split between U (53.8%) and R (46.2%) - urban vs rural almost balanced, suggesting national coverage.
:::

Vehicle brand distribution show high fragmentation with top brand \~10%, many brands with \<1%. We will combine all brands excluding RENAULT, CITROEN, PEUGEOT, VOLKSWAGEN, FORD, SEAT, and OPEL.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 15
#| fig-height: 6
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Then plot them by distribution.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 8
#| fig-height: 4

# Bin
top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")

motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

# Draw chart
ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand_grouped))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: callout-note
## Insights: Vehicle brand

-   The near-equal spread among top brands means risk is **not concentrated** in any single brand's characteristics.
-   Among top 7, French brands (RENAULT, CITROEN, PEUGEOT) collectively represent \~27%, reflecting **Spain's strong preference for French** automobiles.
-   SEAT's presence (\~7.9%) is notable as a **Spanish domestic brand**, expected in a Spanish portfolio.
-   The large "Others" group (40.6%) warrants further investigation - it contains luxury brands (BMW, MERCEDES, AUDI) and niche brands that **may carry higher vehicle values and different risk profiles**.
:::

### Numerical variables

Driver & Vehicle Distributions

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 16
#| fig-height: 8

num_vars <- c("driver_age", "vehicle_age", "age_driving_licence",
              "vehicle_value", "power_to_weight_ratio")

plots <- lapply(num_vars, function(col) {
  ggplot(motoinsu, aes(x = as.numeric(!!sym(col)))) + # as.numeric() inside aes() forces any discrete/factor columns to be treated as continuous
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = paste("Distribution of", col), x = col, y = "Count") +
    theme_minimal()
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights: Driver & Vehicle Distributions

#### Driver & Licence Characteristics

-   **driver_age**: roughly bell-shaped, peaking around **35–50 years**. Young drivers (\<25) and elderly (\>70) are rare, suggesting the portfolio skews toward middle-aged, experienced drivers.
-   **vehicle_age**: bimodal with peaks around **10 and 25 years,** meaning the portfolio covers both newer and older vehicles almost equally.
-   **age_driving_licence**: heavily right-skewed, most policyholders have held licences for **less than 20 years**, consistent with the driver age distribution.

#### Vehicle Characteristics

-   **vehicle_value**: heavily right-skewed, most vehicles are **low value under 50,000**, with a very long tail up to 300,000+ (luxury/high-end vehicles).
-   **power_to_weight_ratio**: concentrated tightly between 10 and 20, with very few high-performance outliers above 40. Portfolio is dominated by **standard passenger cars**.
:::

Premium, Claims & Exposure Distributions

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 15
#| fig-height: 15

num_vars <- c("total_premium", "liability_premium", "property_damage_premium",
              "theft_premium", "fire_premium", "glass_premium",
              "legal_protection_premium", "occupants_premium",
              "total_claims", "liability_claims", "property_claims",
              "total_incurred", "liability_incurred", "property_incurred",
              "total_exposure", "liability_exposure")

plots <- lapply(num_vars, function(col) {
  ggplot(motoinsu, aes(x = as.numeric(!!sym(col)))) + # as.numeric() inside aes() forces any discrete/factor columns to be treated as continuous
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = paste("Distribution of", col), x = col, y = "Count") +
    theme_minimal()
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights: Premium, Claims & Exposure Distributions

#### Premium Variables

-   **total_premium**: right-skewed, most policies have low-to-mid premiums with a long tail of high-premium policies
-   **liability_premium**: concentrated around **100–400**, the most normally distributed of all premiums — makes sense as liability coverage is **mandatory** for all vehicles in Spain
-   **property_damage_premium, theft_premium, fire_premium, glass_premium, legal_protection_premium, occupants_premium**: all **extremely right-skewed with mass at zero**, meaning most policies **do not include** these optional coverages. This is consistent with CC being 58% of policies — only combined policies carry these extras.

#### Claims Variables

-   **total_claims, liability_claims, property_claims**: **overwhelming majority are zero**. Rare claims events with counts of 1–2 for most claimants, very few exceed 3. Classic **low-frequency** insurance pattern.

#### Incurred Loss Variables

-   **total_incurred, liability_incurred, property_incurred**: mostly zero with **right tail** up to 600,000.

#### Exposure Variables

-   **total_exposure & liability_exposure**: U-shaped, **dominated by 1.0 (full year, annual payers)** and a smaller spike near 0.25 (quarterly). The spread between 0 and 1 represents mid-year cancellations, consistent with 10.8% cancellation rate.
:::

## Bivariate Analysis

We run some experiments to find out which pairs are best to visualize and deliver significant insights.

### Profitability Analysis: Incurred/Claims by Policy Type

We using boxplot chart to show relationship of `total_incurred` by `policy_type`. Because numerical variables here are right-skewed with a dozen of zero and with long tail with very large number, scale_y_log10() is used to spaces them logarithmically while maintains real values. We will calculate median filtering out zero values.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
x_var <- "policy_type"
y_var <- "total_incurred"

median_labels <- motoinsu %>%
  filter(!!sym(y_var) > 0) %>%         
  group_by(!!sym(x_var)) %>%
  summarise(med = median(!!sym(y_var)))

motoinsu %>%
  filter(!!sym(y_var) > 0) %>%
  ggplot(aes(y = !!sym(y_var), x = !!sym(x_var), fill = !!sym(x_var))) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_label(data = median_labels,
             aes(x = !!sym(x_var), y = med,
                 label = scales::comma(round(med, 0))),
             fill = "white", size = 3, inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  labs(title = paste("Boxplot of", y_var, "by", x_var),
       x = x_var, y = paste(y_var, "(log scale)")) +
  theme_minimal()

```

::: callout-note
## Insights: Total Incurred by Policy Type

-   **COMP_N** has the highest median incurred (1,392), consistent with no-excess comprehensive coverage encouraging more claims.
-   **TP** surprisingly ranks second (1,114), suggesting liability-only claims tend to be costly when they do occur (bodily injury).
-   **COMP_E** (954) is lower than TP despite broader coverage, likely because the excess/deductible filters out small claims.
-   **CC** and **TPG** share the lowest median (882), high volume products with more mixed, smaller claims.
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
x_var <- "policy_type"
y_var <- "total_claims"

median_labels <- motoinsu %>%
  filter(!!sym(y_var) > 0) %>%         
  group_by(!!sym(x_var)) %>%
  summarise(med = median(!!sym(y_var)))

motoinsu %>%
  filter(!!sym(y_var) > 0) %>%
  ggplot(aes(y = !!sym(y_var), x = !!sym(x_var), fill = !!sym(x_var))) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_label(data = median_labels,
             aes(x = !!sym(x_var), y = med,
                 label = scales::comma(round(med, 0))),
             fill = "white", size = 3, inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  labs(title = paste("Boxplot of", y_var, "by", x_var),
       x = x_var, y = paste(y_var, "(log scale)")) +
  theme_minimal()
```

::: callout-note
## Insights: Total Claims by Policy Type

-   **CC and TP** median is at **1**, most policies only ever have 1 claim at most, very low frequency.
-   **COMP_N shows the widest box** (highest IQR between 1–5), comprehensive without excess generates **more frequent claims**, likely because no excess/deductible discourages small claims
-   **COMP_E** has more outliers reaching 10+ claims, excess policies attract higher-risk customers who still manage to claim frequently despite the deductible barrier.
-   **TP and TPG** are nearly flat at 1, very low claim frequency for basic coverage.
:::

::: callout-warning
**COMP_N is the most volatile product**: high claim frequency AND high incurred losses. The absence of excess (deductible) in COMP_N likely encourages policyholders to file more claims, driving up both frequency and severity.

➨ The insurer should consider **reviewing COMP_N pricing strategy** as it appears to be the highest-risk product in the portfolio.
:::

### Temporal Analysis: performance by Year

With line chart, we perform average of incurred amount, average of premium and loss ratio (total_incurred/total_premium) by year to see the temporal changes.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
motoinsu %>%
  group_by(year) %>%
  summarise(
    avg_incurred = mean(total_incurred, na.rm = TRUE),
    avg_premium  = mean(total_premium, na.rm = TRUE),
    loss_ratio   = sum(total_incurred)/sum(total_premium)*100
  ) %>%
  pivot_longer(-year, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = as.numeric(as.character(year)), y = value, 
             color = metric, group = metric)) +
  geom_line() + 
  geom_point() +
  facet_wrap(~metric, scales = "free_y", nrow = 3) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  theme_minimal() +
  labs(title = "Portfolio Performance Trend 2022-2024", x = "Year", y = "Value")

```

::: callout-note
## Insights: performance by Year

-   **avg_incurred (red)** shows a consistent upward trend from \~187 in 2022 to \~224 in 2024, meaning the average cost per claim is increasing every year - a clear sign of **growing claims severity**.
-   **avg_premium (green)** peaked in 2023 then slightly declined in 2024, suggesting the insurer may have **reduced premiums to attract new business** (consistent with 69.3% new business ratio) while claims costs continued rising.
-   **loss_ratio (blue)** increases from \~66.5% in 2022 to \~67% in 2023 to \~75% in 2024. The jump from 2023 to 2024 is the most alarming signal — a near 8 percentage point spike in a single year. The portfolio is becoming **less profitable at an accelerating rate**.
:::

::: callout-warning
The combination of rising incurred losses, declining premiums, and worsening loss ratio suggests the portfolio is under increasing financial stress. The heavy reliance on **new business acquisition without sufficient premium adjustment** is a key risk driver the insurer should address urgently.
:::

### Loss Ratio Analysis: loss_ratio by policy_type/bonus_score

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 13
#| fig-height: 6

group_var <- "policy_type"

loss_data <- motoinsu %>%
  group_by(!!sym(group_var)) %>%
  summarise(loss_ratio = sum(total_incurred)/sum(total_premium)*100) %>%
  mutate(status = ifelse(loss_ratio >= 70, "Above 70%", "Below 70%"))

ggplot(loss_data, aes(x = reorder(!!sym(group_var), loss_ratio), 
                       y = loss_ratio, fill = status)) +
  geom_col() +
  geom_hline(yintercept = 70, linetype = "dashed", color = "grey30", linewidth = 0.8) +
  annotate("text", x = 0.5, y = 72, label = "70% Benchmark", 
           hjust = 0, size = 3.5, color = "grey30") +
  geom_text(aes(label = paste0(round(loss_ratio, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("Above 70%" = "#F1948A", "Below 70%" = "#AED6F1")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = paste("Loss Ratio by", group_var), x = NULL, y = "Loss Ratio (%)") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

::: callout-note
## **Loss Ratio by Policy Type**

-   COMP_N is the most unprofitable product at 93.5%, far exceeding the 70% benchmark by 23.5 percentage points. Immediate repricing is needed. The 19 point gap between COMP_N (93.5%) and COMP_E (73.9%) demonstrates that the excess/deductible is highly effective at reducing claims cost.
-   TP is the most profitable at 54.6%, sitting 15.4 points below the benchmark. Low claim frequency from limited coverage scope offsets its high individual claim severity.
-   TPG (69.6%) falls just below the 70% threshold, appearing healthy but barely so. It is surprisingly less profitable than CC (63.5%) despite simpler coverage, suggesting glass replacement costs are underpriced relative to their frequency.
-   Overall, two out of five products (COMP_N and COMP_E) exceed the 70% healthy benchmark, signaling underpricing in comprehensive coverage. The remaining three sit below the line but CC and TPG are not far from crossing it, consistent with the worsening loss ratio trend seen in the temporal analysis.
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 12
#| fig-height: 6

group_var <- "bonus_score"

bonus_colors <- c(
  "G" = "#F5B7B1",  # light red
  "B" = "#E74C3C",  # medium red
  "N" = "#922B21"   # dark red
)

motoinsu %>%
  group_by(!!sym(group_var)) %>%
  summarise(loss_ratio = sum(total_incurred)/sum(total_premium)*100) %>%
  ggplot(aes(x = reorder(!!sym(group_var), loss_ratio), 
             y = loss_ratio, fill = !!sym(group_var))) +
  geom_col() +
  geom_text(aes(label = paste0(round(loss_ratio, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_manual(values = bonus_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = paste("Loss Ratio by", group_var), x = NULL, y = "Loss Ratio (%)") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: callout-note
## **Loss Ratio by Bonus Score**

Results follow expected pattern - G (Good history) is most profitable at 69.7%, N (New) is most costly at 81.9%, with B (Bad) in between at 79.6%. The 12-point gap between G and N confirms claims history is a strong predictor of future loss.

With 69.3% of the portfolio being new business, the insurer is heavily exposed to the highest-loss segment, directly contributing to the worsening loss ratio trend seen in 2022 to 2024.
:::

### Density plot: Driver Age by Bonus Score

We use density plot to show if young/new drivers skew toward N/B scores.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
ggplot(motoinsu, 
       aes(x = driver_age, 
           fill = bonus_score)) +
  geom_density(alpha = 0.4) +
  theme_minimal()
```

::: callout-note
-   **N (New, blue): youngest drivers**.
    -   Peak sharply around age 33–35, very narrow distribution.

    -   New policyholders skew younger - first-time insurance buyers.
-   **B (Bad history, pink)**: **mirrors N almost exactly**.
    -   Nearly identical age distribution to N.

    -   Suggests bad-score customers are largely the same cohort as new customers - young drivers who recently accumulated claims.
-   **G (Good history, green)**: **older and wider spread**.
    -   Bimodal: one peak \~40, second bump around 55–65.

    -   Experienced drivers with long clean records dominate the older age bands.

    -   The long right tail to age 80+ is almost entirely G - elderly drivers have good bonus scores from decades of claim-free driving.
:::

### Violin Chart by Premium/Incurred Type

Because most optional coverages contain a large proportion of zero values (policies without that coverage), we exclude zeros and apply a log scale to better visualize the spread and shape of non zero premium distributions.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 12
#| fig-height: 6

premium_vars <- c("total_premium", "liability_premium", "property_damage_premium",
                   "theft_premium", "fire_premium", "glass_premium",
                   "legal_protection_premium", "occupants_premium")

premium_long <- motoinsu %>%
  select(all_of(premium_vars)) %>%
  pivot_longer(cols = everything(), 
               names_to = "premium_type", 
               values_to = "premium_value") %>%
  filter(premium_value > 0)

median_labels <- premium_long %>%
  group_by(premium_type) %>%
  summarise(median_val = median(premium_value))

ggplot(premium_long, aes(x = reorder(premium_type, premium_value, FUN = median), 
                          y = premium_value, fill = premium_type)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.1) +
  geom_label(data = median_labels,
             aes(x = premium_type, y = median_val,
                 label = round(median_val, 1)),
             hjust = -0.2, size = 3.5, fontface = "bold",
             fill = "white", alpha = 0.8, label.size = 0,
             inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Pastel1") +
  coord_flip() +
  labs(title = "Distribution of Premium Types",
       subtitle = "Violin + Boxplot (log scale, zero values excluded)",
       x = NULL, y = "Premium (log scale)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA)
  )
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 12
#| fig-height: 6

incurred_vars <- c("total_incurred", "liability_incurred", "liability_property_incurred",
                    "liability_injury_incurred", "property_incurred", "theft_incurred",
                    "fire_incurred", "glass_incurred", "legal_protection_incurred",
                    "occupants_incurred")

incurred_long <- motoinsu %>%
  select(all_of(incurred_vars)) %>%
  pivot_longer(cols = everything(), 
               names_to = "incurred_type", 
               values_to = "incurred_value") %>%
  filter(incurred_value > 0)

median_labels <- incurred_long %>%
  group_by(incurred_type) %>%
  summarise(median_val = median(incurred_value))

ggplot(incurred_long, aes(x = reorder(incurred_type, incurred_value, FUN = median), 
                           y = incurred_value, fill = incurred_type)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.1, outlier.alpha = 0.1) +
  geom_label(data = median_labels,
             aes(x = incurred_type, y = median_val,
                 label = round(median_val, 1)),
             hjust = -0.2, size = 3.5, fontface = "bold",
             fill = "white", alpha = 0.8, label.size = 0,
             inherit.aes = FALSE) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Pastel2") +
  coord_flip() +
  labs(title = "Distribution of Incurred Loss Types",
       subtitle = "Violin + Boxplot (log scale, zero values excluded)",
       x = NULL, y = "Incurred Loss (log scale)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA)
  )
```

::: callout-note
**Key abnormalities:**

-   **Fire:** Premium median = 1, incurred median = 1,522.9. A 1,500x gap, the most extreme mismatch. The premium violin is flat and concentrated while the incurred violin is the widest in the chart, showing highly unpredictable claim costs against nearly zero pricing.

-   **Occupants:** Premium median = 1.8, incurred median = 635.7. A 350x gap. Premium has the narrowest violin (zero risk differentiation), yet incurred shows a broad, boxy shape suggesting varied claim severity.

-   **Glass:** Incurred violin has a distinctive diamond/star shape with a tight cluster around 387, suggesting standardized repair costs (windshield replacement at fixed rates). Premium (15.7) is too low relative to this.

-   **Theft:** Premium violin is wide and spread (7.2 median) showing some risk differentiation, but incurred (310.4) has a compact rectangular shape, meaning theft claims tend to fall within a predictable cost band. The pricing variation does not match the claim pattern.

-   **Legal protection:** The only coverage where shapes are somewhat proportional. Both violins are compact and narrow, suggesting this is the best priced optional coverage.

    **➨ Overall pattern:** Mandatory coverages (liability, property damage) show wide, well shaped violins on both sides, indicating mature risk based pricing. Optional coverages show flat/narrow premium violins against much larger and varied incurred violins, confirming systematic underpricing of add on products.
:::

### Correlation Matrix

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 8
#| fig-height: 6
#| 
ggstatsplot::ggcorrmat(
  data = motoinsu, 
  cor.vars = c(total_premium, total_incurred, total_claims,
               total_exposure, vehicle_value, driver_age,
               vehicle_age, age_driving_licence, power_to_weight_ratio),
  title = "Correlation Matrix of Key Variables",
  colors = c(RColorBrewer::brewer.pal(3, "BrBG")[c(1, 2, 3)]),
  ggcorrplot.args = list(
    lab_size = 4
  )
)
```

::: callout-note
## Insights

**Strong pairs:**

-   `vehicle_age` and `age_driving_licence` (0.86): nearly duplicate variables
-   `total_exposure` and `total_premium` (0.6): longer policies generate higher premiums, expected
-   `vehicle_value` and `total_premium` (0.32): surprisingly weak, pricing is not strongly value-driven

**Negative relationships:**

-   `power_to_weight_ratio` and `vehicle_value` (-0.49): high-performance vehicles tend to be undervalued relative to their actual risk
-   `power_to_weight_ratio` and `age_driving_licence` (-0.18): younger licence holders gravitate toward higher-performance vehicles
:::

::: callout-warning
`total_incurred` shows near-zero correlation with almost everything individually, meaning claim severity is driven by variable interactions rather than single factors. Multivariate analysis is necessary.
:::

## Multivariate Analysis

### Heatmap: Policy type vs Fuel type

We use heatmap to visualize three variables.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
motoinsu %>%
  group_by(policy_type, fuel_type) %>%
  summarise(loss_ratio = sum(total_incurred) / sum(total_premium) * 100) %>%
  ggplot(aes(x = fuel_type, y = policy_type, fill = loss_ratio)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Loss Ratio Heatmap: Policy Type vs Fuel Type") +
  theme_minimal()



```

::: callout-note
## Insights

-   **COMP_N:** Has the highest loss ratio (\~90%+) for **both** fuel types, with Diesel (D) appearing even more extreme (deep red). This policy type is clearly underperforming and likely unprofitable.
-   **Fuel type (D vs G):** has minimal impact across most policy types, it likely shouldn't be a primary rating factor in pricing models.
:::

### Bubble chart: Loss ratio by Vehicle brand

Let's see the which vehicle brand has higher loss ratio. Note that we remove Others category from Vehicle brand group.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 8
#| fig-height: 6
#| 
motoinsu %>%
  group_by(vehicle_brand_grouped) %>%
  summarise(
    loss_ratio = sum(total_incurred)/sum(total_premium)*100,
    total_premium = sum(total_premium)
  ) %>%
  filter(vehicle_brand_grouped != "Others") %>%
  ggplot(aes(x = total_premium, y = loss_ratio,    # x = total_premium now
             size = total_premium, color = vehicle_brand_grouped)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = vehicle_brand_grouped),
            vjust = -1.5, size = 3, color = "black") +
  geom_text(aes(label = paste0("€", scales::comma(round(total_premium/1e6, 1)), "M")),
            vjust = 2.5, size = 2.8, color = "grey30") +
  scale_x_continuous(labels = scales::comma) +     # x-axis shows premium numbers like y-axis
  scale_size_continuous(range = c(3, 18), guide = "none") +
  labs(
    title = "Loss Ratio vs Vehicle Brand (sized by Total Premium)",
    x = "Total Premium (€)", y = "Loss Ratio (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

::: callout-note
**High risk, low revenue — most concerning:**

-   **OPEL** (77%, €7.4M) — highest loss ratio but one of the smallest bubbles. Worst combination: unprofitable AND small contributor
-   **SEAT** (74%, €7.4M) — same story, small premium volume but elevated risk

**Acceptable risk, decent revenue:**

-   **VOLKSWAGEN** (71%, €8.7M) — largest bubble among named brands, moderate loss ratio. Most important brand to manage carefully given its revenue weight
-   **CITROEN** (69%, €8.5M) — large bubble, below-average loss ratio. Best balance of volume and profitability among French brands

**Best performers:**

-   **PEUGEOT** (66%, €8.1M) and **RENAULT** (67%, €9.1M) — lowest loss ratios with solid premium volumes. RENAULT is particularly valuable: largest revenue among named brands AND most profitable French brand
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 12
#| fig-height: 7

anim_data <- motoinsu %>%
  filter(vehicle_brand_grouped != "Others") %>%
  group_by(year, vehicle_brand_grouped) %>%
  summarise(
    loss_ratio = sum(total_incurred) / sum(total_premium) * 100,
    .groups = "drop"
  ) %>%
  mutate(year_num = as.numeric(as.character(year)))

p <- ggplot(anim_data, aes(x = year_num, y = loss_ratio, 
                            color = vehicle_brand_grouped,
                            group = vehicle_brand_grouped)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = vehicle_brand_grouped),
            vjust = 1.5, size = 3, show.legend = FALSE, fontface = "bold") +
  geom_hline(yintercept = 70, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  labs(title = "Loss Ratio Trend by Vehicle Brand (2022–2024)",
       x = "Year", y = "Loss Ratio (%)", color = "Brand") +
  theme_minimal() +
  transition_reveal(year_num) +
  ease_aes("linear")

animate(p, nframes = 60, fps = 10, width = 800, height = 500,
        renderer = gifski_renderer(),
        res = 100, units = "px")
```

::: callout-note
-   Across three year window, **RENAULT and PEUGEOT** are the most stable, staying in the 63 to 68% range. They are the safest brands in the portfolio and should be prioritized for customer retention.

-   **OPEL** continues to show warning to be the highest volatile brand. This instability suggests a small portfolio where a few large claims can swing the ratio wildly. Unpredictable and hard to price.

-   **SEAT** is consistently the worst among mid range brands, hovering around 70 to 80% across all three years. It never dips below the benchmark, confirming it is structurally underpriced.
:::

::: callout-warning
-   **RENAULT and PEUGEOT** are the most valuable brands, combining high premium volume with the lowest and most stable loss ratios across all three years. These should be prioritized for customer retention.
-   **OPEL and SEAT** generate the least revenue while being the most unprofitable. OPEL is extremely volatile year to year, while SEAT is consistently above the benchmark, confirming structural underpricing. Both require urgent pricing review.
:::

# **References**

::: {#refs}
:::
