---
title: "Technical Report"
author: "Nguyen Kim Hau"
date: "May 15, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Data wrangling

## Importing dataset

```{r}
pacman::p_load(tidyverse,
               psych,
               dplyr,
               patchwork
               ) 
```

The dataset contains anonymised microdata on automobile insurance policies and covers the main operations of the non-life portfolio from January 2022 to December 2024. The database shows 354,140 records and 47 variables, where each row corresponds to a policy transaction and each column to a specific attribute of that transaction.

This dataset uses semicolon delimiters, we use read_csv2() that is a base R shortcut that assumes semicolon delimiters by default.

```{r}
motoinsu <- read_csv2("data/Motor_insurance_portfolio.csv")
```

Out of a total of 47 attributes in the motoinsu tibble data frame, 29 of them are numerical data type (dbl - double) and 28 features are categorical data type (char - character).

```{r}
describe(motoinsu %>% 
           select(where(is.numeric)) # Only show numerical columns
         ) %>%
  select(vars, min, max, mean, skew) # Only get desired columns
```

```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(kableExtra)

var_desc <- read_excel("data/Descriptive_of_variables.xlsx") %>%
  fill(everything(), .direction = "down") %>%
  group_by(Variables, Group) %>%
  summarise(Description = paste(Description, collapse = " "), .groups = "drop")

var_desc %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
glimpse(motoinsu)
```

## Conversing datatype

We realize that we need to change data type of some variables as follows.

```{r}
motoinsu <- motoinsu %>%
  mutate(
    # from dbl to chr (identifier)
    insured_id = as.character(insured_id),
    
    # from dbl to fac (categorical)
    across(c(year, policy_type, policy_status, business_type,
             payment_frequency, bonus_score, fuel_type, seats,
             vehicle_brand, municipality_type, circulation_area), 
           as.factor),
    
    # from char to dbl
    across(c(liability_premium, property_damage_premium,
             theft_premium, fire_premium, glass_premium,
             legal_protection_premium, occupants_premium,
             total_exposure, liability_exposure), 
           as.double)
  )

glimpse(motoinsu)
```

::: callout-tip
-   dbl stands for: double, numbers with decimals.
-   fct: factor, text with fixed, known categories. Factor uses less memory, which matters in such large database.
-   char: character, plain text without predefined categories.
:::

After changing data type, we have 1 identifier, 11 categorical variables and 35 numerical variables.

## Handling missing values

There are six features having missing values. Although all of these rates are small, under 1%, we treat them differently.

```{r}
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

```{r}
colMeans(is.na(motoinsu)) %>% # Caculate missing rate in each column
  round(5) %>%
  .[. > 0] # Filter ratio > 0
```

-   `vehicle_age`, `age_driving_licence` only lack two values, we drop these missing rows.
-   `fire_incurred`, `occupants_incurred` having missing values likely means no claim, in insurance context. Therefore we can change to zero.
-   `fuel_type`, `vehicle_value` have more missing values, which make I consider between using mode/medium or remove missing rows. Because they are seem important features, it is risky to impute data. This issue should be reported to the relevant stakeholders.

```{r}
motoinsu <- motoinsu %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value ) %>%
  mutate(
    fire_incurred    = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(occupants_incurred, 0)
  )
# Check after modification
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

# Visual Analysis

## Univariate Analysis

### Categorical variables

Among 11 categorical columns, vehicle_brand has too many distinct values. Therefore we will visualize its distribution and impute it later. Following is the distributions of 10 categorical features.

```{r}
#| fig-width: 15 # Adjust width of canvas
#| fig-height: 15 # Adjust height of canvas
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

cat_vars <- c("year", "policy_type", "policy_status", "business_type",
              "payment_frequency", "bonus_score", 
              "fuel_type", "seats",
              "municipality_type", "circulation_area")

plots <- lapply(cat_vars, function(col) {
  ggplot(motoinsu, aes(x = fct_infreq(!!sym(col)))) +
    geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
    geom_text(aes(y = after_stat(count/sum(count) * 100),
                  label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
              stat = "count", vjust = -0.5, size = 3) +
    labs(title = paste("Distribution of", col), x = col, y = "Percentage (%)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights:

#### Policy & Business Characteristics

-   **Year**: Data is unbalanced - 2024 (47.4%) has the most records, likely because policies accumulate over time. Dataset covers 3 years of a growing portfolio.
-   **Policy type**: CC dominates (58%), meaning most customers choose combined third-party + additional coverage. Pure TP is rare (1.6%) - customers prefer more protection.
-   **Policy status**: 89.2% active (A) vs 10.8% cancelled (C) - relatively low cancellation rate, portfolio is stable.
-   **Business type**: 69.3% new business (NB) vs 30.7% portfolio (P), which means the insurer is heavily acquiring new customers rather than renewing existing ones. This is a risk signal, new customers have unknown claims history.
-   **Payment frequency**: 81.3% pay annually (A) - most customers prefer single payment, simplifying cash flow for the insurer.
-   **Bonus score**: 96.5% have score G (good/favorable claims history) - the portfolio is dominated by low-risk customers historically.

#### Vehicle Characteristics

-   **Fuel type**: 64.2% diesel (D) vs 35.8% gasoline (G) - diesel vehicles dominate, which may have different risk profiles.
-   **Seats**: 84.9% have 5 seats - standard passenger cars dominate, very homogeneous portfolio.

#### Geographic

-   **Municipality type**: 64.6% type I (Inland) where typically have higher accident/theft risk than coastal and island areas.
-   **Circulation area**: Nearly split between U (53.8%) and R (46.2%) - urban vs rural almost balanced, suggesting national coverage.
:::

Vehicle brand distribution show high fragmentation with top brand \~10%, many brands with \<1%. We will combine all brands excluding RENAULT, CITROEN, PEUGEOT, VOLKSWAGEN, FORD, SEAT, and OPEL.

```{r}
#| fig-width: 15
#| fig-height: 6
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Then plot them by distribution.

```{r}
#| fig-width: 8
#| fig-height: 4

# Bin
top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")

motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

# Draw chart
ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand_grouped))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: callout-note
## Insights:

-   The near-equal spread among top brands means risk is **not concentrated** in any single brand's characteristics.
-   Among top 7, French brands (RENAULT, CITROEN, PEUGEOT) collectively represent \~27%, reflecting **Spain's strong preference for French** automobiles.
-   SEAT's presence (\~7.9%) is notable as a **Spanish domestic brand**, expected in a Spanish portfolio.
-   The large "Others" group (40.6%) warrants further investigation - it contains luxury brands (BMW, MERCEDES, AUDI) and niche brands that **may carry higher vehicle values and different risk profiles**.
:::

### Numerical variables

Driver & vehicle Distributions

```{r}
#| fig-width: 16
#| fig-height: 8

num_vars <- c("driver_age", "vehicle_age", "age_driving_licence",
              "vehicle_value", "power_to_weight_ratio")

plots <- lapply(num_vars, function(col) {
  ggplot(motoinsu, aes(x = as.numeric(!!sym(col)))) + # as.numeric() inside aes() forces any discrete/factor columns to be treated as continuous
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = paste("Distribution of", col), x = col, y = "Count") +
    theme_minimal()
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights

#### Driver & Licence Characteristics

-   **driver_age**: roughly bell-shaped, peaking around **35–50 years**. Young drivers (\<25) and elderly (\>70) are rare, suggesting the portfolio skews toward middle-aged, experienced drivers.
-   **vehicle_age**: bimodal distribution with peaks around **5–10 years** and **20–25 years**. Many both newer and older vehicles insured.
-   **age_driving_licence**: heavily right-skewed, most policyholders have held licences for **less than 20 years**, consistent with the driver age distribution.

#### Vehicle Characteristics

-   **vehicle_value**: extremely right-skewed with most values near **zero** and a long tail. Most insured vehicles are low-to-mid value; a small number are very high-value vehicles (potential outliers/luxury cars).
-   **power_to_weight_ratio**: right-skewed, concentrated around **1000–2000**, with a few extreme high-performance vehicles.
:::

Premium, Claims & Exposure Distributions

```{r}
#| fig-width: 15
#| fig-height: 15

num_vars <- c("total_premium", "liability_premium", "property_damage_premium",
              "theft_premium", "fire_premium", "glass_premium",
              "legal_protection_premium", "occupants_premium",
              "total_claims", "liability_claims", "property_claims",
              "total_incurred", "liability_incurred", "property_incurred",
              "total_exposure", "liability_exposure")

plots <- lapply(num_vars, function(col) {
  ggplot(motoinsu, aes(x = as.numeric(!!sym(col)))) + # as.numeric() inside aes() forces any discrete/factor columns to be treated as continuous
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = paste("Distribution of", col), x = col, y = "Count") +
    theme_minimal()
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights

#### Premium Variables

-   **total_premium**: right-skewed, most policies have low-to-mid premiums with a long tail of high-premium policies
-   **liability_premium**: concentrated around **100–400**, the most normally distributed of all premiums — makes sense as liability coverage is **mandatory** for all vehicles in Spain
-   **property_damage_premium, theft_premium, fire_premium, glass_premium, legal_protection_premium, occupants_premium**: all **extremely right-skewed with mass at zero**, meaning most policies **do not include** these optional coverages. This is consistent with CC being 58% of policies — only combined policies carry these extras.

#### Claims Variables

-   **total_claims, liability_claims, property_claims**: **overwhelming majority are zero**. Rare claims events with counts of 1–2 for most claimants, very few exceed 3. Classic **low-frequency** insurance pattern.

#### Incurred Loss Variables

-   **total_incurred, liability_incurred, property_incurred**: **almost all zero** with extreme outliers reaching 10\^15. These extreme values are suspicious and may indicate **data quality issues** or very catastrophic events worth investigating.

#### Exposure Variables

-   **total_exposure & liability_exposure**: both show a **U-shaped/bimodal pattern** concentrated at **0.25 and 1.0**:

    -   **1.0** = full year policy (annual payment, 81.3%)
    -   **0.25** = quarterly payment (4.3%)
    -   Values between suggest mid-year cancellations (10.8% cancelled policies)
:::

Have a closer look on outlier cases in all numerical variables. Seven features show abnormal maximum values comparing to quantile 99.9%, which are "total_incurred", "liability_incurred", "liability_property_incurred", "liability_injury_incurred", "property_incurred", "glass_incurred", "legal_protection_incurred".

```{r}
incurred_vars <- c("total_incurred", "liability_incurred", "liability_property_incurred",
                    "liability_injury_incurred", "property_incurred", 
                    "glass_incurred", "legal_protection_incurred"
                    )

motoinsu %>%
  summarise(across(all_of(incurred_vars), list(
    max    = ~max(., na.rm = TRUE),
    p999 = ~quantile(., 0.999, na.rm = TRUE),
    q99 = ~quantile(., 0.99, na.rm = TRUE),
    q75 = ~quantile(., 0.75, na.rm = TRUE),
    q50 = ~quantile(., 0.50, na.rm = TRUE),
    q25 = ~quantile(., 0.25, na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  kable(format.args = list(big.mark = ",", scientific = FALSE)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

They seem clearly data entry errors (likely missing decimal points). Therefore, we will remove them for not distort all analysis later. We use 0.999 threshold removes only the truly impossible values while keeping legitimate large claims, minimizing data loss (remove 0.5% of total rows). Such outliers need to inform to relevant stakeholders.

```{r}
original_n <- 354140

motoinsu <- motoinsu %>%
  filter(if_all(all_of(incurred_vars), 
                ~ . <= quantile(., 0.999, na.rm = TRUE)))

cat("Records before:", original_n, "\n",
    "Records after:", nrow(motoinsu), "\n",
    "Records removed:", original_n - nrow(motoinsu), "\n",
    "Percentage removed:", round((original_n - nrow(motoinsu))/original_n * 100, 3), "%\n")
```
