---
title: "Technical Report"
author: "Nguyen Kim Hau"
date: "May 15, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Data wrangling

## Importing dataset

```{r}
pacman::p_load(tidyverse,
               psych,
               dplyr,
               patchwork
               ) 
```

The dataset contains anonymised microdata on automobile insurance policies and covers the main operations of the non-life portfolio from January 2022 to December 2024. The database shows 354,140 records and 47 variables, where each row corresponds to a policy transaction and each column to a specific attribute of that transaction.

This dataset uses semicolon delimiters, we use read_csv2() that is a base R shortcut that assumes semicolon delimiters by default.

```{r}
motoinsu <- read_csv2("data/Motor_insurance_portfolio.csv")
```

Out of a total of 47 attributes in the motoinsu tibble data frame, 29 of them are numerical data type (dbl - double) and 28 features are categorical data type (char - character).

```{r}
describe(motoinsu %>% 
           select(where(is.numeric)) # Only show numerical columns
         ) %>%
  select(vars, min, max, mean, skew) # Only get desired columns
```

```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(kableExtra)

var_desc <- read_excel("data/Descriptive_of_variables.xlsx") %>%
  fill(everything(), .direction = "down") %>%
  group_by(Variables, Group) %>%
  summarise(Description = paste(Description, collapse = " "), .groups = "drop")

var_desc %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
glimpse(motoinsu)
```

## Conversing datatype

We realize that we need to change data type of some variables as follows.

```{r}
motoinsu <- motoinsu %>%
  mutate(
    # from dbl to chr (identifier)
    insured_id = as.character(insured_id),
    
    # from dbl to fac (categorical)
    across(c(year, policy_type, policy_status, business_type,
             payment_frequency, bonus_score, fuel_type, seats,
             vehicle_brand, municipality_type, circulation_area), 
           as.factor),
    
    # from char to dbl
    across(c(liability_premium, property_damage_premium,
             theft_premium, fire_premium, glass_premium,
             legal_protection_premium, occupants_premium,
             total_exposure, liability_exposure), 
           as.double)
  )

glimpse(motoinsu)
```

::: callout-tip
-   dbl stands for: double, numbers with decimals.
-   fct: factor, text with fixed, known categories. Factor uses less memory, which matters in such large database.
-   char: character, plain text without predefined categories.
:::

After changing data type, we have 1 identifier, 11 categorical variables and 35 numerical variables.

## Handling missing values

There are six features having missing values. Although all of these rates are small, under 1%, we treat them differently.

```{r}
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

```{r}
colMeans(is.na(motoinsu)) %>% # Caculate missing rate in each column
  round(5) %>%
  .[. > 0] # Filter ratio > 0
```

-   `vehicle_age`, `age_driving_licence` only lack two values, we drop these missing rows.
-   `fire_incurred`, `occupants_incurred` having missing values likely means no claim, in insurance context. Therefore we can change to zero.
-   `fuel_type`, `vehicle_value` have more missing values, which make I consider between using mode/medium or remove missing rows. Because they are seem important features, it is risky to impute data. This issue should be reported to the relevant stakeholders.

```{r}
motoinsu <- motoinsu %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value ) %>%
  mutate(
    fire_incurred    = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(occupants_incurred, 0)
  )
# Check after modification
colSums(is.na(motoinsu)) %>% 
  .[. > 0]
```

# Visual Analysis

## Univariate Analysis

### Categorical variables

Among 11 categorical columns, vehicle_brand has too many distinct values. Therefore we will visualize its distribution and impute it later. Following is the distributions of 10 categorical features.

```{r}
#| fig-width: 15 # Adjust width of canvas
#| fig-height: 20 # Adjust height of canvas
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

cat_vars <- c("year", "policy_type", "policy_status", "business_type",
              "payment_frequency", "bonus_score", 
              "fuel_type", "seats",
              "municipality_type", "circulation_area")

plots <- lapply(cat_vars, function(col) {
  ggplot(motoinsu, aes(x = fct_infreq(!!sym(col)))) +
    geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
    geom_text(aes(y = after_stat(count/sum(count) * 100),
                  label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
              stat = "count", vjust = -0.5, size = 3) +
    labs(title = paste("Distribution of", col), x = col, y = "Percentage (%)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

wrap_plots(plots, ncol = 3)
```

::: callout-note
## Insights:

#### Policy & Business Characteristics

-   **Year**: Data is unbalanced - 2024 (47.4%) has the most records, likely because policies accumulate over time. Dataset covers 3 years of a growing portfolio.
-   **Policy type**: CC dominates (58%), meaning most customers choose combined third-party + additional coverage. Pure TP is rare (1.6%) - customers prefer more protection.
-   **Policy status**: 89.2% active (A) vs 10.8% cancelled (C) - relatively low cancellation rate, portfolio is stable.
-   **Business type**: 69.3% new business (NB) vs 30.7% portfolio (P), which means the insurer is heavily acquiring new customers rather than renewing existing ones. This is a risk signal, new customers have unknown claims history.
-   **Payment frequency**: 81.3% pay annually (A) - most customers prefer single payment, simplifying cash flow for the insurer.
-   **Bonus score**: 96.5% have score G (good/favorable claims history) - the portfolio is dominated by low-risk customers historically.

#### Vehicle Characteristics

-   **Fuel type**: 64.2% diesel (D) vs 35.8% gasoline (G) - diesel vehicles dominate, which may have different risk profiles.
-   **Seats**: 84.9% have 5 seats - standard passenger cars dominate, very homogeneous portfolio.

#### Geographic

-   **Municipality type**: 64.6% type I (Inland) where typically have higher accident/theft risk than coastal and island areas.
-   **Circulation area**: Nearly split between U (53.8%) and R (46.2%) - urban vs rural almost balanced, suggesting national coverage.
:::

Vehicle brand distribution show high fragmentation with top brand \~10%, many brands with \<1%. We will combine all brands excluding RENAULT, CITROEN, PEUGEOT, VOLKSWAGEN, FORD, SEAT, and OPEL.

```{r}
#| fig-width: 15 # Adjust width of canvas
#| fig-height: 6 # Adjust height of canvas
# gom_bar():  counts rows automatically
# fct_infreq: reorder factor levels based on the frequency of values

ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Then plot them by distribution.

```{r}

# Bin
top_brands <- c("RENAULT", "CITROEN", "PEUGEOT", "VOLKSWAGEN", "FORD", "SEAT", "OPEL")

motoinsu <- motoinsu %>%
  mutate(vehicle_brand_grouped = fct_other(vehicle_brand, 
                                            keep = top_brands, 
                                            other_level = "Others"))

# Draw chart
ggplot(motoinsu, aes(x = fct_infreq(vehicle_brand_grouped))) +
  geom_bar(aes(y = after_stat(count/sum(count) * 100)), fill = "steelblue") +
  geom_text(aes(y = after_stat(count/sum(count) * 100),
                label = paste0(round(after_stat(count/sum(count) * 100), 1), "%")),
            stat = "count", vjust = -0.5, size = 3) +
  labs(title = "Distribution of Vehicle Brand", x = "Vehicle Brand", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: callout-note
## Insights:

-   The near-equal spread among top brands means risk is not concentrated in any single brand's characteristics.

<!-- -->

-   Among top 7, French brands (RENAULT, CITROEN, PEUGEOT) collectively represent \~27%, reflecting Spain's strong preference for French automobiles.

<!-- -->

-   SEAT's presence (\~7.9%) is notable as a Spanish domestic brand, expected in a Spanish portfolio.

<!-- -->

-   The large "Others" group (40.6%) warrants further investigation - it contains luxury brands (BMW, MERCEDES, AUDI) and niche brands that may carry higher vehicle values and different risk profiles.
:::
